---
title: "Snake_Analysis"
author: "Owen McCool, Therese Lamperty"
date: "2025-07-16"
output: word_document
---

# Code Setup & package read in 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read in packages}
library(vegan)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(iNEXT) # species diversity
library(hillR) # species diversity
library(DHARMa) # checking models
library(car) # reporting model stats
library(glmmTMB)
library(unmarked) #density estimates with hierarchical distance sampling models
library(AICcmodavg) # has convenient functions for comparing models based on AIC
library(showtext)
library(ggsignif)
showtext_auto()
font_add(family = "Tahoma", regular = "tahoma.ttf")

```

# Project Overview

Study Site: Research was split between two different zones within the greater Yasuni Biosphere Reserve. Tiputini Biodiversity Station (TBS) served as the faunally intact site, whereas Yasuni Research Station (YRS) served as the defaunated site. Both sites are situated along the Tiputini River within the province of Orellana, Ecuador. TBS, approximately 45 km east of YRS, is only accessible by boat. YRS is along the Maxus Oil Road, and is less integrated into the Amazonian ecosystem. Additionally, unsustainable hunting activities by the indigenous communities have been prominent in the surrounding region at YRS for multiple decades, as noted from 2014. This is thought to be the root cause of defaunation in this zone. 

	I transferred between the two research stations every two weeks, starting my first rotation at TBS on June 10. Time at each station was spent working on 3 separate projects: my herpetofauna surveys, frugivory surveys, and an overarching seed enemy and dispersal project. 

	The rainy season in Northeastern Ecuador is at its peak between March and July. I conducted my study between the months of June and August, 2024. The average high temperature for June in this region is 85.66° F, with the average low of 70.05° F. July has an average high temperature of 86.29° F and an average low of 69.13° F. August has an average high temperature of 91.35° F and an average low of 69.60° F. June receives 143mm of precipitation on average each year, whereas July receives 99.69mm and August receives 77.55mm. (https://weatherandclimate.com/ecuador/orellana/boca-tiputini) 
Put project description here (field sites, time frame of collection,sampling scheme, type of data (eg. species ID,temp, etc.) )

Surveys: Herpetofauna were sampled using visual surveys on foot by two surveyors/night. Trail systems at each site served as the sampling paths. The surveys were repeated around 6 times on each rotation while in the field, and resampled on the second rotations. The surveys typically began around 20:00-20:30 and lasted until 23:00-00:00 CST. At each site, habitat matches were compared and assessed and trails through that habitat were consistently walked. Two different types of habitat, Terra Firma (uplands) and Varzea (flooded forest) were surveyed on the herp nocturnal walks. In addition to this, two significant bodies of water were also surveyed at each site, primarily consisting of oxbow lakes with one year-round large pond. 

	Surveyors used FENIX HM60R headlamps while conducting night surveys. They carefully examined all surrounding habitat, including understory trees, shrubs, vines, water bodies, forest floor, and middle-canopy. Amphibians of the order Anura gave off reflective eyeshine and were easier to sample for, whereas members of Squamata did not. More information about this will be within the discussion section. Effort hours will be calculated based on survey time length and number of surveyors. 

	Additionally, environmental information was taken throughout the course of each survey. Temperature and humidity data was collected with a Mengshen meter in 30 minute intervals on surveys. Understory vegetation density was also measured every 15 minutes along each trail. A 1.5m PVC pipe was randomly placed upright just off the right side of the trail at each measuring point, and the number of leaves making contact with the pipe was counted. 

	Weather and wind were noted before each survey in four nominal categories, with weather either being “clear, partly cloudy, cloudy, rainy” and wind being “no wind, felt on face, branches move, trees move”. I also accounted for precipitation that had already occurred in 6-hour blocks as “time window of last rain event”. 

	Once an animal was found, the first step of the protocol was to note what time of night the find occurred. After this, the temperature and humidity was once again collected. A rangefinder was used to estimate the perpendicular distance of each animal from the trail. A GPS was used to mark and log the coordinates of each animal found. Animals were given a two-letter code based on the Order and Suborder followed by a number, indicating its position of being found. For example, the code of the 9th snake found of the season would be SK09. Species found were also identified to the lowest taxonomic level in the field, and photographs of individuals were taken both to get positive IDs later in the lab when not possible in the field, but also to avoid counting re-captures.

	Snakes found during non-survey times were also recorded. These individuals were split into one of two categories of detection. The first is an “Opportunistic Encounter” (OE), which is any snake that was found by me while out working on other non-herpetofauna projects in the field. The other type of detection is the Third Party Encounter (TPE). This was any snake that was found by other research groups during my rotations at each station, identified to lowest possible taxonomy. The same data was taken for these individuals. GPS coordinates, time of detection, trail, observer, estimation of distance from trail, and habitat type. Because the third-party observers often didn’t have access to the climate data logger, this information was retrieved later on from each station’s respective weather tower data, based on the time of detection. 

Snakes were given a classification as to whether they were small mammal consumers or not. Information on this was retrieved from literature such as online publications, the book Reptiles and Amphibians of the Amazon by R.D. and Patricia Bartlett, and the expert field book Reptiles of Ecuador. 

# Read in and prep data

## Read in


```{r data read in}

#1. all snake records
snakes.master <- read.csv(file = "Data/Cleaned/master_dataset_all_snake_data.csv")

#2. environmental data from both seasons but nights only
enviro_nights <- read.csv(file = "Data/Cleaned/environmental_data_2024-25_nights_only.csv")

#3. all survey records by segment
all_seg_metadata <- read.csv("Data/Cleaned/all_surv_records_by_segments.csv")

#4. amphibian and lizard data 
amph_lz <- read.csv("Data/Cleaned/amph_lz.csv")

#5. environmental data from 2025 (includes afternoons)
env_2025 <- read.csv("Data/Cleaned/environmental_data_2025.csv")

#6. distance and search hours from each survey overall (not broken down by segments)
effort_per_survey <- read.csv("Data/Cleaned/each_survey_effort.csv")

#7. snake data from only surveys
snakes.surv <- read.csv("Data/Cleaned/snakes_survey_data_only.csv")
# Owen added this 2024 amphibian data in

#8. small mammal data
small.mamm.df <- read.csv("Data/Cleaned/small_mammal_data.csv")

#9. analysis-ready lizard and frog detection data
liz_frog_dets <- read.csv("Data/Cleaned/amph_liz_analysis_ready_df.csv")

```


# Small Mammals

## Summarize

```{r summarize small mammal detections}

sum(small.mamm.df$Number_of_Small_Mammals_Spotted[small.mamm.df$Station == "YRS"])

sum(small.mamm.df$Number_of_Small_Mammals_Spotted[small.mamm.df$Station == "TBS"])

small.mamm.df %>%
  group_by(Station) %>%
  summarise(mean_detections_per_survey = mean(Number_of_Small_Mammals_Spotted),
            standard_deviation = sd(Number_of_Small_Mammals_Spotted))

```

## Model

```{r small mammal model}

smam.mod1 <- glmmTMB(smamm.500m ~ Station + (1|Parent_Trail_LongName), data = small.mamm.df, family = tweedie(link = "log"))

sim <- simulateResiduals(fittedModel = smam.mod1, plot = T)

summary(smam.mod1)

testDispersion(smam.mod1, plot = F)
testZeroInflation(smam.mod1, plot = F)

# clean
rm(smam.mod1, sim)
```

## Small Mammal Figure

```{r small mammal figure}

# get mean and SE and SD

smams.distance.mean.se <- small.mamm.df %>%
  group_by(Station) %>%
  summarise(mean = mean(smamm.500m),
            sd = sd(smamm.500m),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# figure

ggplot(smams.distance.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = small.mamm.df, 
              aes(x = Station, y = smamm.500m), 
              width = 0.15, alpha = 0.2) + coord_fixed(ratio = 0.75) + geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21, size = 1.5) + theme_bw() + scale_fill_manual(values = c("blue","orange")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/500m") + ggtitle("Small mammals") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank()) +
  geom_signif(
    y_position = 2.5, xmin = 1, xmax = 2,
    annotation = "***")

# clean
rm(smams.distance.mean.se)
```


# Prep frog, lizard, and env data


Need to check models with the temperature and humidity data

```{r env data prep}

humidity_long <- env_2025 %>%
  select(Date, Station, Survey_Type, 
         Humidity_1:Humidity_8, Humidity_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Humidity_1:Humidity_Final,
               values_to = "Humidity", names_repair = "unique")
# remove NAs

humidity_long <- humidity_long %>%
  na.omit()

# fix date
humidity_long$Start_Date <- mdy(humidity_long$Date)
humidity_long <- humidity_long %>%
  select(-Date)

# repeat with temp
temp_long <- env_2025 %>%
  select(Date, Station, Survey_Type, 
         Temperature_1:Temperature_8, Temperature_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Temperature_1:Temperature_Final,
               values_to = "Temperature", names_repair = "unique")

# remove NAs
temp_long <- temp_long %>%
  na.omit()
# date
temp_long$Start_Date <- mdy(temp_long$Date)
temp_long <- temp_long %>%
  select(-Date)

# understory

understory_long <- env_2025 %>%
  select(Date, Station, Survey_Type, 
         Understory_Density_1:Understory_Density_13, Understory_Density_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Understory_Density_1:Understory_Density_Final,
               values_to = "Understory", names_repair = "unique")
# remove NAs

understory_long <- understory_long %>%
  na.omit()

# fix date
understory_long$Start_Date <- mdy(understory_long$Date)
understory_long <- understory_long %>%
  select(-Date)

# 2024 data
# make date objects correct class
enviro_nights$Start_Date <- ymd(enviro_nights$Start_Date)
enviro_nights$End_Date <- mdy(enviro_nights$End_Date)

# pull out 2024 year only
enviro_nights$Year<- factor(format(enviro_nights$Start_Date, "%Y"))
  
env_2024 <- enviro_nights %>%
  filter(Year == "2024")

# humidity - 2024 - pivot

# simplify
env_2024 <- env_2024 %>%
  select(-Effort_hrs, -Notes, -Number_of_Small_Mammals_Spotted, -Number_of_Water_Bodies_Spotted, -Surveyors, -Searching_Time, -Meters, -Order, -Survey_Start_Time, -Survey_End_Time)

humidity_long_2024 <- env_2024 %>%
  select(Start_Date, Station,
         Humidity_1:Humidity_8, Humidity_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Humidity_1:Humidity_Final,
               values_to = "Humidity", names_repair = "unique")
# remove NAs

humidity_long_2024 <- humidity_long_2024 %>%
  na.omit()

# repeat with temp
temperature_long_2024 <- env_2024 %>%
  select(Start_Date, Station,
         Temperature_1:Temperature_8, Temperature_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Temperature_1:Temperature_Final,
               values_to = "Temperature", names_repair = "unique")
# remove NAs

temperature_long_2024 <- temperature_long_2024 %>%
  na.omit()

# repeat with understory
understory_long_2024 <- env_2024 %>%
  select(Start_Date, Station,
         Understory_Density_1:Understory_Density_13, Understory_Density_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Understory_Density_1:Understory_Density_Final,
               values_to = "Understory", names_repair = "unique")
# remove NAs

understory_long_2024 <- understory_long_2024 %>%
  na.omit()

# join
# add survey type column to 2024 env data
humidity_long_2024$Survey_Type <-"Night"
temperature_long_2024$Survey_Type <-"Night"
understory_long_2024$Survey_Type <-"Night"

humidity_all <- rbind(humidity_long, humidity_long_2024)
temperature_all <- rbind(temp_long, temperature_long_2024)
understory_all <-rbind(understory_long, understory_long_2024)

# get means

humidity_means <- humidity_all %>%
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(mean_relative_humidity = mean(Humidity))

temperature_means <- temperature_all %>%
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(mean_temp_F = mean(Temperature))

understory_means <- understory_all %>%
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(mean_understory_leaves = mean(Understory))

# prep to join
liz_frog_dets$Start_Date<-mdy(liz_frog_dets$Start_Date)
humidity_means$Start_Date<-ymd(humidity_means$Start_Date)
understory_means$Start_Date<-ymd(understory_means$Start_Date)
temperature_means$Start_Date<-ymd(temperature_means$Start_Date)

# add to the amph lz detection data
liz_frog_dets <- left_join(liz_frog_dets, humidity_means)

liz_frog_dets <- left_join(liz_frog_dets, temperature_means)

liz_frog_dets <- left_join(liz_frog_dets, understory_means)
```

```{r hum and temp and und}

humidity_all$Year <- factor(format(humidity_all$Start_Date, "%Y"))

temperature_all$Year <- factor(format(temperature_all$Start_Date, "%Y"))

understory_all$Year <- factor(format(understory_all$Start_Date, "%Y"))


humidity_all <- humidity_all %>%
  filter(Survey_Type != "Morning")

humidity_summary_yr_survey_type <- humidity_all %>%
  filter(!(Year==2024 & Survey_Type == "Afternoon")) %>%
  filter(!(Survey_Type == "Morning")) %>%
  group_by(Station, Survey_Type, Year) %>%
  summarize(mean_humidity = mean(Humidity),
            sd = sd(Humidity),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# temperature
temperature_all <- temperature_all %>%
  filter(Survey_Type != "Morning")

temperature_summary_yr_survey_type <- temperature_all %>%
  filter(!(Year==2024 & Survey_Type == "Afternoon")) %>%
  filter(!(Survey_Type == "Morning")) %>%
  group_by(Station, Survey_Type, Year) %>%
  summarize(mean_temperature = mean(Temperature),
            sd = sd(Temperature),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# understory
understory_all <- understory_all %>%
  filter(Survey_Type != "Morning")

understory_summary_yr_survey_type <- understory_all %>%
  filter(!(Year==2024 & Survey_Type == "Afternoon")) %>%
  filter(!(Survey_Type == "Morning")) %>%
  group_by(Station, Survey_Type, Year) %>%
  summarize(mean_understory = mean(Understory),
            sd = sd(Understory),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

```{r temperature fig}

temp.fig <- ggplot(temperature_summary_yr_survey_type, aes(x=Station, y=mean_temperature, fill = Survey_Type)) + geom_jitter(data = temperature_all, aes( x = Station, y = Temperature), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.45) + theme_bw() +
  scale_fill_manual(values = c("Chocolate", "blue3")) + 
  geom_errorbar(aes(ymin=mean_temperature-se, ymax=mean_temperature+se), width=.5) + geom_point(pch = 21, size = 3) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Temperature (F)") + ggtitle("Survey temperature ranges") + theme_bw(base_family = "Tahoma") + theme(plot.title = element_text(face = "bold", hjust = .5))  + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + facet_wrap(~Year) + labs(fill = "Survey Type")

temp.fig

#ggsave(temp.fig, file="Figures/temperature_fig_updatedJune2026.png", width = 6,  height = 4)

```

```{r hum fig}

hum.fig <- ggplot(humidity_summary_yr_survey_type, aes(x=Station, y=mean_humidity, fill = Survey_Type)) + geom_jitter(data = humidity_all, aes( x = Station, y = Humidity), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.45) + theme_bw() +
  scale_fill_manual(values = c("Chocolate", "blue3")) + 
  geom_errorbar(aes(ymin=mean_humidity-se, ymax=mean_humidity+se), width=.5) + geom_point(pch = 21, size = 3) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Relative humidity (%)") + ggtitle("Survey humidity ranges") + theme_bw(base_family = "Tahoma") + theme(plot.title = element_text(face = "bold", hjust = .5))  + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + facet_wrap(~Year) + labs(fill = "Survey Type")

hum.fig

#ggsave(hum.fig, file="Figures/humidity_fig_updatedJune2026.png", width = 6,  height = 4)

```

```{r und fig}
und.fig <- ggplot(understory_summary_yr_survey_type, aes(x=Station, y=mean_understory, fill = Survey_Type)) + geom_jitter(data = understory_all, aes( x = Station, y = Understory), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.45) + theme_bw() +
  scale_fill_manual(values = c("Chocolate", "blue3")) + 
  geom_errorbar(aes(ymin=mean_understory-se, ymax=mean_understory+se), width=.5) + geom_point(pch = 21, size = 3) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Understory (vertical leaf overlap)") + ggtitle("Survey understory density ranges") + theme_bw(base_family = "Tahoma") + theme(plot.title = element_text(face = "bold", hjust = .5))  + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + facet_wrap(~Year) + labs(fill = "Survey Type")

und.fig

#ggsave(und.fig, file="Figures/understory_fig_updatedJune2026.png", width = 6,  height = 4)
```
## Clean

```{r clean after env org}
rm(understory_long_2024, understory_long, understory_means, understory_summary_yr_survey_type, temperature_long_2024, temperature_means, temperature_summary_yr_survey_type, temp_long, humidity_long, humidity_long_2024, humidity_means, humidity_summary_yr_survey_type, hum.fig, temp.fig, und.fig)
```

# Frog Detections

## Summarize

```{r summarize frogs}

amph_lz_detections %>%
  filter(Survey_Type != "Morning") %>%
  na.omit() %>%
  group_by(Station, Survey_Type) %>%
  summarise(mean_frogs = mean(n_frogs),
            sd = sd(n_frogs),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

### Distance per 500m

Look at response variable as a function of distance now to be consistent with small mammal models. Add in columns representing n individuals/500m first

```{r frog detections per 500m mods}

liz_frog_dets <- liz_frog_dets %>%
  na.omit() %>%
  filter(Survey_Type != "Morning") %>%
  filter(Height_of_Individual != "") %>%

hist(liz_frog_dets$n_frogs_500m)
# null
frog.dist.null <- glmmTMB(n_frogs_500m ~ 1 + (1|Parent_Trail_LongName), data = liz_frog_dets, ziformula = ~1, family = lognormal())

sim <- simulateResiduals(fittedModel = frog.dist.null, plot = T)

summary(frog.dist.null)
# AIC  1430.0

# start simple, only hunting status (station) and survey type
frog.dist.mod1 <- glmmTMB(n_frogs_500m ~ Station + (1|Parent_Trail_LongName), data = liz_frog_dets,  ziformula = ~1, family = lognormal())

sim <- simulateResiduals(fittedModel = frog.dist.mod1, plot = T)
# is ok

summary(frog.dist.mod1)
# AIC 1430.1 
# no sig effect of station on frog detections

# make model more complex, add interaction effect of hunting status with survey type
frog.dist.mod2 <- glmmTMB(n_frogs_500m ~ Station/Survey_Type + (1|Parent_Trail_LongName), data = liz_frog_dets, ziformula = ~1, family = lognormal())

sim <- simulateResiduals(fittedModel = frog.dist.mod2, plot = T)
# passes

summary(frog.dist.mod2)
# AIC worse: 1433

#drop survey type bc doesn't improve model fit

#investigate humidity

frog.dist.mod3 <- glmmTMB(n_frogs_500m ~ mean_relative_humidity + (1|Parent_Trail_LongName), data = liz_frog_dets, ziformula = ~1, family = lognormal())

sim <- simulateResiduals(fittedModel = frog.dist.mod3, plot = T)
# nope, terrible

# temp
frog.dist.mod4 <- glmmTMB(n_frogs_500m ~ mean_temp_F + (1|Parent_Trail_LongName), data = liz_frog_dets, ziformula = ~1, family = lognormal())

sim <- simulateResiduals(fittedModel = frog.dist.mod4, plot = T)
# passes validation
# AIC  worse 301.2, so remove temperature
summary(frog.dist.mod4)

```







### Figure















# Lizards per hour

## Summarize

```{r summarize Lizards}

amph_lz_detections %>%
  group_by(Station) %>%
  summarise(
    tot_liz = sum(n_liz),
    mean_Lizards = mean(n_liz),
            sd = sd(n_liz),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

## Models

### Time

```{r lz mod}

# start simple, only hunting status (station) and survey type
liz.cont.mod1 <- glmmTMB(n_liz_hr ~ Station + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.cont.mod1, plot = T)
# passes validation

summary(liz.cont.mod1)
# AIC 275.8  
# sig effect of station

# make model more complex, add interaction effect of hunting status with survey type, thing on left of / is fixed effect, 
liz.cont.mod2 <- glmmTMB(n_liz_hr ~ Station/Survey_Type + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.cont.mod2, plot = T)
# passes
# AIC:  278.1 (worse)
summary(liz.cont.mod2)

#drop survey type bc doesn't improve model fit

#investigate humidity
liz.cont.mod3 <- glmmTMB(n_liz_hr ~ Station*mean_relative_humidity + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.cont.mod3, plot = T)
# passes validation
# AIC 279.6 (worse)
summary(liz.cont.mod3)

# temp?
liz.cont.mod4 <- glmmTMB(n_liz_hr ~ Station*mean_temp_F + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.cont.mod4, plot = T)
# doesnt pass validation
# AIC  270.5, substantial improvement! this is top mod, difference appear to be drive by temperature
summary(liz.cont.mod4)

#make sure top model (mod 1) isn't over dispersed or zero inflated
testDispersion(liz.cont.mod4, plot = F)
testZeroInflation(liz.cont.mod4, plot = F)

#looks good: Get scatterplot about lixzard tedections with temp, how different are they between day and night
```

### Figure

```{r liz plots}

liz.mean.se <- amph_lz_detections %>%
  group_by(Station) %>%
  summarise(mean = mean(n_liz_hr),
            sd = sd(n_liz_hr),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

liz.fig1 <- ggplot(liz.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = amph_lz_detections, 
              aes(x = Station, y = n_liz_hr), 
              width = 0.15, alpha = 0.2) + coord_fixed(ratio = 0.75) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21) + theme_bw() + scale_fill_manual(values = c("lightgrey", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Lizards") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma", margin = margin(t = 8)), axis.text.y = element_text(family = "Tahoma", margin = margin(l = 8)), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, face = "bold", size = 12, margin = margin(b = 12))) +
  geom_signif(
    y_position = 4, xmin = 1, xmax = 2,
    annotation = "N.S.")

liz.fig1 

# simplifying code:

liz.fig2 <- ggplot(liz.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = amph_lz_detections, 
              aes(x = Station, y = n_liz_hr), 
              width = 0.15, alpha = 0.2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21) + theme_bw() + scale_fill_manual(values = c("lightgrey", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Lizards") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma"), axis.text.y = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, face = "bold", size = 12)) +
  geom_signif(
    y_position = 4, xmin = 1, xmax = 2,
    annotation = "N.S.")

liz.fig2
```
### Distance

```{r lz mod dist}

# null mod
liz.dist.null <- glmmTMB(n_liz_500m ~ 1 + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.dist.null, plot = T)
# passes
summary(liz.dist.null)
# AIc 192.9

# start simple, only hunting status (station) and survey type
liz.dist.mod1 <- glmmTMB(n_liz_500m ~ Station + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.dist.mod1, plot = T)
# passes validation

summary(liz.dist.mod1)
# AIC 188.4  
# sig effect of station

# make model more complex, add interaction effect of hunting status with survey type, thing on left of / is fixed effect, 
liz.dist.mod2 <- glmmTMB(n_liz_500m ~ Station/Survey_Type + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.dist.mod2, plot = T)
# passes

# AIC:  191.8 (worse)
summary(liz.dist.mod2)

#drop survey type bc doesn't improve model fit

#investigate humidity
liz.dist.mod3 <- glmmTMB(n_liz_500m ~ Station/mean_relative_humidity + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.dist.mod3, plot = T)
# passes validation
# AIC 192.1 (worse)
summary(liz.dist.mod3)

# temp?
liz.dist.mod4 <- glmmTMB(n_liz_500m ~ Station*mean_temp_F + (1|Parent_Trail_LongName), data = amph_lz_detections, family = tweedie())

sim <- simulateResiduals(fittedModel = liz.dist.mod4, plot = T)

# AIC  185.5, substantial improvement! this is top mod, difference appear to be drive by temperature
summary(liz.dist.mod4)

#make sure top model (mod 1) isn't over dispersed or zero inflated
testDispersion(liz.dist.mod4, plot = F)
testZeroInflation(liz.dist.mod4, plot = F)

#looks good: Get scatterplot about lixzard tedections with temp, how different are they between day and night
```

### Figure 



# Snakes per hour

```{r snake detection wide df}
# First, need to make a wide dataframe similar to amph_lz_detections

# Have to keep zeros and fix repeat SK04 and SK05 observations

snakes.sur2 <- snakes.sur %>%
  distinct() %>%
  mutate(n = if_else(is.na(Snake_ID), 0L, 1L))

snake_counts <- snakes.sur2 %>%
  group_by(Station, Start_Date, Parent_Trail_LongName) %>%
  summarise(n_snakes = sum(n), .groups = "drop")

# sanity check to make sure correct number of observations is here 

sum(snake_counts$n_snakes)

# Okay now time to add environmental variables: 

humidity_night <- humidity_means %>%
  filter(Survey_Type == "Night")

snake_counts <- snake_counts %>%
  left_join(
    humidity_night %>%
      select(Station, Start_Date, mean_relative_humidity),
    by = c("Station", "Start_Date")
  )

temp_night <- temperature_means %>%
  filter(Survey_Type == "Night")

snake_counts <- snake_counts %>%
  left_join(
    temp_night %>%
      select(Station, Start_Date, mean_temp_F), 
    by = c("Station", "Start_Date")
  )

understory_night <- understory_means %>%
  filter(Survey_Type == "Night")

snake_counts <- snake_counts %>%
  left_join(
    understory_night %>%
      select(Station, Start_Date, mean_understory_leaves), 
    by = c("Station", "Start_Date")
  )

# Almost ready: now just need to add in the Effort column and Meters column from the enviro sheet and then get n_snakes per hour 

snake_counts <- snake_counts %>%
  left_join(
    env_24_25_nights %>%
      select(
        Station,
        Start_Date,
        Parent_Trail_LongName,
        Meters,
        Effort_hrs
      ),
    by = c("Station", "Start_Date", "Parent_Trail_LongName")
  )

snake_counts <- snake_counts %>%
  mutate(n_snakes_hr = n_snakes / Effort_hrs)


```

## Summarize

```{r snake summaries}

snake_counts %>%
  group_by(Station) %>%
  summarise(
    tot_snake = sum(n_snakes),
    mean_snake = mean(n_snakes),
            sd = sd(n_snakes),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

## Model snakes per hour

```{r snake models}

# start simple, only hunting status (station) and survey type
snake.mod1 <- glmmTMB(n_snakes_hr ~ Station + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.mod1, plot = T)
# passes validation

summary(snake.mod1)
# AIC 129.3

# Okay so station doesn't determine any significant differences I don't think

snake.mod2 <- glmmTMB(n_snakes_hr ~ Station*mean_relative_humidity + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.mod2, plot = T)
# passes validation
# AIC 130.1 but barely (dAIC = 0.8)
summary(snake.mod2)

# Humidity doesn't seem to impact detections when looking at station differences

# Time to check temp

snake.mod3 <- glmmTMB(n_snakes_hr ~ Station*mean_temp_F + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.mod3, plot = T)
# passes validation
# AIC 130.9 
summary(snake.mod3)

# Lastly understory density

snake.mod4 <- glmmTMB(n_snakes_hr ~ Station*mean_understory_leaves + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.mod4, plot = T)
# passes validation
# AIC 129.8
summary(snake.mod4)
```

# Snakes per 3 guilds per hour

## Format

```{r detections pr hr by guild split into three categories}

# effort hours
snakes.sur <- left_join(snakes.sur, effort)

# isolate surveys for which no snakes were found
zeros.3.diets <- snakes.sur %>%
  filter(is.na(Snake_ID)) %>%
  select(Start_Date, Station, Diet_Type_3categories, Parent_Trail_LongName, Effort_hrs)

# get them into the format we want (tally doesn't actually matter here)
z.tally.no.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Station, Parent_Trail_LongName, Diet_Type_3categories, Effort_hrs) %>% tally()

# for the nights when 0 snakes were found, make a dataframe of instances that specifies zero found for each diet type category

# no mammals
z.tally.no.mammals.3.diets$Diet_Type_3categories <- "no_mammals"

# mostly mammals
z.tally.mostly.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Station, Diet_Type_3categories, Parent_Trail_LongName, Effort_hrs) %>% tally()

z.tally.mostly.mammals.3.diets$Diet_Type_3categories <- "mostly_mammals"

# mixed
z.tally.mixed.3.diets <- zeros.3.diets %>%
  group_by(Start_Date,Station, Parent_Trail_LongName, Diet_Type_3categories, Effort_hrs) %>% tally()

z.tally.mixed.3.diets$Diet_Type_3categories <- "mixed"

# stack them together
z.tallies.3 <- rbind(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

#clean
rm(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

# also, figure out why the zero detections on CHIO-03 6-11-2024 are entered twice, as are 6-23 MAPO03, 6-23 MURC-01, and 6-23 PUEN-01 (all 2024) - ignoring for now bc it's an error I'm pretty sure

# remove the n column (meaningless)
z.tallies.3 <- z.tallies.3 %>%
  select(!n)

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

length(surveys.no.zeros$Snake_ID) # SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally.3 <- surveys.no.zeros %>%
  group_by(Start_Date, Station, Parent_Trail_LongName, Diet_Type_3categories, Effort_hrs) %>% tally()

# now, we can add the surveys and the segments with zero detections back

snake_guild_per_hr <- full_join(z.tallies.3, survey.tally.3)

# make NAs zeros
snake_guild_per_hr <- snake_guild_per_hr %>%
  mutate(n = ifelse(is.na(n), 0, n))

snake_guild_per_hr$snakes_hr <- snake_guild_per_hr$n/snake_guild_per_hr$Effort_hrs

rm(z.tallies.3, zeros.3.diets, survey.tally.3)
```

## Summarize

```{r summarize snakes per hr per guild}

snake_guild_per_hr %>%
  group_by(Station, Diet_Type_3categories) %>%
  summarise(
    tot_snakes_per_surv = sum(n),
    mean_snakes_per_surv = mean(n),
            sd = sd(n),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

## Model snake guilds per hour

```{r snakes per guild mod}
snakes.3.guilds.m1 <- glmmTMB(snakes_hr ~ Station/Diet_Type_3categories + (1|Parent_Trail_LongName), data = snake_guild_per_hr, family = tweedie())

sim <- simulateResiduals(fittedModel = snakes.3.guilds.m1, plot = T)
# passes validation

summary(snakes.3.guilds.m1)

Anova(snakes.3.guilds.m1)

```

## Figure

```{r snake guild figure}

snakes.per.guild.mean.se <- snake_guild_per_hr %>%
  group_by(Station, Diet_Type_3categories) %>%
  summarise(mean = mean(snakes_hr),
            sd = sd(snakes_hr),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

snake.guild.fig <- ggplot(snakes.per.guild.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = snake_guild_per_hr, 
              aes(x = Station, y = snakes_hr), 
              width = 0.15,alpha = 0.2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21) + theme_bw() + scale_fill_manual(values = c("lightgrey", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Snakes by feeding guild") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma"), axis.text.y = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5))

# Create a custom dataframe for annotations
ann_df <- data.frame(Station = c("TBS", "YRS", "TBS", "YRS", "TBS", "YRS"),
  Diet_Type_3categories = c("mixed", "mostly_mammals", "no_mammals"),          # The faceting variable
  start_group = c("TBS", "TBS", "TBS"),    # Left side of bracket
  end_group   = c("YRS", "YRS", "YRS"),    # Right side of bracket
  y_position  = c(0.9, 0.9, 2.6),       # Tailored height for each individual facet
  label       = c("N.S.", "***", "N.S.") # Custom text or stars
)

snake.guild.fig.final <- snake.guild.fig +
  geom_signif(
    data = ann_df,
    aes(xmin = 1, xmax = 2, y_position = y_position, annotations = label),
    manual = TRUE) + facet_wrap(~Diet_Type_3categories, scales = "free", labeller = labeller(Diet_Type_3categories = 
    c("mixed" = "Mixed diet",
      "mostly_mammals" = "Primarily mammal diet",
      "no_mammals" = "Non-mammal diet"))) + theme(strip.text = element_text(size = 12))

snake.guild.fig.final
```

# Snakes per 2 guilds per hour

## Format

```{r detections pr hr by guild split into two categories}

# isolate surveys for which no snakes were found
zeros.2.diets <- snakes.sur %>%
  filter(is.na(Snake_ID)) %>%
  select(Start_Date, Station, Diet_Type, Parent_Trail_LongName, Effort_hrs)

# get them into the format we want (tally doesn't actually matter here)
z.tally.no.mammals.2.diets <- zeros.2.diets %>%
  group_by(Start_Date, Station, Parent_Trail_LongName, Diet_Type, Effort_hrs) %>% tally()

# for the nights when 0 snakes were found, make a dataframe of instances that specifies zero found for each diet type category

# no mammals
z.tally.no.mammals.2.diets$Diet_Type <- "no_mammals"

# mostly mammals
z.tally.mammals.2.diets <- zeros.2.diets %>%
  group_by(Start_Date, Station, Diet_Type, Parent_Trail_LongName, Effort_hrs) %>% tally()

z.tally.mammals.2.diets$Diet_Type <- "consumes_mammals"

# stack them together
z.tallies.2 <- rbind(z.tally.no.mammals.2.diets, z.tally.mammals.2.diets)

# also, figure out why the zero detections on CHIO-03 6-11-2024 are entered twice, as are 6-23 MAPO03, 6-23 MURC-01, and 6-23 PUEN-01 (all 2024) - ignoring for now bc it's an error I'm pretty sure

# remove the n column (meaningless)
z.tallies.2 <- z.tallies.2 %>%
  select(!n)

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

length(surveys.no.zeros$Snake_ID) # SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally.2 <- surveys.no.zeros %>%
  group_by(Start_Date, Station, Parent_Trail_LongName, Diet_Type, Effort_hrs) %>% tally()

# now, we can add the surveys and the segments with zero detections back

snake_2guilds_per_hr <- full_join(z.tallies.2, survey.tally.2)

# make NAs zeros
snake_2guilds_per_hr <- snake_2guilds_per_hr %>%
  mutate(n = ifelse(is.na(n), 0, n))

snake_2guilds_per_hr$snakes_hr <- snake_2guilds_per_hr$n/snake_2guilds_per_hr$Effort_hrs

# clean
rm(z.tallies.2, zeros.2.diets, survey.tally.2)
```

## Summarize

```{r summarize snakes per hr per two feeding guilds}

snake_2guilds_per_hr %>%
  group_by(Station, Diet_Type) %>%
  summarise(
    tot_snakes_per_surv = sum(n),
    mean_snakes_per_surv = mean(n),
            sd = sd(n),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

## Model

```{r snakes per guild mod}

snakes.2.guilds.m1 <- glmmTMB(snakes_hr ~ Station/Diet_Type + (1|Parent_Trail_LongName), data = snake_2guilds_per_hr, family = tweedie)

sim <- simulateResiduals(fittedModel = snakes.2.guilds.m1, plot = T)
# passes validation

summary(snakes.2.guilds.m1)

Anova(snakes.2.guilds.m1)

# significant interaction effect

```

## Figure

```{r snake guild figure}

snakes.per.2guilds.mean.se <- snake_2guilds_per_hr %>%
  group_by(Station, Diet_Type) %>%
  summarise(mean = mean(snakes_hr),
            sd = sd(snakes_hr),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

snake.2guilds.fig <- ggplot(snakes.per.2guilds.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = snake_2guilds_per_hr, 
              aes(x = Station, y = snakes_hr), 
              width = 0.15, size = 1.5, alpha = 0.3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21, size = 3) + theme_bw() + scale_fill_manual(values = c("lightgrey", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Snakes by feeding guild") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma"), axis.text.y = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, face = "bold", size = 12))

snake.2guilds.fig 

# Create a custom dataframe for annotations
ann_df2 <- data.frame(Station = c("TBS", "YRS", "TBS", "YRS", "TBS", "YRS"),
  Diet_Type = c("consumes_mammals", "no_mammals"),          # The faceting variable
  start_group = c("TBS", "TBS", "TBS"),    # Left side of bracket
  end_group   = c("YRS", "YRS", "YRS"),    # Right side of bracket
  y_position  = c(1.5, 3),       # Tailored height for each individual facet
  label       = c("***", "N.S.") # Custom text or stars
)

snake.2guilds.fig.final <- snake.2guilds.fig +
  geom_signif(
    data = ann_df2,
    aes(xmin = 1, xmax = 2, y_position = y_position, annotations = label),
    manual = TRUE) + facet_wrap(~Diet_Type, scales = "free", labeller = labeller(Diet_Type = 
    c("consumes_mammals" = "Consumes mammals",
      "no_mammals" = "Non-mammal diet")
  )) 

snake.2guilds.fig.final
```


# Snakes per family per hour

## Format

```{r format data per family}

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

# SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally <- surveys.no.zeros %>%
  filter(Family != "Viperidae") %>%
filter(Family != "Elapidae") %>%
  group_by(Start_Date, Station, Family, Parent_Trail_LongName, Effort_hrs) %>% tally()

# generate all possible combinations based on family to get zeros

snake.family.tallies <- survey.tally %>%
  as_tibble() %>%
  complete(Family, nesting(Start_Date, Station, Effort_hrs, Parent_Trail_LongName), fill = list(n = 0))

# get rate per hour

snake.family.tallies$n_hr <- snake.family.tallies$n/snake.family.tallies$Effort_hrs

```

## Summarize & some modeling

```{r snakes per fam summary}
snake.family.tallies %>%
  group_by(Station, Family) %>%
  summarise(
    tot_per_fam = sum(n),
    mean_per_fam_per_surv = mean(n),
            sd = sd(n),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

#Extract meters and dates to mertge to snake_tallies_family
snake_surv_distances <- snake_counts %>%
  select(Start_Date, Meters)

snake.family.tallies <- left_join(snake.family.tallies, snake_surv_distances)

snake.family.tallies$n_snakes_500m <- (snake.family.tallies$n/snake.family.tallies$Meters)*500



```

## Model

```{r model dets per fam}
snakes.fam.m1 <- glmmTMB(n ~ Station/Family + (1|Parent_Trail_LongName) + offset(log(Effort_hrs)), data = snake.family.tallies, ziformula = ~1, family = nbinom2)

sim <- simulateResiduals(fittedModel = snakes.fam.m1, plot = T)
# passes validation

summary(snakes.fam.m1)

Anova(snakes.fam.m1)
```

## Figures

```{r snake per fam det fig}

snakes.per.fam.mean.se <- snake.family.tallies %>%
  group_by(Station, Family) %>%
  summarise(mean = mean(n_hr),
            sd = sd(n_hr),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

snake.fam.fig <- ggplot(snakes.per.fam.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = snake.family.tallies, 
              aes(x = Station, y = n_hr), 
              width = 0.15, alpha = 0.2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21) + theme_bw() + scale_fill_manual(values = c("lightgrey", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Snakes by family") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma"), axis.text.y = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, size = 12))

# Create a custom dataframe for annotations
ann_df2 <- data.frame(Station = c("TBS", "YRS", "TBS", "YRS"),
  Family = c("Boidae", "Colubridae"),          # The faceting variable
  start_group = c("TBS", "TBS"),    # Left side of bracket
  end_group = c("YRS", "YRS"),    # Right side of bracket
  y_position = c(1, 2.6),       # Tailored height for each individual facet
  label = c("***", "N.S.") # Custom text or stars
)

snake.fam.fig.final <- snake.fam.fig +
  geom_signif(
    data = ann_df2,
    aes(xmin = 1, xmax = 2, y_position = y_position, annotations = label),
    manual = TRUE) + facet_wrap(~Family, scales = "free") + theme(strip.text = element_text(size = 12))

```

# Detections Panel Figure

```{r panel detections figure}

# ok so we have frog.fig1, liz.fig1, smamm.fig, snake.fam.fig.final, snake.guild.fig.final


panel1<-ggarrange(snake.fam.fig.final, snake.guild.fig.final, labels = c("(a)", "(b)"), widths = c(1,2))

panel1

ggsave(panel1, filename = "Figures/snake_detections.png", height = 2.5, width = 5)

panel2<-ggarrange(frog.fig2, liz.fig2, smamm.fig, labels = c("(c)", "(d)", "(e)"), nrow =1)

panel2

#ggsave(panel2, filename = "Figures/prey_detections.png", height = 2.5, width = 5)
```


# Snakes per 500m

## Overall regardless of species

```{r mod overall snake detections per 500m}

# Add in columns for detections per 500m 
snake_counts$n_snakes_500m <- (snake_counts$n_snakes/snake_counts$Meters)*500

# look at data
hist(snake_counts$n_snakes_500m) # tweedie should be ok

# null
snake.dist.null <- glmmTMB(n_snakes_500m ~ 1 + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.null, plot = T)
# passes

summary(snake.dist.null)
# AIC 67.3

# mod 1 Station
snake.dist.mod1 <- glmmTMB(n_snakes_500m ~ Station + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.mod1, plot = T)
# passes

summary(snake.dist.mod1)
# AIC 69.2

# Station makes model worse, so omit

# Investigate temperature
snake.dist.mod2 <- glmmTMB(n_snakes_500m ~ mean_temp_F + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.mod2, plot = T)
# passes validation

summary(snake.dist.mod2)
# AIC 68.8
# doesn't improve fit

# Investigate humidity
snake.dist.mod3 <- glmmTMB(n_snakes_500m ~ mean_relative_humidity + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.mod3, plot = T)
# passes validation

summary(snake.dist.mod3)
# AIC 68.5

# Null is actually the best fit model for overall snake detections regardless of species

```

## Per family

```{r mod snakes per 3 guilds}

# null
#snake.dist.null <- glmmTMB(n_snakes_500m ~ 1 + (1|Parent_Trail_LongName), data = snake.family.tallies, ziformula = ~ 1, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.null, plot = T)
# doesn't pass

summary(snake.dist.null)

# 

# mod 1 Station
snake.dist.mod1 <- glmmTMB(n_snakes_500m ~ Station/Family + (1|Parent_Trail_LongName), data = snake.family.tallies, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.mod1, plot = T)
# passes

summary(snake.dist.mod1)
# AIC 50.1

Anova(snake.dist.mod1)

testDispersion(snake.dist.mod1, plot = F)
testZeroInflation(snake.dist.mod1, plot = F)




# Station makes model worse, so omit

# Investigate temperature
snake.dist.mod2 <- glmmTMB(n_snakes_500m ~ mean_temp_F + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.mod2, plot = T)
# passes validation

summary(snake.dist.mod2)
# AIC 68.8
# doesn't improve fit

# Investigate humidity
snake.dist.mod3 <- glmmTMB(n_snakes_500m ~ mean_relative_humidity + (1|Parent_Trail_LongName), data = snake_counts, family = tweedie())

sim <- simulateResiduals(fittedModel = snake.dist.mod3, plot = T)
# passes validation

summary(snake.dist.mod3)
# AIC 68.5

```

# Figure 

```{r snake per fam det fig}

snakes.per.fam.mean.se <- snake.family.tallies %>%
  group_by(Station, Family) %>%
  summarise(mean = mean(n_snakes_500m),
            sd = sd(n_snakes_500m),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

snake.fam.fig <- ggplot(snakes.per.fam.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = snake.family.tallies, 
              aes(x = Station, y = n_snakes_500m), 
              width = 0.15, alpha = 0.2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21) + theme_bw() + scale_fill_manual(values = c("lightgrey", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/500m") + ggtitle("Snakes by family") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma"), axis.text.y = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, size = 12))

snake.fam.fig + facet_wrap( ~ Family, scales = "free")

# Create a custom dataframe for annotations
ann_df2 <- data.frame(Station = c("TBS", "YRS", "TBS", "YRS"),
  Family = c("Boidae", "Colubridae"),          # The faceting variable
  start_group = c("TBS", "TBS"),    # Left side of bracket
  end_group = c("YRS", "YRS"),    # Right side of bracket
  y_position = c(1, 2.6),       # Tailored height for each individual facet
  label = c("***", "N.S.") # Custom text or stars
)

snake.fam.fig.final <- snake.fam.fig +
  geom_signif(
    data = ann_df2,
    aes(xmin = 1, xmax = 2, y_position = y_position, annotations = label),
    manual = TRUE) + facet_wrap(~Family, scales = "free") + theme(strip.text = element_text(size = 12))

```





# Density estimates

Note I think it's a good idea to follow the methods in this paper: Hierarchical distance sampling to estimate population sizes of common lizards across a desert ecoregion (in Zotero shared library): OWEN sounds like a plan I can read that this weekend. 

## Binning

First, create distance bins. Also look at some things.

```{r review data more} 

# pull out year to make a way to distinguish between seasons

snakes.sur$Year <- factor(format(snakes.sur$Start_Date, "%Y"))

snakes.sur %>%
  group_by(Year, Station) %>%
    reframe(N_surveys = length(unique(Start_Date))) #Effort comparisons between stations

# Deciding on distance bins in distance sampling involves selecting cutpoints that adequately capture the shape of the detection function, typically ensuring more bins near the line (where detection probability changes rapidly) and fewer further away. The goal is to minimize bias from data rounding while ensuring enough observations per bin for accurate model fitting. 

# Smaller bins closer to zero

# Truncate your data to remove the farthest detections - WE CANNOT DO THIS DUE TO THE BOA BIAS THAT OWEN POINTED OUT

# First, look at a histogram of distance from trail to get an idea of detection distribution

hist(snakes.sur$Distance_From_Trail_m, breaks = seq(0, 12, by = 0.5))

```


```{r divide up bins broader}

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 1, "0_1", NA))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 2, "1_2", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 3, "2_3", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 3 & Distance_From_Trail_m <= 4, "3_4", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 4 & Distance_From_Trail_m <= 5, "4_5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 5, "5_8", Distance_bin_m))

```


## 3 Diet types

Mixed, primarily mammal consumers, non-mammal consumers

```{r format for unmarked 3 diet types}

# isolate surveys for which no snakes were found
zeros.3.diets <- snakes.sur %>%
  filter(is.na(Snake_ID)) %>%
  select(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m)

# get them into the format we want (tally doesn't actually matter here)
z.tally.no.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

# for the nights when 0 snakes were found, make a dataframe of instances that specifies zero found for each diet type category

# no mammals
z.tally.no.mammals.3.diets$Diet_Type_3categories <- "no_mammals"

# mostly mammals
z.tally.mostly.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

z.tally.mostly.mammals.3.diets$Diet_Type_3categories <- "mostly_mammals"

# mixed

# mixed
z.tally.mixed.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

z.tally.mixed.3.diets$Diet_Type_3categories <- "mixed"

# stack them together
z.tallies.3 <- rbind(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

#clean
rm(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

# also, figure out why the zero detections on CHIO-03 6-11-2024 are entered twice, as are 6-23 MAPO03, 6-23 MURC-01, and 6-23 PUEN-01 (all 2024) - ignoring for now bc it's an error I'm pretty sure

# remove the n column (meaningless)
z.tallies.3 <- z.tallies.3 %>%
  select(!n)

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

length(surveys.no.zeros$Snake_ID) # SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally.3 <- surveys.no.zeros %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m, Distance_bin_m) %>% tally()

# pivot wider that wider
density_est_df.3 <- survey.tally.3 %>% 
  pivot_wider(names_from = Distance_bin_m,
              values_from = n, 
              values_fill = 0)

# now, we can add the surveys and the segments with zero detections back

density_est_df.3 <- full_join(z.tallies.3, density_est_df.3)

# make NAs zeros - except the temp and humidty NAs

density_est_df.3 <- density_est_df.3 %>%mutate_all(~replace(., is.na(.), 0))

# clean
rm(survey.tally.3, surveys.no.zeros, z.tallies.3, zeros.3.diets)
```

```{r make umf three diet types}

density_est_df.3$Year <- as.factor(density_est_df.3$Year)
density_est_df.3$Diet_Type_3categories <- as.factor(density_est_df.3$Diet_Type_3categories)
density_est_df.3$Seg_dist_m <- as.numeric(density_est_df.3$Seg_dist_m)
density_est_df.3$Segment <- as.factor(density_est_df.3$Segment)
density_est_df.3$Station <- as.factor(density_est_df.3$Station)

# separate out the covariates
covs.3 <- density_est_df.3[,c("Station", "Year", "Diet_Type_3categories", "Segment", "Seg_dist_m")]

umf.3 <- unmarkedFrameDS(y=as.matrix(density_est_df.3[,7:ncol(density_est_df.3)]), siteCovs=data.frame(covs.3), dist.breaks=c(0, 1, 2, 3, 4, 5, 8), unitsIn="m", survey="line", tlength=covs.3$Seg_dist_m)

hist(umf.3, xlab="distance (m)", main="", cex.lab=0.8, cex.axis=0.8)

```

### GOF function

Goodness of fit tests

```{r gof function}

# using this resource: https://www.montana.edu/screel/teaching/bioe-440r-521/bioe-440-distance-sampling.html

# Goodness of fit test: good fit if results of these tests show p > 0.05. First, fit a model: use your best-supported model (Lowest AIC). Then check GOF for that model.

# Function returning three fit-statistics.

fitstats <- function(fm) {

  observed <- getY(fm@data)

  expected <- fitted(fm)

  resids <- residuals(fm)

  sse <- sum(resids^2)

  chisq <- sum((observed - expected)^2 / expected)

  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2)

  out <- c(SSE=sse, Chisq=chisq, freemanTukey=freeTuke)

  return(out)

}

```


## Model 3 Diet Types

```{r mods three diet types no env covs}

null.3 <- distsamp(~1 ~ 1, umf.3, unitsOut = "kmsq")
summary(null.3)

Diet_det <- distsamp(~ Diet_Type_3categories ~ 1, umf.3, unitsOut = "kmsq")
summary(Diet_det)# diet type as detection covariate improves model

Year.3 <- distsamp(~ Diet_Type_3categories ~ Year, umf.3, unitsOut = "kmsq")
summary(Year.3)
# doesn't significantly improve fit (< 2 AIC)

Station.3 <- distsamp(~Diet_Type_3categories ~ Station, umf.3, unitsOut = "kmsq")
summary(Station.3)
# station makes things worse

StationXGuild.3 <- distsamp(~Diet_Type_3categories ~ Station/Diet_Type_3categories, umf.3, unitsOut = "kmsq")
summary(StationXGuild.3)
# Make a table to compare models

# put mods in a list
mod.set.3diets <- list(null.3 = null.3, Year.3 = Year.3, StationXGuild.3 = StationXGuild.3, Station.3 = Station.3, Diet_det = Diet_det)

# make table
aic_table.3 <- aictab(cand.set = mod.set.3diets, sort = TRUE)

aic_table.3

```

### GOF 3 diet types

```{r goodness of fit three guilds}
#pb_3cats <- parboot(StationXGuild.3, fitstats, nsim=1000)
#pb_3cats
# fine
```


## Predicted vals 3 diet types

```{r predicted values three guilds}

new.data3 <- density_est_df.3 %>%
  ungroup() %>%
  dplyr::select(Station, Diet_Type_3categories) %>%
  distinct()

#Predict

predictions3 <- predict(StationXGuild.3, type="state", newdata=new.data3, appendData=TRUE)

```

```{r plot three guilds, message=FALSE, warning=FALSE}
predictions3$Diet_Type_3categories <- factor(
  predictions3$Diet_Type_3categories,
  levels = c("no_mammals", "mixed", "mostly_mammals", 
  labels = c("No Mammals", "Mixed", "Mostly Mammals"))
) #make sure that the order goes nomamm > mixed > mostly mamm

p_3cats<-ggplot(predictions3, aes(x=Station, y=Predicted, fill = Diet_Type_3categories, group = Station)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = .2) + geom_point(pch = 21, size = 4) + theme_bw() + scale_fill_manual(values = c("goldenrod1", "green4", "black")) + xlab("") + scale_color_discrete(labels = "") +  scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake Density Estimates by Dietary Guild") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(text = element_text(family = "Tahoma")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12, margin = margin(b = 10))) + scale_fill_manual(
  name = "Diet Type",
  values = c("goldenrod1", "green4", "black"),
  labels = c("No mammals", "Mixed", "Mostly mammals")
) + facet_wrap(~Diet_Type_3categories, scales = "free", labeller = labeller(Diet_Type_3categories = c(
    no_mammals = "No Mammals",
    mixed = "Mixed",
    mostly_mammals = "Mostly Mammals"
  ))
)

p_3cats

#ggsave(p_3cats, file="snk_dens_3diet.png", width = 8, height = 5)


```

Appears to be elevated numbers in hunted forest in the guild that is mostly mammal consumers.


## Family level

The two most detected families are Boidae and Colubridae. Limitting the analysis to these 2 families allows the models to run without error with Family as a covariate. Including all families detected creates model issues with uncertainty because the other families were only detected 1 or 2 times. 

```{r format for unmarked based on fams}

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 1, "0_1", NA))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 2, "2", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 4, "2_4", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 4, "4_8", Distance_bin_m))

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

# SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally <- surveys.no.zeros %>%
  filter(Family != "Viperidae") %>%
filter(Family != "Elapidae") %>%
  group_by(Start_Date, Year, Station, Family, Segment, Seg_dist_m, Distance_bin_m) %>% tally()

# generate all possible combinations based on family to get zeros

df_long <- survey.tally %>%
  as_tibble() %>%
  complete(Family, Distance_bin_m, nesting(Start_Date, Year, Station, Segment, Seg_dist_m), fill = list(n = 0))

# also, which families eat what?
snakes.sur %>%
  group_by(Family) %>%
  reframe(unique(Diet_Type)) # colubridae is only possibility for mixed in our data?

# pivot wider that wider
density_fam_df <- df_long %>% 
  pivot_wider(names_from = Distance_bin_m,
              values_from = n, 
              values_fill = 0)
```


## Model family level density

```{r model at fam level}
# separate out the covariates

# model not running bc of sparse data at fam level - removed elapidae, which only had 1 detection, and Viperidae which only had 2

covs.fam <- density_fam_df[,c("Station", "Year", "Family", "Segment", "Seg_dist_m")]

umf.fam <- unmarkedFrameDS(y=as.matrix(density_fam_df[,7:ncol(density_fam_df)]), siteCovs=data.frame(covs.fam), dist.breaks=c( 0,1,2,4,8), unitsIn="m", survey="line", tlength=covs.fam$Seg_dist_m)

# run models
# null
null.fam <- distsamp(~1 ~ 1, umf.fam, unitsOut = "kmsq")
summary(null.fam)

# family as detection covariate, no density covariate
fam.null  <- distsamp(~Family ~ 1, umf.fam, unitsOut = "kmsq")
summary(fam.null)

# add year as density covariate
year <- distsamp(~ Family ~ Year, umf.fam, unitsOut = "kmsq")
summary(year)

# add station as density covariate
station <- distsamp(~Family ~ Station, umf.fam, unitsOut = "kmsq")
summary(station)

# add station and station x family interaction term 

station_stationXfam <- distsamp(~Family~ Station/Family, umf.fam, unitsOut = "kmsq")

summary(station_stationXfam)
# best mod by far

# put mods in a list
mod.set.family <- list(null.fam = null.fam, fam.null = fam.null, year = year, station = station, station_stationXfam = station_stationXfam)

# make table
aic_table.fam <- aictab(cand.set = mod.set.family, sort = TRUE)

aic_table.fam
```


## Predicted vals fam

```{r predicted values three guilds}

new.data.fam <- density_fam_df %>%
  ungroup() %>%
  dplyr::select(Station, Family) %>%
  distinct()
  

#Predict

predictions.fam <- predict(station_stationXfam, type="state", newdata=new.data.fam, appendData=TRUE)


```


```{r plot fam, message=FALSE, warning=FALSE}

p_3cats_fam<-ggplot(predictions.fam, aes(x=Station, y=Predicted, fill = Family, group = Station)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = .2) + geom_point(pch = 21) + theme_bw() + scale_fill_manual(values = c("purple2","goldenrod1", "black")) + xlab("") + scale_color_discrete(labels = "") +  scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake Density Estimates by Family") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(text = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12)) 

p_3cats_fam

#ggsave(p_3cats_fam, file="snk_dens_3diet_fam.png", width = 8, height = 5)
```

Appears to be elevated numbers in hunted forest in the family that is mostly mammal consumers.


# Community composition

## Pie Charts

Below I mess around with some basic community composition visualizations.

```{r species occurrences}

# just get an idea of n detections from all survey types per station, including the instances with no ID
raw_count <- snakes.master %>%
  group_by(Station) %>%
  reframe(n = n())

# total at YRS is 111, at TBS is 108 (great, super comparable numbers when pooled)

# subfamily level may help
# create this column and populate it:
snakes.master$Subfamily <- NA
snakes.master <- snakes.master %>%
  dplyr::mutate(Subfamily = ifelse(Genus == "Siphlophis" | Genus == "Imantodes" | Genus == "Dipsas" | Genus == "Drepanoides" | Genus == "Oxyrhopus" | Genus == "Leptodeira" | Genus == "Hydrops" | Genus == "Helicops" | Genus == "Atractus" | Genus == "Erythrolamprus" | Genus == "Taeniophallus" | Genus == "Pseudoboa", "Dipsadinae", 
                            ifelse(Genus == "Epicrates" | Genus == "Corallus" | Genus == "Eunectes", "Boinae",
                                   ifelse(Genus == "Chironius" | Genus == "Drymobius" | Genus == "Tantilla" | Genus == "Drymarchon" | Genus == "Oxybelis" | Genus == "Spilotes" | Genus == "Dendrophidion" | Genus =="Leptophis", "Colubrinae",
                                          ifelse(Genus == "Bothrops" | Genus == "Bothrocophias", "Crotalinae",
                                                 ifelse(Genus == "Xenoxybelis", "Xenodontinae",
                                                        ifelse(Genus == "Micrurus", "Elapinae", Subfamily)))))))

#Siphlophis = Dipsadinae
#Imantodes = Dipsadinae 
#Epicrates = Boinae
#Dipsas = Dipsadinae
#Drepanoides = Dipsadinae  
#Chironius = Colubrinae
#Corallus = Boinae
#Oxyrhopus = Dipsadinae
#Leptodeira = Dipsadinae
#Hydrops = Dipsadinae
#Bothrops = Crotalinae
#Xenoxybelis = Xenodontinae
#Drymobius = Colubrinae
#Micrurus = Elapinae
#Helicops = Dipsadinae
#Eunectes = Boinae   
#Atractus = Dipsadinae
#Tantilla = Colubrinae
#Drymarchon = Colubrinae
#Erythrolamprus = Dipsadinae
#Taeniophallus = Dipsadinae
#Bothrocophias = Crotalinae
#Spilotes = Colubrinae
#Pseudoboa = Dipsadinae
#Oxybelis = Colubrinae
#Dendrophidion = Colubrinae
#Leptophis = Colubrinae
#Bothrocophias = Crotalinae

# make an edge list of n times an individual was seen at each station for each subfamily group
sub.fam.edgelist <- snakes.master %>%
  group_by(Station, Subfamily) %>%
  tally() %>%
  na.omit() 

# piechart
ggplot(sub.fam.edgelist, aes(x="", y= n, fill=Subfamily)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) + # remove background, grid, numeric labels
  scale_fill_brewer(palette="Set1") # remove background, grid, numeric labels

# only the survey snake data, at subfamily level
surveys <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Subfamily) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys, aes(x="", y= n, fill=Subfamily)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) + 
  scale_fill_brewer(palette="Set1") 

# Genus level, all detection methods:
all.genus <- snakes.master %>%
  group_by(Station, Genus) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(all.genus, aes(x="", y= n, fill=Genus)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station)

# Genus level, surveys only:
surveys.genus <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Genus) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys.genus, aes(x="", y= n, fill=Genus)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) 

# family

surveys.family <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Family) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys.family, aes(x="", y= n, fill=Family)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) 
```


## All records community composition

### Jaccard PERMANOVA 

```{r jaccard permanova}

# going to re-run this since some snakes were missing previously
snakes.master$ID <- paste(snakes.master$Station, snakes.master$Date, snakes.master$Snake_ID, sep = "//")

edgelist.full <- snakes.master %>%
  filter(Genus != "unknown") %>%
  group_by(Station, Species, Date, ID) %>%
  tally() %>%
  na.omit()

# now, let's use dplyr again to transpose (flip) your sp.abun dataframe so that now, you have the species as column names and the number of times they were found at each trail and date will be the data reported (expect lots of 0s and 1s)

snake.occ.mat <- edgelist.full %>% 
  pivot_wider(names_from = Species, #makes the column into the rownames
              values_from = n, 
              values_fill = 0) %>% #will fill spaces with no snakes as 0
  column_to_rownames(var = "ID") #will make this new combo the unique identifier as rownames

community <- snake.occ.mat[, c(3:42)] #made matrix from all data columns with n snake data

predictors <- snake.occ.mat[, c(1,2)] #separating non-snake stuff (station and zone)

j.dis.mat <- vegdist(community, method = "jaccard") #jaccard only accounts for presence absence 

pnova.j <- adonis2(j.dis.mat ~ predictors$Station) #adonis looks at specific differences between groups, our group factor is station

# look at results
print(pnova.j) # marginally sig/nonsig
```

### Jaccard NMDS

```{r jaccard nmds visual}

```

## Survey records community composition

### Jaccard PERMANOVA

### Jaccard NMDS

### Horn-Morisita PERMANOVA

### Horn-Morisita NMDS

# Diversity

```{r diversity & hill numbers}

# Diversity

# need data in a new species x site matrix
unique(snakes.sur$Species)

temp <- snakes.sur %>%
  group_by(Station, Species) %>% tally()

temp <- temp %>%
  na.omit()

# Need to modify the matrix so we have an n column, etc.


# get the new format, species as row names, station as columns
div.mat <- temp %>% 
  pivot_wider(names_from = Station, 
              values_from = n, 
              values_fill = 0) %>%
  column_to_rownames(var = "Species")

# now we use div.mat to get the diversity and 95% CIs
# new function uses iNEXT package (confusingly also the name of the function)
species.diversity = iNEXT(div.mat, q = c(0,1,2), datatype = "abundance", nboot = 200)

# extract values we need
species.diversity.output <- species.diversity[["iNextEst"]][["coverage_based"]]

div.vals <- species.diversity.output %>% 
  filter(Assemblage == "TBS" & m == 100 | Assemblage == "YRS" & m == 100)

# plot diversity
diversity.plot <- ggplot(data = div.vals, aes(x = Assemblage, y = qD, fill = as.factor(Order.q), shape = as.factor(Order.q))) + theme_light() + geom_pointrange(aes(ymin = qD.LCL, ymax = qD.UCL), position = position_dodge2(width = 0.35), color = "black", size = 0.6) + xlab(NULL) + ylab("Effective Number of Species") + scale_x_discrete(labels = c("Unhunted", "Hunted")) + scale_fill_manual(values = c("red", "goldenrod", "cornflowerblue"), labels = c("q = 0\n(richness)", "q = 1\n(exp. Shannon)", "q = 2\n(inv. Simpson)")) + coord_fixed(ratio = 0.1) + scale_shape_manual(values = c(21, 22, 23)) +
  theme(axis.title.y = element_text(size = 11, colour = "black", family = "Tahoma", 
                                    margin = margin(r = 10))) +
  theme(axis.title.x = element_text(size = 11, colour = "black", family = "Tahoma")) +
  theme(axis.text = element_text(size = 11, colour = "black", family = "Tahoma")) +
  theme(strip.text.x = element_text(size = 11, colour = "black", family = "Tahoma")) + 
  theme(legend.title = element_blank()) + theme(legend.text = element_text(size = 10)) +
  theme(plot.title = element_text(size = 11, colour = "black", family = "Tahoma", 
                                  hjust = .5, face = "bold")) + theme(legend.position = "bottom") +
  guides(fill = guide_legend(override.aes = list(shape = c(21, 22, 23)))) + guides(shape = "none") + ggtitle(str_wrap("Station-Level Diversity Measured with Hill Numbers (q = 0, 1, 2)", width=40))

diversity.plot

#ggsave("snk_diversity_plot.png", plot = diversity.plot, width = 6, height = 6.5)

```

Species accumulation curves

```{r accum curves}

ggiNEXT(species.diversity, type=1, facet.var="Order.q", grey=TRUE) 

```

Right so no difference there.
