---
title: "Snake_Analysis"
author: "Owen McCool, Therese Lamperty"
date: "2025-07-16"
output: word_document
---

# Code Setup & package read in 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read in packages}
library(vegan)
library(stringr)
library(ggplot2)
library(tidyverse)
library(iNEXT) # species diversity
library(hillR) # species diversity
library(DHARMa) # checking models
library(car) # reporting model stats
library(glmmTMB)
library(unmarked) #density estimates with hierarchical distance sampling models
library(tidyr)
library(AICcmodavg) # has convenient functions for comparing models based on AIC

```

# Project Overview

Study Site: Research was split between two different zones within the greater Yasuni Biosphere Reserve. Tiputini Biodiversity Station (TBS) served as the faunally intact site, whereas Yasuni Research Station (YRS) served as the defaunated site. Both sites are situated along the Tiputini River within the province of Orellana, Ecuador. TBS, approximately 45 km east of YRS, is only accessible by boat. YRS is along the Maxus Oil Road, and is less integrated into the Amazonian ecosystem. Additionally, unsustainable hunting activities by the indigenous communities have been prominent in the surrounding region at YRS for multiple decades, as noted from 2014. This is thought to be the root cause of defaunation in this zone. 

	I transferred between the two research stations every two weeks, starting my first rotation at TBS on June 10. Time at each station was spent working on 3 separate projects: my herpetofauna surveys, frugivory surveys, and an overarching seed enemy and dispersal project. 

	The rainy season in Northeastern Ecuador is at its peak between March and July. I conducted my study between the months of June and August, 2024. The average high temperature for June in this region is 85.66° F, with the average low of 70.05° F. July has an average high temperature of 86.29° F and an average low of 69.13° F. August has an average high temperature of 91.35° F and an average low of 69.60° F. June receives 143mm of precipitation on average each year, whereas July receives 99.69mm and August receives 77.55mm. (https://weatherandclimate.com/ecuador/orellana/boca-tiputini) 
Put project description here (field sites, time frame of collection,sampling scheme, type of data (eg. species ID,temp, etc.) )

Surveys: Herpetofauna were sampled using visual surveys on foot by two surveyors/night. Trail systems at each site served as the sampling paths. The surveys were repeated around 6 times on each rotation while in the field, and resampled on the second rotations. The surveys typically began around 20:00-20:30 and lasted until 23:00-00:00 CST. At each site, habitat matches were compared and assessed and trails through that habitat were consistently walked. Two different types of habitat, Terra Firma (uplands) and Varzea (flooded forest) were surveyed on the herp nocturnal walks. In addition to this, two significant bodies of water were also surveyed at each site, primarily consisting of oxbow lakes with one year-round large pond. 

	Surveyors used FENIX HM60R headlamps while conducting night surveys. They carefully examined all surrounding habitat, including understory trees, shrubs, vines, water bodies, forest floor, and middle-canopy. Amphibians of the order Anura gave off reflective eyeshine and were easier to sample for, whereas members of Squamata did not. More information about this will be within the discussion section. Effort hours will be calculated based on survey time length and number of surveyors. 

	Additionally, environmental information was taken throughout the course of each survey. Temperature and humidity data was collected with a Mengshen meter in 30 minute intervals on surveys. Understory vegetation density was also measured every 15 minutes along each trail. A 1.5m PVC pipe was randomly placed upright just off the right side of the trail at each measuring point, and the number of leaves making contact with the pipe was counted. 

	Weather and wind were noted before each survey in four nominal categories, with weather either being “clear, partly cloudy, cloudy, rainy” and wind being “no wind, felt on face, branches move, trees move”. I also accounted for precipitation that had already occurred in 6-hour blocks as “time window of last rain event”. 

	Once an animal was found, the first step of the protocol was to note what time of night the find occurred. After this, the temperature and humidity was once again collected. A rangefinder was used to estimate the perpendicular distance of each animal from the trail. A GPS was used to mark and log the coordinates of each animal found. Animals were given a two-letter code based on the Order and Suborder followed by a number, indicating its position of being found. For example, the code of the 9th snake found of the season would be SK09. Species found were also identified to the lowest taxonomic level in the field, and photographs of individuals were taken both to get positive IDs later in the lab when not possible in the field, but also to avoid counting re-captures.

	Snakes found during non-survey times were also recorded. These individuals were split into one of two categories of detection. The first is an “Opportunistic Encounter” (OE), which is any snake that was found by me while out working on other non-herpetofauna projects in the field. The other type of detection is the Third Party Encounter (TPE). This was any snake that was found by other research groups during my rotations at each station, identified to lowest possible taxonomy. The same data was taken for these individuals. GPS coordinates, time of detection, trail, observer, estimation of distance from trail, and habitat type. Because the third-party observers often didn’t have access to the climate data logger, this information was retrieved later on from each station’s respective weather tower data, based on the time of detection. 

Snakes were given a classification as to whether they were small mammal consumers or not. Information on this was retrieved from literature such as online publications, the book Reptiles and Amphibians of the Amazon by R.D. and Patricia Bartlett, and the expert field book Reptiles of Ecuador. 

# Read in and prep data

## Read in

This data was cleaned some in initial steps in 2024-2025 (see archived code). Note that temp and humidity data was removed for 4 surveys at YRS that took place after July 11, 2025, during which the team lost confidence in the data collection instrument. 

```{r data read in}
snakes.master <- read.csv(file = "snake.master.csv")
enviro <- read.csv(file = "enviro.csv")
segs <- read.csv("Segment_Distances.csv")
```

## Data prep

We need to make a dataframe that contains the dates that each segment was surveyed, preferably also merged with the length of each of those segments.

```{r segment master list}

segs_expanded <- enviro %>%
  select(!c(6,9,11:45)) %>%
  separate_rows(Order, sep = ",")

# get ready to add the meters per segment. Need names to match
segs <- segs %>%
  rename(Segment = SEGMENT) %>%
  rename(Parent_Trail = PARENT.TRAIL) %>%
  rename(Seg_dist_m = DISTANCE_M)

# and fix the names in segs_expanded to match and to be easy to understand
segs_expanded <- segs_expanded %>%
  rename(Segment = Order) %>%
  rename(Parent_Trail_LongName = Trail_Walked)

# change this in enviro as well
enviro <- enviro %>%
  rename(Parent_Trail_LongName = Trail_Walked)

# join them together
segments_master <- left_join(segs_expanded, segs)

# there are some missing distances
# check for typos
unique(segs_expanded$Segment)

# there's just one where there is a hyphen separating two segments that need to be in separate rows
segs_expanded <- segs_expanded %>%
  mutate(Segment = ifelse(Segment == "NUMA-02-CHIO-03", 
                          "NUMA-02, CHIO-03", Segment)) %>%
  separate_rows(Segment, sep = ",")

# remove all the spaces
segs_expanded <- segs_expanded %>%
  mutate(Segment = trimws(Segment))

# fix one case where there is just a space between two segments that need to be in separate rows
segs_expanded <- segs_expanded %>%
  separate_rows(Segment, sep = " ")

# okay this looks more promising, try to join again
segments_master <- left_join(segs_expanded, segs)

# beautiful
```

Okay now other little things with the datasets

```{r df prepping}

# make dates dates
snakes.master$Start_Date <- mdy(snakes.master$Start_Date)

snakes.master$Date <- mdy(snakes.master$Date)

enviro$Start_Date <- mdy(enviro$Start_Date)

segments_master$Start_Date<-mdy(segments_master$Start_Date)
segments_master$End_Date<-mdy(segments_master$End_Date)
# modify diet type to be binary = eats mammals, does not eat mammals

# mixed = "includes mammals"
snakes.master <- snakes.master %>%
  mutate(Diet_Type = ifelse(Diet_Type == "mixed", 
                          "includes_mammals", Diet_Type)) 
# mostly mammals = includes mammals
snakes.master <- snakes.master %>%
  mutate(Diet_Type = ifelse(Diet_Type == "mostly_mammals", 
                          "includes_mammals", Diet_Type))

# sampling effort
snakes.sur <- snakes.master %>%
  filter(Type_of_Encounter == "Survey") #piping with dplyr, make new dataframe with only surveyed snakes since those are only obs. with effort associated (not sure if this is needed since effort hours is now its own column)

#Now we want to put in an environment sheet because it has our effort hours, and we want to convert those to usable numeric values

# NOTE this is the time spent actively searching, not the time difference between start and stop of survey for the night

enviro$Effort_hrs <- sapply(strsplit(enviro$Searching_Time, ":"), function(x) {
  as.numeric(x[1]) + as.numeric(x[2]) / 60
}) #not really sure how this function works (from internet) but it makes the hours go from 2:30 to 2.50 hours, etc. 

# join date and time columns first
# note I added end date columns manually in the enviro datasheet to account for the 2 surveys that went past midnight
enviro$Survey_Start_Date_Time <- paste(enviro$Start_Date, enviro$Survey_Start_Time)

# use lubridate to make these date and time in R
enviro$Survey_Start_Date_Time <-ymd_hm(enviro$Survey_Start_Date_Time)

# repeat with end date and time
enviro$Survey_End_Date_Time <- paste(enviro$End_Date, enviro$Survey_End_Time)

enviro$Survey_End_Date_Time <-mdy_hm(enviro$Survey_End_Date_Time)

# small mammals 
 # add the n small mammals found per survey to the snake.sur dataframe because that has the effort hours already in it

# subset data - just want the n small mammmals and keep the date column to join by

small.mamms.temp <- enviro %>%
  select(Station, Start_Date, Number_of_Small_Mammals_Spotted, Parent_Trail_LongName) #creating temporary dataframe with just these parameters

small.mamms.temp$Number_of_Small_Mammals_Spotted <-  as.numeric(small.mamms.temp$Number_of_Small_Mammals_Spotted) #changing the n small mammals to a number from a character (idk why it was a character)
class(small.mamms.temp$Number_of_Small_Mammals_Spotted) #looks correct now

# isolate effort hours per survey
effort <- enviro %>%
  select(Start_Date, Effort_hrs)

# join effort to small mamms
small.mamms.temp <- left_join(small.mamms.temp, effort)

# clean 
rm(effort)

# get rate of n mammals per hour
small.mamms.temp$smammals.hr <- small.mamms.temp$Number_of_Small_Mammals_Spotted/small.mamms.temp$Effort_hrs

# small mammal dataframe done

# add segment lengths to snake sur df
snakes.sur <- snakes.sur %>%
  rename(Segment = Location)
```

```{r check and fix tricky join}

length(unique(segments_master$Segment)) # 89 unique segments surveyed
length(unique(segments_master$Start_Date)) #61 surveys

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Segment)) # snakes were found on 46 of them

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Snake_ID)) # 101 snakes found

length(unique(snakes.sur$Start_Date)) #60, what date is in segments master but not in snakes.sur?

missing_startdate <- anti_join(segments_master, snakes.sur, by = "Start_Date")
 
# MAPO-01 MAPO-02 HARP-01 on 2024-06-21 is in segments master but not snakes.sur. Do we know why?

# okay so do we have distances for all segments snakes were found on?

#missing_segs <- anti_join(segments_master, snakes.sur, by = "Segment")
# 157 instances of segments being surveyed and no snakes were found?

# also, in theory, segments master has all the cases that a survey occurred, and so I can remove the cases of surveys with no snakes found from the snakes.sur dataframe before merging (double check with owen that this sounds right)

snakes.sur.zeros.dropped <- snakes.sur %>%
 filter(!is.na(Snake_ID))

snakes.sur.zeros.dropped <- snakes.sur.zeros.dropped %>%
  select(Start_Date, Station, Segment, Snake_ID, Distance_From_Trail_m, Diet_Type, Family, Genus, Species, Common_Name)

snakes.sur <- full_join(segments_master, snakes.sur.zeros.dropped, relationship = "many-to-many")

length(unique(snakes.sur$Segment)) # 89 unique segments surveyed

length(unique(snakes.sur$Start_Date)) #61 surveys

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Segment)) # snakes were found on 46 of them

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Snake_ID)) # 101 snakes found

# all parts are here, problem fixed

rm(snakes.sur.zeros.dropped, missing_startdate)

```

# Data summaries

## Sampling Time and Distance

```{r survey samp effort}

# pull out mean and standard deviation overall and for each station
mean(enviro$Effort_hrs)
sd(enviro$Effort_hrs)

# per station - this will give output at bottom of code chunk and should be reported in the Results or Methods section
enviro %>%
  group_by(Station) %>%
  summarise(mean_duration = mean(Effort_hrs),
            standard_deviation = sd(Effort_hrs))

#Also want to see what the distance effort is

class(enviro$Meters) #says it is a character and can't be summed, we need to convert it
enviro$Meters <- as.numeric(enviro$Meters) #now they are numbers

sum(enviro$Meters[enviro$Station == "TBS"]) # 73,864 meters at TBS
sum(enviro$Meters[enviro$Station == "YRS"]) # 76,915 meters at YRS

# per night meters
enviro %>%
  group_by(Station) %>%
  summarise(mean_m_per_night = mean(Meters),
            standard_deviation = sd(Meters))
```

## Small mammals

```{r summarize small mammal numbers}
#Raw mammal counts

sum(small.mamms.temp$Number_of_Small_Mammals_Spotted[small.mamms.temp$Station == "YRS"])

sum(small.mamms.temp$Number_of_Small_Mammals_Spotted[small.mamms.temp$Station == "TBS"])

# I think this is good to report (total numbers) but maybe also with mean and sd of n encountered per survey at each station:

enviro %>%
  group_by(Station) %>%
  summarise(mean_detections_per_hr = mean(Number_of_Small_Mammals_Spotted),
            standard_deviation = sd(Number_of_Small_Mammals_Spotted))

```

# Analysis

## Small Mammals

### Figures

#### Smamm Per hr boxplot

```{r small mammal boxplot}

# make a boxplot!

smammals.plot <- ggplot(data = small.mamms.temp, aes(x = Station, y = smammals.hr, fill = Station)) + geom_boxplot() + theme_bw() + scale_fill_manual(values = c("cornflowerblue", "goldenrod1"), labels = c("Unhunted", "Hunted")) + xlab("Station") +
  theme(legend.title=element_blank()) + ylab("N Small Mammals Detected Per Hour") + ggtitle("Small Mammal Detection Rates by Station") 

smammals.plot #view the plot

# clean
rm(smammals.plot)

# save plot to computer files
#ggsave(smammals.plot, file="small_mammal_plot.png", width = 6, 
      # height = 4)
# can add dimension specifications with width = x and height = x
```

#### Smamm Per hr mean and se

```{r se and mean}
# another way to plot that makes judging significance easier NEEDS EXPLAINING
smams.mean.se <- small.mamms.temp %>%
  group_by(Station) %>%
  summarise(mean = mean(smammals.hr),
            sd = sd(smammals.hr),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

smamm.fig <- ggplot(smams.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = small.mamms.temp, 
              aes(x = Station, y = smammals.hr), 
              width = 0.15, size = 1.5, alpha = 0.5) + coord_fixed(ratio = 0.75) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21, size = 3) + theme_bw() + scale_fill_manual(values = c("cornflowerblue", "goldenrod1")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Small Mammal Detections by Station") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma", margin = margin(t = 8)), axis.text.y = element_text(family = "Tahoma", margin = margin(l = 8)), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, face = "bold", size = 12, margin = margin(b = 12))) 

smamm.fig #view plot, also seems significant NEED TO UNDERSTAND GGPLOT CODE

ggsave(smamm.fig, file="small-mammal-mean-se.png", width = 6,  height = 6)

rm(smams.mean.se, smamm.fig)

```

### Smamm Model

```{r small mammal model}

# note: nbinom2 with offset term and zero inflation term will run and seems okay, but it doesn't pass validation cleanly. Other attempts (with and without random effects) cause model convergence problems. 

# Model as continuous response variable rather than raw count data

smam.cont <- glmmTMB(smammals.hr ~ Station + (1|Parent_Trail_LongName), data = small.mamms.temp, family = tweedie(link = "log"))

sim <- simulateResiduals(fittedModel = smam.cont, plot = T)

summary(smam.cont) # great, can report this

testDispersion(smam.cont, plot = F)
testZeroInflation(smam.cont, plot = F)

# could report in this format:
Anova(smam.cont)

#

#https://stats.stackexchange.com/questions/603155/seeking-feedback-on-the-interpretation-and-reporting-of-a-glmm-output
```



```{r clean env after smamm}
rm(sim, small.mamms.temp, smam.cont)
```

## Understory

### Understory Set-Up

Get understory data reformatted. We took these readings every 15 min. First step will be looking at variation between stations. 

```{r reformat understory}
understory_long <- enviro %>%
  select(Survey_Start_Date_Time, Station, Parent_Trail_LongName, Understory_Density_1:Understory_Density_13, Understory_Density_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Understory_Density_1:Understory_Density_Final,
               values_to = "Understory_leaves", names_repair = "unique")

# remove NAs

understory_long <- understory_long %>%
  na.omit()

```

### Und Figure

Histogram per station

```{r understory histo}

under.hist <- ggplot(understory_long, aes(x= Understory_leaves)) +
  geom_histogram() +facet_wrap(~Station) + theme_bw()

under.hist

# summary
understory_long %>%
  na.omit() %>%
  group_by(Station) %>%
  summarise(mean = mean(Understory_leaves),
            sd = sd(Understory_leaves),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

And do a mean and se with points overlain

```{r understory mean se figure}

und.mean.se <- understory_long %>%
  na.omit() %>%
  group_by(Station) %>%
  summarise(mean = mean(Understory_leaves),
            sd = sd(Understory_leaves),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

und.fig <- ggplot(und.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) + geom_jitter(data = understory_long, aes( x = Station, y = Understory_leaves), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.75) + theme_bw() +
  scale_fill_manual(values = c("lightgreen","lightgreen")) + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5) + geom_point(pch = 21, size = 3) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Understory (vertical leaf overlap)") + ggtitle("Understory density") + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(legend.position = "none") + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

und.fig

#ggsave(und.fig, file="understory-mean-se.png", width = 6,  height = 4)

rm(und.fig)
```

### Und Model

NOTE THIS DOES NOT NEED TO BE REPORTED SO IM COMMENTING IT OUT

```{r understory model}

#und.mod <- glmmTMB(Understory_leaves ~ Station + (1|Parent_Trail_LongName), data = understory_long, family = nbinom2())

#sim <- simulateResiduals(fittedModel = und.mod, plot = T)

#summary(und.mod) # no difference between understory

#testDispersion(und.mod, plot = F)
#testZeroInflation(und.mod, plot = F)

# could report in this format:
#Anova(und.mod)

```

## Temperature

### Temp Set-Up

Get temperature data reformatted. We took these readings every 30 min and also every time an individual was found. First step will be looking at variation between stations. 

```{r reformat temperature}
# First, let's add column to differentiate by season (2024 or 2025)
enviro$Season <- factor(format(enviro$Start_Date, "%Y"))

# pull out per data record enviro data from snake.sur

env_at_detection <- snakes.sur %>%
  select(Station, Start_Date, Temperature_at_Time_of_Detection, Humidity_at_Time_of_Detection) %>% na.omit()

enviro2 <- left_join(enviro, env_at_detection)

temperature_long <- enviro2 %>%
  select(Survey_Start_Date_Time, Station, Season, Parent_Trail_LongName, 
         Temperature_1:Temperature_7, Temperature_Final, Temperature_at_Time_of_Detection) %>%
  pivot_longer(names_to = "Measurement",
               cols = Temperature_1:Temperature_at_Time_of_Detection,
               values_to = "Temperature", names_repair = "unique")
# remove NAs

temperature_long <- temperature_long %>%
  na.omit()
```

### Temp Figure

Histogram per station & season

```{r temperature histo}

temperature.hist <- ggplot(temperature_long, aes(x= Temperature)) +
  geom_histogram() +facet_grid(Season~Station) + theme_bw()

temperature.hist
```

And do a mean and se with points overlain

```{r temperature mean se figure}

temp.mean.se <- temperature_long %>%
  na.omit() %>%
  group_by(Station, Season) %>%
  summarise(mean = mean(Temperature),
            sd = sd(Temperature),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

temp.fig <- ggplot(temp.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) + geom_jitter(data = temperature_long, aes( x = Station, y = Temperature), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.75) + theme_bw() +
  scale_fill_manual(values = c("lightgreen","lightgreen")) + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5) + geom_point(pch = 21, size = 3) + facet_wrap(~Season) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Measured Temperature") + ggtitle("Temps by Field Season & Station") + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(legend.position = "none") + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

temp.fig

#Interesting, seems 2025 was def colder at YRS, and a little bit overall as well

#ggsave(temp.fig, file="temp-mean-se.png", width = 6,  height = 4)

rm(temp.fig)
```

```{r clean after temp}
rm(temp.mean.se, temperature.hist)
```

## Humidity

### Hum Set-Up

Get temperature data reformatted. We took these readings every 30 min and also every time an individual was found. First step will be looking at variation between stations. 

```{r reformat humidity}

humidity_long <- enviro2 %>%
  select(Survey_Start_Date_Time, Station, Season, Parent_Trail_LongName, 
         Humidity_1:Humidity_7, Humidity_Final, Humidity_at_Time_of_Detection) %>%
  pivot_longer(names_to = "Measurement",
               cols = Humidity_1:Humidity_at_Time_of_Detection,
               values_to = "Humidity", names_repair = "unique")
# remove NAs

humidity_long <- humidity_long %>%
  na.omit()

# Sanity check to make sure pivoted temp and humidity DFs are same length
length(humidity_long$Survey_Start_Date_Time)
length(temperature_long$Survey_Start_Date_Time) 
```

### Hum Figure

Histogram per station & season

```{r humidity histo}

humidity.hist <- ggplot(humidity_long, aes(x= Humidity)) +
  geom_histogram() +facet_grid(Season~Station) + theme_bw()
humidity.hist
```

And do a mean and se with points overlain

```{r humidity mean se figure}

hum.mean.se <- humidity_long %>%
  na.omit() %>%
  group_by(Station, Season) %>%
  summarise(mean = mean(Humidity),
            sd = sd(Humidity),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

hum.fig <- ggplot(hum.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) + geom_jitter(data = humidity_long, aes( x = Station, y = Humidity), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.75) + theme_bw() +
  scale_fill_manual(values = c("lightgreen","lightgreen")) + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5) + geom_point(pch = 21, size = 3) + facet_wrap(~Season) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("% Relative Humidity") + ggtitle("Humidity by Field Season & Station") + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(legend.position = "none") + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

hum.fig

#Huge amount of variation in humidity, overall more humid at YRS but 
# definitely more humid in 2025, likely from all the huge rain events 

#ggsave(hum.fig, file="humidty-mean-se.png", width = 6,  height = 4)

#rm(hum.fig)
```


```{r clean env after hum}
rm(hum.fig, hum.mean.se, humidity.hist)
```

## Handling Humidity and Temp

We want to summarize humidity and temp in a way that allows us to incorporate it into our model. As of 2/23, I'm thinking the best way to do this is to take means per survey date.

Let's take the mean vals per date

```{r temp hum means} 

temp_date_means <- temperature_long %>%
  na.omit() %>%
  group_by(Survey_Start_Date_Time, Station) %>%
  reframe(mean_temp_F = mean(Temperature))

hum_date_means <- humidity_long %>%
  na.omit() %>%
  group_by(Survey_Start_Date_Time, Station) %>%
  reframe(mean_humidity_percent = mean(Humidity))

temp_hum_df <- left_join(temp_date_means, hum_date_means)

temp_hum_df$Start_Date <- as.Date(temp_hum_df$Survey_Start_Date_Time)

temp_hum_df <- temp_hum_df %>%
  select(-Survey_Start_Date_Time)

# clean
rm(temperature_long, humidity_long, temp_date_means, hum_date_means, understory_long)
```

## Density estimates

First, create distance bins. Also look at some things.

```{r review data more} 

# pull out year to make a way to distinguish between seasons

snakes.sur$Year <- factor(format(snakes.sur$Start_Date, "%Y"))

snakes.sur %>%
  group_by(Year, Station) %>%
    reframe(N_surveys = length(unique(Start_Date))) #Effort comparisons between stations

enviro %>%
  group_by(Season, Station) %>%
    reframe(Accum_hrs = sum(Effort_hrs))

# Deciding on distance bins in distance sampling involves selecting cutpoints that adequately capture the shape of the detection function, typically ensuring more bins near the line (where detection probability changes rapidly) and fewer further away. The goal is to minimize bias from data rounding while ensuring enough observations per bin for accurate model fitting. 

# Smaller bins closer to zero

# Truncate your data to remove the farthest detections - WE CANNOT DO THIS DUE TO THE BOA BIAS THAT OWEN POINTED OUT

# First, look at a histogram of distance from trail to get an idea of detection distribution

hist(snakes.sur$Distance_From_Trail_m, breaks = seq(0, 12, by = 0.5))

```

Proceed with specifying bins and truncating the data to remove the 11.5m away detection.

```{r divide up bins} 

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 0.25, "0_0.25", NA))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 0.25 & Distance_From_Trail_m <= 0.5, "0.25_0.5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 0.5 & Distance_From_Trail_m <= 1, "0.5_1", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 1.5, "1_1.5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1.5 & Distance_From_Trail_m <= 2, "1.5_2", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 2.5, "2_2.5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2.5 & Distance_From_Trail_m <=3.5, "2.5_3.5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 3.5 & Distance_From_Trail_m <=5, "3.5_5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 5, "5_8", Distance_bin_m))

# join temp and humidity means to the snakes.sur data
snakes.sur <- left_join(snakes.sur, temp_hum_df)
```

```{r divide up bins broader}

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 1, "0_1", NA))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 2, "1_2", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 3, "2_3", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
  #mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 3 & Distance_From_Trail_m <= 4, "3_4", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
#  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 4 & Distance_From_Trail_m <= 5, "4_5", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 5, "5_8", Distance_bin_m))

# join temp and humidity means to the snakes.sur data
#snakes.sur <- left_join(snakes.sur, temp_hum_df)

```

```{r format for unmarked}

# isolate surveys for which no snakes were found
zeros <- snakes.sur %>%
  filter(is.na(Snake_ID)) %>%
  select(Start_Date, Year, Station, Diet_Type, Segment, Seg_dist_m, mean_temp_F, mean_humidity_percent)

# get them into the format we want (tally doesn't actually matter here)
z.tally.no.mammals <- zeros %>%
  group_by(Start_Date, Year, Station, Diet_Type, Segment, Seg_dist_m, mean_temp_F, mean_humidity_percent) %>% tally()

# for the nights when 0 snakes were found, make a dataframe of instances that specifies zero non-mammal eaters were found specifically
z.tally.no.mammals$Diet_Type <- "no_mammals"

# repeat with making a df of instances when zero mammal eating snakes were found
z.tally.yes.mammals <- zeros %>%
  group_by(Start_Date, Year, Station, Diet_Type, Segment, Seg_dist_m, mean_temp_F, mean_humidity_percent) %>% tally()

# make all diet types = no mammals

z.tally.yes.mammals$Diet_Type <- "includes_mammals"

# stack them together

z.tallies <- rbind(z.tally.no.mammals, z.tally.yes.mammals)

#clean
rm(z.tally.yes.mammals, z.tally.no.mammals)

# also, figure out why the zero detections on CHIO-03 6-11-2024 are entered twice, as are 6-23 MAPO03, 6-23 MURC-01, and 6-23 PUEN-01 (all 2024) - ignoring for now bc it's an error I'm pretty sure

# remove the n column (meaningless)
z.tallies <- z.tallies %>%
  select(!n)

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

length(surveys.no.zeros2$Snake_ID) # SK04 and SK05 are duplicated
# remove the duplicate entries
# OWEN to go back and figure out where/why and remove earlier
surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally <- surveys.no.zeros %>%
  group_by(Start_Date, Year, Station, Diet_Type, Segment, Seg_dist_m, Distance_bin_m, mean_temp_F, mean_humidity_percent) %>% tally()

# pivot wider that wider
density_est_df <- survey.tally %>% 
  pivot_wider(names_from = Distance_bin_m,
              values_from = n, 
              values_fill = 0)

# now, we can add the surveys and the segments with zero detections back

density_est_df <- full_join(z.tallies, density_est_df)

# make NAs zeros - except the temp and humidty NAs
density_est_df[9:17][is.na(density_est_df[9:17])] <- 0

# clean
rm(survey.tally, surveys.no.zeros, z.tallies, zeros)
```

Note I think it's a good idea to follow the methods in this paper: Hierarchical distance sampling to estimate population sizes of common lizards across a desert ecoregion (in Zotero shared library)


```{r density estimation} 

# Package things up into an unmarkedFrame 

# Seasons
# Feedback from Morgan indicates that combining years is not a problem in our case. We may want to explore incorporating year as a factor though. 

density_est_df$Year <- as.factor(density_est_df$Year)
density_est_df$Diet_Type <- as.factor(density_est_df$Diet_Type)
density_est_df$Seg_dist_m <- as.numeric(density_est_df$Seg_dist_m)
density_est_df$Segment <- as.factor(density_est_df$Segment)
density_est_df$Station <- as.factor(density_est_df$Station)


# separate out the covariates
covs <- density_est_df[,c("Station", "Year", "Diet_Type", "Segment", "Seg_dist_m")]

umf <- unmarkedFrameDS(y=as.matrix(density_est_df[,9:17]), siteCovs=data.frame(covs), dist.breaks=c(0, 0.25, 0.5, 1, 1.5, 2, 2.5, 3.5, 5, 8), unitsIn="m", survey="line", tlength=covs$Seg_dist_m)

hist(umf, xlab="distance (m)", main="", cex.lab=0.8, cex.axis=0.8)

```

### Mods without Env

First compare models based on AIC without worrying about temperature and humidity.

```{r density models no env covs}

# Fit model 1, null model
# since fixing the mistake that lost some observations, we now get an error regareing vmmin not being finite. Royle seems a little mystified by this too in a similar example https://groups.google.com/g/unmarked/c/XreznyG1xo4

null <- distsamp(~1 ~ 1, umf, unitsOut = "kmsq")

# Mod 2
Year <- distsamp(~ 1 ~ Year, umf, unitsOut = "kmsq")

# Mod 3

Year_Station <- distsamp(~1 ~ Year + Station, umf, unitsOut = "kmsq")

# Mod 4
Year_Station_StationXGuild <- distsamp(~1 ~ Year + Station/Diet_Type, umf, unitsOut = "kmsq")

# Mod 5
Station <- distsamp(~1 ~ Station, umf, unitsOut = "kmsq")

# Mod 7
Station_StationXGuild <-distsamp(~1 ~ Station/Diet_Type, umf, unitsOut = "kmsq")

# Make a table to compare models

# put mods in a list
mod.set.1 <- list(null = null, Year = Year, Year_Station = Year_Station, Year_Station_StationXGuild = Year_Station_StationXGuild, Station_StationXGuild = Station_StationXGuild, Station = Station)

# make table
aic_table.1 <- aictab(cand.set = mod.set.1, sort = TRUE)

aic_table.1
```

Lowest AIC model has fixed terms of year, station, and the interaction between station and diet type

```{r best model without env vars}

summary(Year_Station_StationXGuild)
summary(Station_StationXGuild)

# Get coefficients for covariate effects (on log scale)

coef(Year_Station_StationXGuild)
```

```{r goodness of fit without env covs}

# using this resource: https://www.montana.edu/screel/teaching/bioe-440r-521/bioe-440-distance-sampling.html

# Goodness of fit test: good fit if results of these tests show p > 0.05. First, fit a model: use your best-supported model (Lowest AIC). Then check GOF for that model.

Year_Station_StationXGuild

# Function returning three fit-statistics.

fitstats <- function(fm) {

  observed <- getY(fm@data)

  expected <- fitted(fm)

  resids <- residuals(fm)

  sse <- sum(resids^2)

  chisq <- sum((observed - expected)^2 / expected)

  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2)

  out <- c(SSE=sse, Chisq=chisq, freemanTukey=freeTuke)

  return(out)

}



pb <- parboot(Year_Station_StationXGuild, fitstats, nsim=25, report=1)

pb

# from my not thorough understanding of this, the tests indicate or top model passes goodness of fit tests. 
```


```{r extract predicted values without env vars}

new.data1 <- density_est_df %>%
  ungroup() %>%
  dplyr::select(Station, Year, Diet_Type) %>%
  distinct()
  

#Predict

predictions1 <- predict(Year_Station_StationXGuild, type="state", newdata=new.data1, appendData=TRUE)

predictions1  

```

```{r plot without env covs, warning=FALSE}

p1<-ggplot(predictions1, aes(x=Station, y=Predicted, fill = Year, group = Station)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = .2) + geom_point(pch = 21, size = 3) + theme_bw() + scale_fill_manual(values = c("cornflowerblue", "goldenrod1")) + xlab("") + scale_color_discrete(labels = "") +  scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates by year and diet type") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12)) + facet_wrap(~Diet_Type)
p1
```

Repeat without year because of the added difficulty that brings to interpretation AND it's not a hypothesis we're testing. We don't expect snakes to vary between years and in the second year, effort was higher, which I believe is reflected by the smaller error bars for 2025.

```{r assess mod fit without year}
# this model only has station and the station x diet type interaction. 

pb2 <- parboot(Station_StationXGuild, fitstats, nsim=25, report=1)

pb2 # is okay based on these tests

```

```{r extract predictions without year}
new.data2 <- density_est_df %>%
  ungroup() %>%
  dplyr::select(Station, Diet_Type) %>%
  distinct()
  
#Predict

predictions2 <- predict(Station_StationXGuild, type="state", newdata=new.data2, appendData=TRUE)

predictions2

```
Plot predictions from model without Year included

```{r plot without year}

p2<-ggplot(predictions2, aes(x=Station, y=Predicted, fill = Diet_Type, group = Diet_Type)) + geom_line(linetype = "dashed", position = position_dodge(width = 0.5))+
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = 0.25, position = position_dodge(width = 0.5)) + geom_point(pch = 21, size = 3, position = position_dodge(width = 0.5)) + theme_bw() + scale_fill_manual(values = c("blue4", "goldenrod3"), labels = c("Mammal eater", "Non-mammal eater")) + xlab("")  + theme(legend.title = element_text("Diet type")) + scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates by diet type") + labs(fill = "Diet type") + 
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12))

p2

# without connector line

p3<-ggplot(predictions2, aes(x=Station, y=Predicted, fill = Diet_Type, group = Diet_Type)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = 0.25, position = position_dodge(width = 0.5)) + geom_point(pch = 21, size = 3, position = position_dodge(width = 0.5)) + theme_bw() + scale_fill_manual(values = c("blue4", "goldenrod3"), labels = c("Mammal eater", "Non-mammal eater")) + xlab("")  + theme(legend.title = element_text("Diet type")) + scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates by diet type") + labs(fill = "Diet type") + 
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12))

p3
```

Clean plots and the other models from above and some other misc things

```{r clean plots and mods part one}
rm(new.data1, new.data2, null, p1, p2, p3, pb, pb2, Station, umf, covs, env_at_detection, enviro2, temp_hum_df, Year, Year_Station, Year_Station_StationXGuild, und.mean.se, under.hist, segs_expanded, segs, mod.set.1)

```

### Mods with env covs

Let's look at how humidity and temperature affect model AIC values. 

```{r density est mods hum and temp}

# HUMIDITY AND TEMPERATURE INCLUSION

# The AIC of the models below can't be compared with mods 1-4 above bc of the NA removal necessary to include the humidity and the temperature data

# make a df that excludes all surveys associated with the NA readings of temp and humidity

density_est_df.shortened <- density_est_df %>%
  na.omit()

# scale and center continuous covs
density_est_df.shortened$temp_scaled <-as.numeric(scale(density_est_df.shortened$mean_temp_F, center = TRUE))

density_est_df.shortened$hum_scaled <-as.numeric(scale(density_est_df.shortened$mean_humidity_percent, center = TRUE))

covs.shortened <- density_est_df.shortened[,c("Station", "Diet_Type", "Year", "Segment", "Seg_dist_m", "temp_scaled", "hum_scaled")]

umf.shortened <- unmarkedFrameDS(y=as.matrix(density_est_df.shortened[,9:17]), siteCovs=data.frame(covs.shortened), dist.breaks=c(0, 0.25, 0.5, 1, 1.5, 2, 2.5, 3.5, 5, 8), unitsIn="m", survey="line", tlength=covs.shortened$Seg_dist_m)
```

```{r now structure and compare models}

# set levels here

# Fit model 1, null model

null2 <- distsamp(~1 ~ 1, umf.shortened, unitsOut = "kmsq")

# Mod 2
Year2 <- distsamp(~ 1 ~ Year, umf.shortened, unitsOut = "kmsq")

# Mod 3
Year_Station2 <- distsamp(~1 ~ Year + Station, umf.shortened, unitsOut = "kmsq")

# Mod 4
Year_Station_StationXGuild2 <- distsamp(~1 ~ Year + Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Mod 5

Station2 <- distsamp(~1 ~ Station, umf.shortened, unitsOut = "kmsq")

# Mod 7
Station_StationXGuild2 <-distsamp(~1 ~ Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Mod 8
# Substitute temperature for year
Temp_Station_StationXGuild <- distsamp(~1 ~ temp_scaled + Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Mod 9
# Substitute humidity for year
Humid_Station_StationXGuild <- distsamp(~1 ~ hum_scaled + Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Mod 10
# Temperature and year
Temp_Yr_Stn_StnXGuild <- distsamp(~1 ~ Year + temp_scaled + Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Mod 11
# Humidity and year
Humid_Yr_Stn_StnXGuild <- distsamp(~1 ~ Year + hum_scaled + Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Mod 12
# Humidity, Temp, Year (and Station and Station x Diet type)
Humid_Yr_Temp_Stn_StnXGuild <- distsamp(~1 ~ Year + hum_scaled + temp_scaled + Station/Diet_Type, umf.shortened, unitsOut = "kmsq")

# Make a table to compare models

# put mods in a list
mod.set.2 <- list(null = null2, Year = Year2, Year_Station = Year_Station2, Year_Station_StationXGuild = Year_Station_StationXGuild2, Station_StationXGuild = Station_StationXGuild2, Station = Station2, Temp_Station_StationXGuild = Temp_Station_StationXGuild, Humid_Station_StationXGuild = Humid_Station_StationXGuild, Temp_Yr_Stn_StnXGuild = Temp_Yr_Stn_StnXGuild, Humid_Yr_Temp_Stn_StnXGuild = Humid_Yr_Temp_Stn_StnXGuild)

# make table
aic_table.2 <- aictab(cand.set = mod.set.2, sort = TRUE)

aic_table.2

summary(Temp_Yr_Stn_StnXGuild)
```

Can explore changing baseline in model.


So.... we can look at *Temp_Yr_Stn_StnXGuild* (top mod based on AIC), which has a lower AIC than the model that only includes Year, Station, and the Station x Guild interaction (dAIC = 2.95 between those two models). That difference in AIC values indicates that Temperature could be an important covariate to consider. However, when year is not used (which I'm like 99% sure is what we want for testing our specific hypotheses which have nothing to do with differences between years, between which sampling effort differed), then **temperature does not improve model fit** and actually makes it a tad worse: *Station_StationXGuild* AIC = 700.86, *Temp_Station_StationXGuild* = 701.39 (dAIC = 0.53) (all of which perform better than the null by a lot). 

Assuming the three models above pass GOF tests, we probably want to report all 3 of them in the supplementary materials, along with the top model from the non-shortened dataset that didn't consider env covs (the section above this one) which includes year. This would be to show that we checked and confirmed results were relatively consistent, but we I think it probably makes sense to stick with reporting the results from the Station_StationXGuild model from the first set of tests in the main paper (so the mod that incorporates all data collected, rather than the one that has shortened data to accommodate for the NAs for the missing temp and humidity data on some surveys).


Okay moving on, look at GOF (Goodness of fit)

```{r GOF env covs and shortened df mods}

# GOF for top model from above (Temp_Yr_Stn_StnXGuild)
Temp_Yr_Stn_StnXGuild_GOF <- parboot(Temp_Yr_Stn_StnXGuild, fitstats, nsim=25, report=1)

Temp_Yr_Stn_StnXGuild_GOF
# again, based on my not great understanding of this, it passes

# GOF for Station_StationXGuild
Station_StationXGuild_GOF <- parboot(Station_StationXGuild, fitstats, nsim=25, report=1)

Station_StationXGuild_GOF # passes. I think something is up with the ChiSq part of this though for all the GOF tests I've run with this function tbh

# clean
rm(Station_StationXGuild_GOF, Temp_Yr_Stn_StnXGuild_GOF)

# one last GOF test using the temp and station and stationxguild model without year (which doesn't perform as well as the model that leaves both temp and year out)

bp<-Temp_Station_StationXGuild_GOF <- parboot(Temp_Station_StationXGuild, fitstats, nsim=25, report=1)

bp

```

Model predictions - first looking at *Temp_Yr_Stn_StnXGuild*

```{r mod predictions temp}
# new data frame for prediction extraction, first, look at trend with temperature because I'm curious
new.data.temp.continuous <- density_est_df.shortened %>%
  ungroup() %>%
  dplyr::select(Station, Diet_Type, Year, temp_scaled) %>%
  distinct()


#Predict

predictions_temp <- predict(Temp_Yr_Stn_StnXGuild, type="state", newdata=new.data.temp.continuous, appendData=TRUE)

predictions_temp 
```

Plot trend with temp based on model with year (which has 2024 predictions that are wack)

```{r temp trend model with year}

# trend exists in both guilds, fewer detections when hotter (not what I'd expect but I guess it's always kinda hot there)

p1 <-ggplot(predictions_temp, aes(x=temp_scaled, y=Predicted, fill = Diet_Type))  + geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE)) +
geom_point(pch = 21, size = 2) + theme_bw() + scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates and temperature") + scale_fill_manual(values = c("darkred", "lightgreen"), labels = c("Mammal eater", "Non-mammal eater")) + xlab("")  + theme(legend.title = element_text("Diet type")) 

p1

# omit 2024 data

p2 <- predictions_temp %>%
  filter(Year == "2025") %>%
  ggplot(aes(x=temp_scaled, y=Predicted, fill = Diet_Type))  + geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE)) +
geom_point(pch = 21, size = 2) + theme_bw() + scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates and temperature") + scale_fill_manual(values = c("darkred", "lightgreen"), labels = c("Mammal eater", "Non-mammal eater")) + xlab("")  + theme(legend.title = element_text("Diet type")) 

p2
```

Okay cool well I'm satisfied. It appears that temperature has a slight negative effect on detections. Now I guess we need to check to make sure our results are still the same when temperature is included in the model.


```{r mod predictions yr station stationxguild}

# new data frame, hold temp constant at mean
temp_mean <- mean(density_est_df.shortened$temp_scaled) # get mean temp (scaled)

new.data.temp.constant <-density_est_df.shortened %>% # isolate covariates in new df
  ungroup() %>%
  dplyr::select(Station, Diet_Type, Year) %>%
  distinct()

new.data.temp.constant$temp_scaled <- temp_mean # insert mean for temp (needs to have same name as cov when entered in umf, hence 'temp_scaled')

#Predict with temp held constant at mean

predictions_temp.constant <- predict(Temp_Yr_Stn_StnXGuild, type="state", newdata=new.data.temp.constant, appendData=TRUE)

predictions_temp.constant 
```

Plot

```{r plot with temp held constant}

ggplot(predictions_temp.constant, aes(x=Station, y=Predicted, fill = Diet_Type, group = Diet_Type)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = 0.25, position = position_dodge(width = 0.5)) + geom_point(pch = 21, size = 3, position = position_dodge(width = 0.5)) + theme_bw() + scale_fill_manual(values = c("blue4", "goldenrod3"), labels = c("Mammal eater", "Non-mammal eater")) + xlab("")  + theme(legend.title = element_text("Diet type")) + scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates by diet type\n using model that accounts for temperature") + labs(fill = "Diet type") + 
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12)) + geom_line(linetype = "dashed", position = position_dodge(width = 0.5)) + facet_wrap(~Year, scales = "free")

```


Okay great things are looking consistent. Now lets do this with the model that accounts for temperature but not for year (which... technically does not perform better based on AIC when compared to the model that omits temperature and year).

What to report?
Morgan says betas and 95% CIs, or SE (for comparison) - everything log likelihood.



```{r mod predictions no year yes temp}

# new data frame, hold temp constant at mean
temp_mean <- mean(density_est_df.shortened$temp_scaled) # get mean temp (scaled)

new.data.temp.constant.no.yr <-density_est_df.shortened %>% # isolate covariates in new df
  ungroup() %>%
  dplyr::select(Station, Diet_Type) %>%
  distinct()

new.data.temp.constant.no.yr$temp_scaled <- temp_mean # insert mean for temp (needs to have same name as cov when entered in umf, hence 'temp_scaled')

#Predict with temp held constant at mean

predictions_temp.constant.no.yr <- predict(Temp_Station_StationXGuild, type="state", newdata=new.data.temp.constant.no.yr, appendData=TRUE)

```

And plot it

```{r plot no year yes temp}

ggplot(predictions_temp.constant.no.yr, aes(x=Station, y=Predicted, fill = Diet_Type, group = Diet_Type)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = 0.25, position = position_dodge(width = 0.5)) + geom_point(pch = 21, size = 3, position = position_dodge(width = 0.5)) + theme_bw() + scale_fill_manual(values = c("blue4", "goldenrod3"), labels = c("Mammal eater", "Non-mammal eater")) + xlab("")  + theme(legend.title = element_text("Diet type")) + scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake density estimates by diet type\n using model that accounts for temperature") + labs(fill = "Diet type") + 
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12)) + geom_line(linetype = "dashed", position = position_dodge(width = 0.5)) 

```

Okay I feel pretty confident we can say even when year and temperature are controlled for, we find the same trends with our data. 

```{r clean some}

rm(null2, p1, p2, predictions_temp, predictions_temp.constant, predictions_temp.constant.no.yr, predictions1, predictions2, new.data.temp.constant, new.data.temp.continuous, new.data.temp.constant.no.yr, Station_StationXGuild, Station_StationXGuild2, Station2, Temp_Station_StationXGuild, Temp_Yr_Stn_StnXGuild, umf.shortened, Year2, mod.set.2, Year_Station2, Year_Station_StationXGuild2, Temp_Station_StationXGuild_GOF, Humid_Station_StationXGuild, Humid_Yr_Stn_StnXGuild, Humid_Yr_Temp_Stn_StnXGuild, density_est_df.shortened, aic_table.1, aic_table.2, bp, density_est_df, covs.shortened, temp_mean)

```

# Community composition

I have tried NMDS about a hundred different ways (aggregated by 7 days, 10 days, 14 days, 20, a month, a year; species level, genus level, sub family, family level; all data from all detection methods and only the survey detections). Nothing works (continually get a warning stress is basically 0, you-dont-have enough-data message).

Below I mess around with some basic community composition visualizations.

```{r species occurrences}

# just get an idea of n detections from all survey types per station, including the instances with no ID
raw_count <- snakes.master %>%
  group_by(Station) %>%
  reframe(n = n())

# total at YRS is 111, at TBS is 108 (great, super comparable numbers when pooled)

# subfamily level may help
# create this column and populate it:

snakes.master <- snakes.master %>%
  mutate(Subfamily = ifelse(Genus == "Siphlophis" | Genus == "Imantodes" | Genus == "Dipsas" | Genus == "Drepanoides" | Genus == "Oxyrhopus" | Genus == "Leptodeira" | Genus == "Hydrops" | Genus == "Helicops" | Genus == "Atractus" | Genus == "Erythrolamprus" | Genus == "Taeniophallus" | Genus == "Pseudoboa", "Dipsadinae", 
                            ifelse(Genus == "Epicrates" | Genus == "Corallus" | Genus == "Eunectes", "Boinae",
                                   ifelse(Genus == "Chironius" | Genus == "Drymobius" | Genus == "Tantilla" | Genus == "Drymarchon" | Genus == "Oxybelis" | Genus == "Spilotes" | Genus == "Dendrophidion" | Genus =="Leptophis", "Colubrinae",
                                          ifelse(Genus == "Bothrops" | Genus == "Bothrocophias", "Crotalinae",
                                                 ifelse(Genus == "Xenoxybelis", "Xenodontinae",
                                                        ifelse(Genus == "Micrurus", "Elapinae", Subfamily)))))))

#Siphlophis = Dipsadinae
#Imantodes = Dipsadinae 
#Epicrates = Boinae
#Dipsas = Dipsadinae
#Drepanoides = Dipsadinae  
#Chironius = Colubrinae
#Corallus = Boinae
#Oxyrhopus = Dipsadinae
#Leptodeira = Dipsadinae
#Hydrops = Dipsadinae
#Bothrops = Crotalinae
#Xenoxybelis = Xenodontinae
#Drymobius = Colubrinae
#Micrurus = Elapinae
#Helicops = Dipsadinae
#Eunectes = Boinae   
#Atractus = Dipsadinae
#Tantilla = Colubrinae
#Drymarchon = Colubrinae
#Erythrolamprus = Dipsadinae
#Taeniophallus = Dipsadinae
#Bothrocophias = Crotalinae
#Spilotes = Colubrinae
#Pseudoboa = Dipsadinae
#Oxybelis = Colubrinae
#Dendrophidion = Colubrinae
#Leptophis = Colubrinae
#Bothrocophias = Crotalinae

# make an edge list of n times an individual was seen at each station for each subfamily group
edgelist <- snakes.master %>%
  group_by(Station, Subfamily) %>%
  tally() %>%
  na.omit() 

# piechart
ggplot(edgelist, aes(x="", y= n, fill=Subfamily)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) + # remove background, grid, numeric labels
  scale_fill_brewer(palette="Set1") # remove background, grid, numeric labels

# only the survey snake data, at subfamily level
surveys <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Subfamily) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys, aes(x="", y= n, fill=Subfamily)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) + 
  scale_fill_brewer(palette="Set1") 

# Genus level, all detection methods:
all.genus <- snakes.master %>%
  group_by(Station, Genus) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(all.genus, aes(x="", y= n, fill=Genus)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station)

# Genus level, surveys only:
surveys.genus <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Genus) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys.genus, aes(x="", y= n, fill=Genus)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) 
```

# Diversity

```{r diversity & hill numbers}

# Diversity

# need data in a new species x site matrix
unique(snakes.sur$Species)

temp <- snakes.sur %>%
  group_by(Station, Species) %>% tally()

temp <- temp %>%
  na.omit()

# Need to modify the matrix so we have an n column, etc.


# get the new format, species as row names, station as columns
div.mat <- temp %>% 
  pivot_wider(names_from = Station, 
              values_from = n, 
              values_fill = 0) %>%
  column_to_rownames(var = "Species")

# now we use div.mat to get the diversity and 95% CIs
# new function uses iNEXT package (confusingly also the name of the function)
species.diversity = iNEXT(div.mat, q = c(0,1,2), datatype = "abundance", nboot = 200)

# extract values we need
species.diversity.output <- species.diversity[["iNextEst"]][["coverage_based"]]

div.vals <- species.diversity.output %>% 
  filter(Assemblage == "TBS" & m == 68 | Assemblage == "YRS" & m == 68)

# plot diversity
diversity.plot <- ggplot(data = div.vals, aes(x = Assemblage, y = qD, fill = as.factor(Order.q), shape = as.factor(Order.q))) + theme_light() + geom_pointrange(aes(ymin = qD.LCL, ymax = qD.UCL), position = position_dodge2(width = 0.35), color = "black", size = 0.6) + xlab(NULL) + ylab("Effective Number of Species") + scale_x_discrete(labels = c("Unhunted", "Hunted")) + scale_fill_manual(values = c("red", "goldenrod", "cornflowerblue"), labels = c("q = 0\n(richness)", "q = 1\n(exp. Shannon)", "q = 2\n(inv. Simpson)")) + coord_fixed(ratio = 0.1) + scale_shape_manual(values = c(21, 22, 23)) +
  theme(axis.title.y = element_text(size = 11, colour = "black", family = "Tahoma", 
                                    margin = margin(r = 10))) +
  theme(axis.title.x = element_text(size = 11, colour = "black", family = "Tahoma")) +
  theme(axis.text = element_text(size = 11, colour = "black", family = "Tahoma")) +
  theme(strip.text.x = element_text(size = 11, colour = "black", family = "Tahoma")) + 
  theme(legend.title = element_blank()) + theme(legend.text = element_text(size = 10)) +
  theme(plot.title = element_text(size = 11, colour = "black", family = "Tahoma", 
                                  hjust = .5, face = "bold")) + theme(legend.position = "bottom") +
  guides(fill = guide_legend(override.aes = list(shape = c(21, 22, 23)))) + guides(shape = "none") + ggtitle(str_wrap("Station-Level Diversity Measured with Hill Numbers (q = 0, 1, 2)", width=40))

diversity.plot

#ggsave("snk_diversity_plot.png", plot = diversity.plot, width = 6, height = 6.5)

```

Species accumulation curves

```{r accum curves}

ggiNEXT(species.diversity, type=1, facet.var="Order.q", grey=TRUE) 

```

Right so no difference there.