---
title: "Snake_Analysis"
author: "Owen McCool, Therese Lamperty"
date: "2025-07-16"
output: word_document
---

# Code Setup & package read in 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read in packages}
library(vegan)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(iNEXT) # species diversity
library(hillR) # species diversity
library(DHARMa) # checking models
library(car) # reporting model stats
library(glmmTMB)
library(unmarked) #density estimates with hierarchical distance sampling models
library(AICcmodavg) # has convenient functions for comparing models based on AIC
library(showtext)
library(ggsignif)
showtext_auto()
font_add(family = "Tahoma", regular = "tahoma.ttf")

```

# Project Overview

Study Site: Research was split between two different zones within the greater Yasuni Biosphere Reserve. Tiputini Biodiversity Station (TBS) served as the faunally intact site, whereas Yasuni Research Station (YRS) served as the defaunated site. Both sites are situated along the Tiputini River within the province of Orellana, Ecuador. TBS, approximately 45 km east of YRS, is only accessible by boat. YRS is along the Maxus Oil Road, and is less integrated into the Amazonian ecosystem. Additionally, unsustainable hunting activities by the indigenous communities have been prominent in the surrounding region at YRS for multiple decades, as noted from 2014. This is thought to be the root cause of defaunation in this zone. 

	I transferred between the two research stations every two weeks, starting my first rotation at TBS on June 10. Time at each station was spent working on 3 separate projects: my herpetofauna surveys, frugivory surveys, and an overarching seed enemy and dispersal project. 

	The rainy season in Northeastern Ecuador is at its peak between March and July. I conducted my study between the months of June and August, 2024. The average high temperature for June in this region is 85.66° F, with the average low of 70.05° F. July has an average high temperature of 86.29° F and an average low of 69.13° F. August has an average high temperature of 91.35° F and an average low of 69.60° F. June receives 143mm of precipitation on average each year, whereas July receives 99.69mm and August receives 77.55mm. (https://weatherandclimate.com/ecuador/orellana/boca-tiputini) 
Put project description here (field sites, time frame of collection,sampling scheme, type of data (eg. species ID,temp, etc.) )

Surveys: Herpetofauna were sampled using visual surveys on foot by two surveyors/night. Trail systems at each site served as the sampling paths. The surveys were repeated around 6 times on each rotation while in the field, and resampled on the second rotations. The surveys typically began around 20:00-20:30 and lasted until 23:00-00:00 CST. At each site, habitat matches were compared and assessed and trails through that habitat were consistently walked. Two different types of habitat, Terra Firma (uplands) and Varzea (flooded forest) were surveyed on the herp nocturnal walks. In addition to this, two significant bodies of water were also surveyed at each site, primarily consisting of oxbow lakes with one year-round large pond. 

	Surveyors used FENIX HM60R headlamps while conducting night surveys. They carefully examined all surrounding habitat, including understory trees, shrubs, vines, water bodies, forest floor, and middle-canopy. Amphibians of the order Anura gave off reflective eyeshine and were easier to sample for, whereas members of Squamata did not. More information about this will be within the discussion section. Effort hours will be calculated based on survey time length and number of surveyors. 

	Additionally, environmental information was taken throughout the course of each survey. Temperature and humidity data was collected with a Mengshen meter in 30 minute intervals on surveys. Understory vegetation density was also measured every 15 minutes along each trail. A 1.5m PVC pipe was randomly placed upright just off the right side of the trail at each measuring point, and the number of leaves making contact with the pipe was counted. 

	Weather and wind were noted before each survey in four nominal categories, with weather either being “clear, partly cloudy, cloudy, rainy” and wind being “no wind, felt on face, branches move, trees move”. I also accounted for precipitation that had already occurred in 6-hour blocks as “time window of last rain event”. 

	Once an animal was found, the first step of the protocol was to note what time of night the find occurred. After this, the temperature and humidity was once again collected. A rangefinder was used to estimate the perpendicular distance of each animal from the trail. A GPS was used to mark and log the coordinates of each animal found. Animals were given a two-letter code based on the Order and Suborder followed by a number, indicating its position of being found. For example, the code of the 9th snake found of the season would be SK09. Species found were also identified to the lowest taxonomic level in the field, and photographs of individuals were taken both to get positive IDs later in the lab when not possible in the field, but also to avoid counting re-captures.

	Snakes found during non-survey times were also recorded. These individuals were split into one of two categories of detection. The first is an “Opportunistic Encounter” (OE), which is any snake that was found by me while out working on other non-herpetofauna projects in the field. The other type of detection is the Third Party Encounter (TPE). This was any snake that was found by other research groups during my rotations at each station, identified to lowest possible taxonomy. The same data was taken for these individuals. GPS coordinates, time of detection, trail, observer, estimation of distance from trail, and habitat type. Because the third-party observers often didn’t have access to the climate data logger, this information was retrieved later on from each station’s respective weather tower data, based on the time of detection. 

Snakes were given a classification as to whether they were small mammal consumers or not. Information on this was retrieved from literature such as online publications, the book Reptiles and Amphibians of the Amazon by R.D. and Patricia Bartlett, and the expert field book Reptiles of Ecuador. 

# Read in and prep data

## Read in

This data was cleaned some in initial steps in 2024-2026 (see archived code). Note that temp and humidity data was removed for 4 surveys at YRS that took place after July 11, 2025, during which the team lost confidence in the data collection instrument. 

```{r data read in}
all_surv_records <- read.csv("Data/all_survey_records_aug2026.csv")
# a record of every segment surveyed, afternoon and night, with length and time

amph_lz <- read.csv("Data/amph_lz_cleaned_aug2026.csv")
# cleaned amphibian and lizard data - a record of every individual found, when, and what segment
# make sure there are no duplicates...
amph_lz <- amph_lz %>%
  distinct(Animal_ID, .keep_all = TRUE) # evidently there was 1

# I cleaned frogs and added a tadpole phase column, now they are a sep df
frogs_df <- read.csv("frogs.csv")
frogs_df$Start_Date <- mdy(frogs_df$Start_Date)

effort <- read.csv("Data/effort_aug2026.csv")
# Date, effort hours, segment length

env_2025 <- read.csv("Data/env_2025_cleaned.csv", encoding = "latin1")
# 2025 env data

env_24_25_nights <- read.csv("Data/env_24_25_nights_aug2026.csv")
# night time survey environmental data both years

small.mamms.df <- read.csv("Data/small_mammals_aug2026.csv")
# records of small mammals spotted during surveys

snakes.master <- read.csv("Data/snakes_master_cleaned_aug2026.csv")
# all snake records

snakes.sur <- read.csv("Data/snake_surveys_cleaned_aug2026.csv")
# all the snake survey records, cleaned
# this version of snakes.sur has those two duplicates (SK04, 05) removed

```


# Sampling Effort Summaries

```{r mean efforts}

# mean and sd search effort per station per survey type
all_surv_records %>%
  group_by(Station, Survey_Type) %>%
  summarize(mean_search_time_hrs = mean(Effort_hrs),
            sd = sd(Effort_hrs))

# mean and sd meters searched per station per survey type
temp <-all_surv_records %>%
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(total_meters_surveyed = sum(Seg_dist_m))

temp %>%
  group_by(Station, Survey_Type) %>%
  summarize(mean_length = mean(total_meters_surveyed),
            sd_length = sd(total_meters_surveyed))

rm(temp)

all_surv_records %>%
  group_by(Station, Survey_Type) %>%
  summarize(cumulative_m = sum(Seg_dist_m),
            cumulative_surveys = n_distinct(Start_Date_Start_Time))

```

# Small Mammals

## small mammal model 

```{r small mammal model}

smam.mod1 <- glmmTMB(smamm.500m ~ Station + (1|Parent_Trail_LongName), data = small.mamms.df, family = tweedie(link = "log"))

sim <- simulateResiduals(fittedModel = smam.mod1, plot = T)

summary(smam.mod1) # great, can report this

testDispersion(smam.mod1, plot = F)
testZeroInflation(smam.mod1, plot = F)

Anova(smam.mod1)
rm(sim, smam.mod1)
```

## small mammal summary data

```{r small mammal summs}

# make summary of mean and SE and keep for plotting later
smams.distance.mean.se <- small.mamms.df %>%
  group_by(Station) %>%
  summarise(mean = mean(smamm.500m),
            sd = sd(smamm.500m),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# also look at raw mammal counts
sum(small.mamms.df$Number_of_Small_Mammals_Spotted[small.mamms.df$Station == "YRS"])

sum(small.mamms.df$Number_of_Small_Mammals_Spotted[small.mamms.df$Station == "TBS"])

# report mean and sd of n encountered per survey at each station:

env_24_25_nights %>%
  group_by(Station) %>%
  summarise(mean_detections_per_survey = mean(Number_of_Small_Mammals_Spotted),
            standard_deviation = sd(Number_of_Small_Mammals_Spotted))

```

# Prep frog and lizard data

Outcome of below is 1 df to use for lizards and 1 to use for frogs - lizards are tallied just based on gross ID and frogs are tallied based on Family

```{r frog lz raw detection rates}

# the amph lz data frame is being used for the lizards (and should be eventually renamed to clear up potential confusion)
# the frogs are done separately bc T added extra columns to that df

# add segment distance data
segments <- all_surv_records %>%
  select(Segment, Seg_dist_m) %>%
  distinct()

# add to amph lizard df
amph_lz <- left_join(amph_lz, segments)

# working with frogs separately in my augmented df (has litter dependency columns)

frogs_df<- left_join(frogs_df, segments)

# columns are crazy, reduce and simplify

amph_lz <- amph_lz %>%
  select(Station, Start_Date, End_Date, Start_Date_Start_Time, End_Date_End_Time, Survey_Type, Gross_ID, Common_Name, Family, Genus, Species, Segment, Seg_dist_m, Parent_Trail_LongName, Height_of_Individual, Habitat_Type)

# tallies
amph_lz_counts <- amph_lz %>%
  group_by(Station, Start_Date, Survey_Type, Parent_Trail_LongName, Segment, Seg_dist_m, Gross_ID) %>% tally()

# repeat with lizard data, family level
lz_counts <- amph_lz %>%
  filter(Gross_ID == "Lizard") %>%
  group_by(Station, Start_Date, Survey_Type, Parent_Trail_LongName, Segment, Seg_dist_m, Family) %>% tally()

# repeat with the frog data, family level
frog_counts <- frogs_df %>%
  group_by(Station, Start_Date, Survey_Type, Segment, Seg_dist_m, Family) %>% tally()

zeros<-anti_join(all_surv_records, amph_lz_counts, by = "Segment") # I think these records where no specimen was found on a segment should be added in with zeros for each gross ID tally

zeros_frogs<-anti_join(all_surv_records, frog_counts, by = "Segment") 

zeros_liz <- anti_join(all_surv_records, lz_counts, by = "Segment")

amph_lz_wide <- amph_lz_counts %>%
  pivot_wider(
    names_from = c(Gross_ID), 
    values_from = n, 
    values_fill = 0)

frogs_wide <- frog_counts %>%
  pivot_wider(
    names_from = Family, 
    values_from = n, 
    values_fill = 0)

lz_wide <- lz_counts %>%
  pivot_wider(
    names_from = Family, 
    values_from = n, 
    values_fill = 0)

# prep to add in the zeros, pull out columns I want
zeros <- zeros %>%
  select(Station, Start_Date, Survey_Type, Segment, Seg_dist_m, Parent_Trail_LongName) 

zeros_frogs <- zeros_frogs %>%
  select(Station, Start_Date, Survey_Type, Segment, Seg_dist_m) 

#bind it with the amph_lz_wide df so that it has the full record of all survey results, even zeros
amph_lz_wide <- rbind(amph_lz_wide, zeros)
lz_wide <- rbind(lz_wide, zeros)
zeros_frogs$Start_Date <- ymd(zeros_frogs$Start_Date)

frogs_wide <- rbind(frogs_wide, zeros_frogs)

# make NA cases (from when I added in the zeros) to zero
amph_lz_wide<- amph_lz_wide %>% mutate(across(c(Lizard, Frog, Salamander), ~replace_na(.x, 0)))

frogs_wide<- frogs_wide %>% mutate(across(where(is.numeric), ~replace_na(.x, 0)))

lz_wide<- lz_wide %>% mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# detection rates per 500 m - calculated based per segment (as opposed to being calculated across the entire survey, which I think can affect the mean values and the SD for sure)
amph_lz_detections <- amph_lz_wide %>%
  group_by(Start_Date, Station, Survey_Type, Parent_Trail_LongName, Segment, Seg_dist_m) %>%
  reframe(n_frogs = sum(Frog),
            n_sal = sum(Salamander),
            n_liz = sum(Lizard))

frog_detections <- frogs_wide %>%
  select(Start_Date, Station, Survey_Type, Segment, Seg_dist_m, 
         Aromobatidae:Pipidae) %>%
  pivot_longer(names_to = "Family",
               cols = Aromobatidae:Pipidae,
               values_to = "n", names_repair = "unique")

lz_detections <- lz_wide %>%
  select(Start_Date, Station, Survey_Type, Segment, Seg_dist_m, 
         Teiidae:Sphaerodactylidae) %>%
  pivot_longer(names_to = "Family",
               cols = Teiidae:Sphaerodactylidae,
               values_to = "n", names_repair = "unique")

amph_lz_detections <- amph_lz_detections %>%
  mutate(n_frogs_500m = (n_frogs/Seg_dist_m)*500,
         n_sal_500m = 500*(n_sal/Seg_dist_m),
           n_liz_500m = 500*(n_liz/Seg_dist_m))

frog_detections$n_500m <- (frog_detections$n/frog_detections$Seg_dist_m)*500

lz_detections$n_500m <- (lz_detections$n/lz_detections$Seg_dist_m)*500

```

# Prep snake survey data

### 3 guilds

```{r prep three guild snake df}

# make one data frame of only the snake records (remove NAs where no snakes were found) 
surveys_with_snakes <- snakes.sur %>% filter(!is.na(Snake_ID))

# likewise, make one df of only cases where no snakes were found and keep relevant columns with it
surveys_without_snakes <- snakes.sur %>% filter(is.na(Snake_ID)) %>%
  select(Start_Date, Station, Parent_Trail_LongName, Segment, Seg_dist_m)

# get counts of n snakes found per diet type per segment per survey
n_snakes_per_3guilds <- surveys_with_snakes %>%
  group_by(Start_Date, Station, Diet_Type_3categories, Parent_Trail_LongName, Segment, Seg_dist_m) %>%
  reframe(n_snakes = n()) 

# account for cases when no snakes for a given guild were found on a segment
# this is step 1 of accounting for cases where a snake in a given guild wasn't found on a segment and needs to be noted with a zero
n_snakes_per_3guilds <- n_snakes_per_3guilds %>% 
  complete(Diet_Type_3categories, nesting(Start_Date, Segment, Station, Parent_Trail_LongName, Seg_dist_m), fill = list(n_snakes = 0))

# add in the instances where no snakes were found for a given survey
n_snakes_per_3guilds <- bind_rows(n_snakes_per_3guilds, surveys_without_snakes)

# and repeat the process of making sure we have a numerical record (including 0s if none were found) for each diet type for each segment
n_snakes_per_3guilds <-n_snakes_per_3guilds %>% 
  complete(Diet_Type_3categories, nesting(Start_Date, Segment, Station, Parent_Trail_LongName, Seg_dist_m), fill = list(n_snakes = 0))

n_snakes_per_3guilds <- n_snakes_per_3guilds %>% na.omit

```

### Colubridae vs Boidae

```{r prep snake fam df}

# make one data frame of only the snake records (remove NAs where no snakes were found) 
surveys_with_snakes <- snakes.sur %>% filter(!is.na(Snake_ID))

# likewise, make one df of only cases where no snakes were found and keep relevant columns with it
surveys_without_snakes <- snakes.sur %>% filter(is.na(Snake_ID)) %>%
  select(Start_Date, Station, Segment, Seg_dist_m)

# get counts of n snakes found per diet type per segment per survey
n_colub_boid <- surveys_with_snakes %>%
  filter(Family == "Colubridae" | Family == "Boidae") %>%
  group_by(Start_Date, Station, Family, Segment, Seg_dist_m) %>%
  reframe(n_snakes = n()) 

# account for cases when no snakes for a given guild were found on a segment
# this is step 1 of accounting for cases where a snake in a given fam wasn't found on a segment and needs to be noted with a zero
n_colub_boid <- n_colub_boid %>% 
  complete(Family, nesting(Start_Date, Segment, Station, Seg_dist_m), fill = list(n_snakes = 0))

# add in the instances where no snakes were found for a given survey
surveys_without_snakes_boid <- surveys_without_snakes
surveys_without_snakes_boid$Family<- "Boidae"
surveys_without_snakes_col <- surveys_without_snakes
surveys_without_snakes_col$Family<- "Colubridae"

n_colub_boid <- bind_rows(n_colub_boid, surveys_without_snakes_col, surveys_without_snakes_boid)

n_colub_boid[is.na(n_colub_boid)] <- 0

# check, should have 61 unique surveys
length(unique(n_colub_boid$Start_Date))

# each survey should have 2 species
n_colub_boid %>%
  group_by(Start_Date) %>%
  reframe(n_fam = n_distinct(Family))

# checks out

# make column for n found per 500m
n_colub_boid$n_500m <- (n_colub_boid$n_snakes/n_colub_boid$Seg_dist_m)*500
```

# Env data

Format the humidity and understory and the temp data and then get means and add them to the herp dfs

```{r prep env and join to obs dfs}

humidity_long <- env_2025 %>%
  select(Date, Station, Survey_Type, 
         Humidity_1:Humidity_8, Humidity_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Humidity_1:Humidity_Final,
               values_to = "Humidity", names_repair = "unique")
# remove NAs
humidity_long <- humidity_long %>%
  na.omit()

# fix date
humidity_long$Start_Date <- mdy(humidity_long$Date)

humidity_long <- humidity_long %>%
  select(-Date)
# there are 3 humidity readings that are very low (<70) and are all the 1 (or 2nd) reading of the night - this could be error, and our sampling intensity is robust enough for us to not worry about influencing results by removing these outliers. 

humidity_long <- humidity_long %>%
  filter(Humidity > 70)

# repeat with temp
temp_long <- env_2025 %>%
  select(Date, Station, Survey_Type, 
         Temperature_1:Temperature_8, Temperature_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Temperature_1:Temperature_Final,
               values_to = "Temperature", names_repair = "unique")

# remove NAs
temp_long <- temp_long %>%
  na.omit()

# date
temp_long$Start_Date <- mdy(temp_long$Date)
temp_long <- temp_long %>%
  select(-Date)

# understory
understory_long <- env_2025 %>%
  select(Date, Station, Survey_Type, 
         Understory_Density_1:Understory_Density_13, Understory_Density_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Understory_Density_1:Understory_Density_Final,
               values_to = "Understory", names_repair = "unique")
# remove NAs
understory_long <- understory_long %>%
  na.omit()

# fix date
understory_long$Start_Date <- mdy(understory_long$Date)
understory_long <- understory_long %>%
  select(-Date)

# 2024 data

# pull out 2024 year only
env_24_25_nights$Start_Date <- ymd(env_24_25_nights$Start_Date)
env_24_25_nights$Year<- factor(format(env_24_25_nights$Start_Date, "%Y"))
  
env_2024 <- env_24_25_nights %>%
  filter(Year == "2024")

# humidity - 2024 - pivot

# simplify
env_2024 <- env_2024 %>%
  select(-Effort_hrs, -Notes, -Number_of_Small_Mammals_Spotted, -Number_of_Water_Bodies_Spotted, -Surveyors, -Searching_Time, -Meters, -Order, -Survey_Start_Time, -Survey_End_Time)

humidity_long_2024 <- env_2024 %>%
  select(Start_Date, Station,
         Humidity_1:Humidity_8, Humidity_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Humidity_1:Humidity_Final,
               values_to = "Humidity", names_repair = "unique")
# remove NAs

humidity_long_2024 <- humidity_long_2024 %>%
  na.omit()

# repeat with temp
temperature_long_2024 <- env_2024 %>%
  select(Start_Date, Station,
         Temperature_1:Temperature_8, Temperature_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Temperature_1:Temperature_Final,
               values_to = "Temperature", names_repair = "unique")
# remove NAs

temperature_long_2024 <- temperature_long_2024 %>%
  na.omit()

# repeat with understory
understory_long_2024 <- env_2024 %>%
  select(Start_Date, Station,
         Understory_Density_1:Understory_Density_13, Understory_Density_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Understory_Density_1:Understory_Density_Final,
               values_to = "Understory", names_repair = "unique")
# remove NAs

understory_long_2024 <- understory_long_2024 %>%
  na.omit()

# join
# add survey type column to 2024 env data
humidity_long_2024$Survey_Type <-"Night"
temperature_long_2024$Survey_Type <-"Night"
understory_long_2024$Survey_Type <-"Night"

humidity_all <- rbind(humidity_long, humidity_long_2024)
temperature_all <- rbind(temp_long, temperature_long_2024)
understory_all <-rbind(understory_long, understory_long_2024)

# get means
humidity_means <- humidity_all %>%
  filter(Humidity > 73) %>% # two afternoon survey outliers
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(mean_relative_humidity = mean(Humidity))

temperature_means <- temperature_all %>%
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(mean_temp_F = mean(Temperature))

understory_means <- understory_all %>%
  group_by(Station, Start_Date, Survey_Type) %>%
  summarize(mean_understory_leaves = mean(Understory))

# center and scale
humidity_means$humidity_scaled <-scale(humidity_means$mean_relative_humidity)

temperature_means$temp_scaled <-scale(temperature_means$mean_temp_F)

# understory not included in models so not bothering to center and scale here

# add to the amph lz detection data
# fix date first
amph_lz_detections$Start_Date <- ymd(amph_lz_detections$Start_Date)
# add humidity to lizard df
amph_lz_detections <- left_join(amph_lz_detections, humidity_means)
# and add temp
amph_lz_detections <- left_join(amph_lz_detections, temperature_means)

# and add to snakes - same process
# fix dates - guild df
n_snakes_per_3guilds$Start_Date <- mdy(n_snakes_per_3guilds$Start_Date)
# and fam df
n_colub_boid$Start_Date <-mdy(n_colub_boid$Start_Date)
#add in the humidity and temp
n_snakes_per_3guilds <- left_join(n_snakes_per_3guilds, humidity_means)
n_snakes_per_3guilds <- left_join(n_snakes_per_3guilds, temperature_means)
# and do that for snake family df too 
n_colub_boid <- left_join(n_colub_boid, humidity_means)
n_colub_boid <- left_join(n_colub_boid, temperature_means)

# add to frogs, same process

frog_detections$Start_Date<-ymd(frog_detections$Start_Date)
frog_detections <- left_join(frog_detections, humidity_means)
frog_detections <- left_join(frog_detections, temperature_means)

# and to lizards
lz_detections$Start_Date<-ymd(lz_detections$Start_Date)
lz_detections <- left_join(lz_detections, humidity_means)
lz_detections <- left_join(lz_detections, temperature_means)
```

```{r summarize hum and temp and und}

humidity_all$Year <- factor(format(humidity_all$Start_Date, "%Y"))

temperature_all$Year <- factor(format(temperature_all$Start_Date, "%Y"))

understory_all$Year <- factor(format(understory_all$Start_Date, "%Y"))


humidity_all <- humidity_all %>%
  filter(Survey_Type != "Morning")

humidity_summary_yr_survey_type <- humidity_all %>%
  filter(!(Year==2024 & Survey_Type == "Afternoon")) %>%
  filter(!(Survey_Type == "Morning")) %>%
  group_by(Station, Survey_Type, Year) %>%
  reframe(mean_humidity = mean(Humidity),
            sd = sd(Humidity),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# temperature
temperature_all <- temperature_all %>%
  filter(Survey_Type != "Morning")

temperature_summary_yr_survey_type <- temperature_all %>%
  filter(!(Year==2024 & Survey_Type == "Afternoon")) %>%
  filter(!(Survey_Type == "Morning")) %>%
  group_by(Station, Survey_Type, Year) %>%
  reframe(mean_temperature = mean(Temperature),
            sd = sd(Temperature),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# understory
understory_all <- understory_all %>%
  filter(Survey_Type != "Morning")

understory_summary_yr_survey_type <- understory_all %>%
  filter(!(Year==2024 & Survey_Type == "Afternoon")) %>%
  filter(!(Survey_Type == "Morning")) %>%
  group_by(Station, Survey_Type, Year) %>%
  reframe(mean_understory = mean(Understory),
            sd = sd(Understory),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

# Frogs

## Frog summary data

```{r frog summaries}

# n frogs per site total
frogs_df %>%
  filter(Survey_Type != "Morning") %>%
  group_by(Station) %>%
  reframe(total = n())

# n species per site total
frogs_df %>%
  filter(Survey_Type != "Morning") %>%
  group_by(Station) %>%
  reframe(species = n_distinct(Species))
  
# get a summary of n found per survey and per 500m
frogs.mean.se <- amph_lz_detections %>%
  filter(Survey_Type != "Morning") %>%
  group_by(Station) %>%
  summarise(mean = mean(n_frogs_500m),
           sd = sd(n_frogs_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))
# mean individuals found per 500m were 2.25 (0.23 SE, 2.67 SD) in the unhunted study area and 3.21 (0.29 SE, 3.48 SD) in the hunted study area

# what does this look like when we look at total frogs per survey totaled and then divided by n meters surveyed that night, rather than doing it based on per segment numbers?
frogs_per_surv <- amph_lz_detections %>%
  group_by(Station, Survey_Type, Start_Date) %>%
  reframe(Total_m = sum(Seg_dist_m),
          Total_frogs = sum(n_frogs))

frogs_per_surv$n_500m <- (frogs_per_surv$Total_frogs/frogs_per_surv$Total_m)*500


frogs.per.surv.mean.se <- frogs_per_surv %>%
  filter(Survey_Type != "Morning") %>%
  group_by(Station) %>%
  summarise(mean = mean(n_500m),
           sd = sd(n_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))

# the difference is less pronounced
#TBS	2.174821	2.053009	36	0.3421682
#YRS	2.842114	3.251057	37	0.5344704

```

### Frogs

```{r frog per fam plot}

# could also check our visuals of n frogs in each fam per 500m calculated based off cumulative meters surveyed per night/day survey to see if that makes a clearer story

# visualization, SE 
frog_fam_df <- frog_detections %>%
  na.omit() %>%
  group_by(Station, Family) %>%
  reframe(mean_n_per_500m = mean(n_500m),
         sd = sd(n_500m),
           sample_size = n(),
         total_frogs_found = sum(n)) %>%
  mutate(se = sd / sqrt(sample_size))

# A width of 0.9 matches the default behavior of geom_col
spec_dodge <- position_dodge(width = 0.5)

frog_plot <- ggplot(frog_fam_df, aes(x = reorder(Family, -mean_n_per_500m), y = mean_n_per_500m, fill = Station)) +
  geom_errorbar(aes(ymin = mean_n_per_500m - se, ymax = mean_n_per_500m + se), position = spec_dodge, width = 0.2) + labs(fill = NULL)+
  geom_point(position = spec_dodge, color = "black", shape = 21) + scale_fill_manual(values = c("cyan3", "chocolate"), labels = c("Unhunted", "Hunted")) + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") + theme_bw() +
  theme(axis.text.x = element_text(color = "black", face = "plain", angle = 90)) 

frog_plot

# Craugastoridae: Many species, specifically C. megacephalus, in this family have been found to be more abundant when peccaries are around (Reider et al 2013, Biol. Cons, "Effects of collared peccary (Pecari tajacu) exclusion on leaf litter amphibians and reptiles in Neotropical wet forest, Costa Rica") https://doi.org/10.1016/j.biocon.2012.12.015

# Craugastoridae - no tadpole stage, litter dwelling (Oreobates shows strong trend)
rm(frog_fam_df)

# remove cases when < 10 individuals were found in a family for both sites 

remove <- c("Centrolenidae", "Pipidae", "Strabomantidae", "Aromobatidae")

frog_detections.red <- frog_detections %>% 
  filter(!Family %in% remove)

```

## Frog models

We will have to use families in our models

```{r frog models}

# remove morning and instances where env data are missing
frog_detections <- frog_detections %>%
  na.omit()

# pull out year... may need to put this in the models
frog_detections$Year <- factor(format(frog_detections$Start_Date, "%Y"))
```

### Family level analysis

```{r frog family model comparison, warning = FALSE}

# null
frog.null <- glmmTMB(n ~ 1 + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

frog.hunt <- glmmTMB(n ~ Station + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hunt)

simulateResiduals(fittedModel = frog.hunt, plot = T)

testDispersion(frog.hunt, plot = F)
testZeroInflation(frog.hunt, plot = F)

# year
frog.hunt.yr <- glmmTMB(n ~ Station + Year + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())
# year doesn't matter

# humidity
frog.hunt.hum <- glmmTMB(n ~ Station + humidity_scaled + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hunt.hum) # 3083.2 not sig fixed effect

# temperature
frog.hunt.hum.temp <- glmmTMB(n ~ Station + humidity_scaled + temp_scaled + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hunt.hum.temp)
# not sig, worse AIC

# hunting status is sig, better than null
# humidity doesn't make a difference
# year doesn't make a difference
# temperature decreases model fit considerably

# add family
frog.hum.hunt.fam <-glmmTMB(n ~ Station + Family + (1|Segment)+ offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hum.hunt.fam) # AIC 2500.6  big improvement

# Investigate interaction between Family and humidity
frog.hum.hunt.famXhum <-glmmTMB(n ~ humidity_scaled*Family + Station + (1|Segment)+ offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hum.hunt.famXhum) # better, AIC 2486.8

# What about interaction between Family and hunting status instead?
frog.hum.hunt.famXhunt <-glmmTMB(n ~ Station*Family + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hum.hunt.famXhunt) #better, AIC 2469.0 

# what about both interactions?
frog.hunt.famXhunt.famXhum <-glmmTMB(n ~ Station*Family + Family:humidity_scaled + (1|Segment) + offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())

summary(frog.hunt.famXhunt.famXhum) #2466.4 best

# what about humidity*family, no hunting status?
frog.hum.famXhum <-glmmTMB(n ~ Family*humidity_scaled + (1|Segment)+ offset(log(Seg_dist_m)), data = frog_detections, family = nbinom2())
summary(frog.hum.famXhum) # nope

# compare (won't run bc I need to fix names, commenting out for now bc lazy)
#frog_mod_names <- c("frog.null", "frog.hunt", "frog.hunt.yr", "frog.hunt.hum", "frog.hunt.hum.temp", "frog.hum.hunt.fam", "frog.hum.hunt.famXhum", "frog.hum.hunt.famXhunt", "frog.hunt.famXhunt", "frog.hunt.famXhunt.famXhum", "frog.hum.famXhum")

#aictab(cand.set = list(frog.null, frog.hunt, frog.hunt.yr, frog.hunt.hum, frog.hunt.hum.temp, frog.hum.hunt.fam, frog.hum.hunt.famXhum, frog.hum.hunt.famXhunt, frog.hunt.famXhunt, frog.hunt.famXhunt.famXhum, frog.hum.famXhum), modnames = frog_mod_names)

Anova(frog.hunt.famXhunt.famXhum)

```

```{r visualize frog interaction terms}

frog_humidity_scatterplot <- frog_detections %>%
ggplot(aes(x = mean_relative_humidity, y = n_500m, fill = Family)) + geom_point(color = "black", shape = 21, alpha = 0.75, position = "jitter") + geom_smooth(method = "glm", color = "black", linewidth = 0.5) + theme_bw()+ylab("# frogs/500m") + xlab("% relative humidity") + facet_wrap(~Family, scales = "free")

frog_humidity_scatterplot
```


```{r frog families between sites}

# overall
frogs.mean.se <- frog_detections %>%
  group_by(Family, Station) %>%
  summarise(mean = mean(n_500m),
           sd = sd(n_500m),
           n_samps = n()) %>%
  mutate(se = sd / sqrt(n_samps))

spec_dodge <- position_dodge(width = 0.3)

frog_plot <- ggplot(frogs.mean.se, aes(x = reorder(Family, -mean), y = mean, fill = Station)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = spec_dodge, width = 0.2) +
  geom_point(position = spec_dodge, color = "black", shape = 21) + scale_fill_manual(values = c("cyan3", "chocolate"), labels = c("Unhunted", "Hunted")) + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") + theme_bw() +
  theme(axis.text.x = element_text(color = "black", face = "plain", angle = 90)) + labs(x=NULL)  + theme(legend.title = element_blank())

frog_plot

```

# Lizards

## Lizard summary data

```{r liz summaries}

# n per site total
amph_lz %>%
  filter(Gross_ID == "Lizard") %>%
  group_by(Station) %>%
  reframe(total = n())

# n species per site total
amph_lz %>%
  filter(Gross_ID == "Lizard") %>%
  group_by(Station) %>%
  reframe(species = n_distinct(Common_Name))
  
# get a summary of n found per survey and per 500m
lizards.mean.se <- amph_lz_detections %>%
  group_by(Station) %>%
  summarise(mean = mean(n_liz_500m),
           sd = sd(n_liz_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))


```


```{r visualize}

# visualization, SE 
lz_fam_df <- lz_detections %>%
  na.omit() %>%
  group_by(Station, Family) %>%
  reframe(mean_n_per_500m = mean(n_500m),
         sd = sd(n_500m),
           sample_size = n(),
         total_lzs_found = sum(n)) %>%
  mutate(se = sd / sqrt(sample_size))

# Maybe thin out to any cases where we found fewer than 12 individuals of a family?

lz_plot <- ggplot(lz_fam_df, aes(x = reorder(x = Family, -mean_n_per_500m), y = mean_n_per_500m, fill = Station)) +
  geom_errorbar(aes(ymin = mean_n_per_500m - se, ymax = mean_n_per_500m + se), position = spec_dodge, width = 0.2) + labs(fill = NULL)+
  geom_point(position = spec_dodge, color = "black", shape = 21) + scale_fill_manual(values = c("cyan3", "chocolate"), labels = c("Unhunted", "Hunted")) + 
  theme_minimal() + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") +
  theme(axis.text.x = element_text(color = "black", face = "plain", angle = 90)) + labs(x = NULL)

lz_plot


```

## Lizard models

```{r lizard models}

# null
lz.null <- glmmTMB(n ~ 1 + (1|Segment) + offset(log(Seg_dist_m)), data = lz_detections, family = poisson)
summary(lz.null)
# AIC 908.7

# Hunting status (station)
lz.st <-glmmTMB(n ~ Station + (1|Segment) + offset(log(Seg_dist_m)), data = lz_detections, family = poisson)
# AIC 910
summary(lz.st)
# station on it's own doesn't help model

# Family
lz.fam <-glmmTMB(n ~ Family + (1|Segment) + offset(log(Seg_dist_m)), data = lz_detections, family = poisson)

summary(lz.fam)
# AIC 731 (improves, not surprising)

# add humidity
lz.fam.hum <-glmmTMB(n ~ Family + humidity_scaled + (1|Segment) + offset(log(Seg_dist_m)), data = lz_detections, family = poisson)

summary(lz.fam.hum)

# AIC 676.7  huge improvement

# Add temperature
lz.fam.hum.temp <-glmmTMB(n ~ Family+ humidity_scaled + temp_scaled + (1|Segment) + offset(log(Seg_dist_m)), data = lz_detections, family = poisson)

summary(lz.fam.hum.temp) # AIC 676.0 small improvement

# Family, temperature, humidity, and Family:Station interaction
lz.all.station <-glmmTMB(n ~ Family/Station + humidity_scaled + temp_scaled+ (1|Segment)+ offset(log(Seg_dist_m)), data = lz_detections, family = poisson()) # poisson

summary(lz.all.station)
# AIC 670 improves - top model 

Anova(lz.all.station)

simulateResiduals(fittedModel = lz.all.station, plot = T) # passes
testZeroInflation(lz.all.station) # passes
testDispersion(lz.all.station) # passes

```

# Snakes

## guilds snake summary data

```{r snake summaries}
n_snakes_per_3guilds$n_snakes_500m <- 500*(n_snakes_per_3guilds$n_snakes/n_snakes_per_3guilds$Seg_dist_m)

#scale and center predictors
n_snakes_per_3guilds$temp_scaled <- scale(n_snakes_per_3guilds$mean_temp_F)
n_snakes_per_3guilds$humidity_scaled <- scale(n_snakes_per_3guilds$mean_relative_humidity)

n_snakes_per_3guilds %>%
  group_by(Station, Diet_Type_3categories) %>%
  reframe(mean_snakes_500m = mean(n_snakes_500m),
         sd = sd(n_snakes_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))

snakes.mean.se <- n_snakes_per_3guilds %>%
  group_by(Station, Diet_Type_3categories) %>%
  reframe(mean = mean(n_snakes_500m),
           sd = sd(n_snakes_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))

```

## guilds snake models

```{r snake mods}

n_snakes_per_3guilds.na.removed <- n_snakes_per_3guilds %>%
  na.omit()

# null mod
null <- glmmTMB(n_snakes_500m ~ 1 + (1|Segment), data = n_snakes_per_3guilds.na.removed, family = tweedie)

sim <- simulateResiduals(fittedModel = null, plot = T)
# passes

# guild and station
guildXstation <- glmmTMB(n_snakes_500m ~ Diet_Type_3categories/Station + (1|Segment), data =  n_snakes_per_3guilds.na.removed, family = tweedie())

sim <- simulateResiduals(fittedModel = guildXstation, plot = T)
# passes

# compare
snake_mod_names <- c("null", "guildXstation")
aictab(cand.set = list(null, guildXstation), modnames = snake_mod_names)

guildXstation.temp <- glmmTMB(n_snakes_500m ~ Diet_Type_3categories/Station + temp_scaled + (1|Segment), data =  n_snakes_per_3guilds.na.removed, family = tweedie)

sim <- simulateResiduals(fittedModel = guildXstation.temp, plot = T)
# passes validation

# compare
snake_mod_names <- c("null", "guildXstation", "guildXstation.temp")

aictab(cand.set = list(null, guildXstation, guildXstation.temp), modnames = snake_mod_names)

# temp does not make model better

# humidity?
guildXstation.hum <- glmmTMB(n_snakes_500m ~ Diet_Type_3categories/Station + humidity_scaled + (1|Segment), data =  n_snakes_per_3guilds.na.removed, family = tweedie)

# compare
snake_mod_names <- c("null", "guildXstation", "guildXstation.temp", "guildXstation.hum")
aictab(cand.set = list(null, guildXstation, guildXstation.temp, guildXstation.hum), modnames = snake_mod_names)

# humidity makes it a tad better but not sig
summary(guildXstation.hum)

Anova(guildXstation)
```

## family snake summary data

```{r snake fam summaries}

#scale and center predictors

n_colub_boid <- n_colub_boid %>% mutate_at(c(10, 12), ~(scale(.) %>% as.vector))

n_colub_boid <- n_colub_boid %>% mutate(Survey_Type = ifelse(is.na(Survey_Type), "Night", Survey_Type))

n_colub_boid %>%
  group_by(Station, Family) %>%
  reframe(mean_snakes_500m = mean(n_500m),
         sd = sd(n_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))

snakes.boid.colub.mean.se <- n_colub_boid %>%
  group_by(Station, Family) %>%
  summarise(mean = mean(n_500m),
           sd = sd(n_500m),
           n = n()) %>%
  mutate(se = sd / sqrt(n))

```


## family snake models

```{r family snake data exploration}

# remove NAs because we need to compare AIC value with and without env vars (which have a few NAs)

n_colub_boid.na.removed <- n_colub_boid %>%
  na.omit()

# what do the data look like?
plot(n_colub_boid.na.removed$temp_scaled, n_colub_boid.na.removed$n_500m) #
plot(n_colub_boid.na.removed$mean_temp_F, n_colub_boid.na.removed$n_500m) #
# humidity?
plot(n_colub_boid.na.removed$mean_relative_humidity, n_colub_boid.na.removed$n_500m)

hist(n_colub_boid.na.removed$n_500m)
```

```{r snake fam mods} 

# null mod
# tweedie looks ok but doesn't pass
# try nbinom with count data 
fam.null <- glmmTMB(n_500m ~ 1 + (1|Segment), data = n_colub_boid.na.removed, family = tweedie)

simulateResiduals(fittedModel = fam.null, plot = T)
testZeroInflation(fam.null)
testDispersion(fam.null)

# family and station
familyXstation<- glmmTMB(n_500m ~ Family/Station + (1|Segment), data = n_colub_boid.na.removed, family = tweedie)

simulateResiduals(fittedModel = familyXstation, plot = T)
testZeroInflation(familyXstation)
testDispersion(familyXstation)

familyXstation.temp <- glmmTMB(n_500m ~ Family/Station + temp_scaled + (1|Segment), data =  n_colub_boid.na.removed, family = tweedie)

simulateResiduals(fittedModel = familyXstation.temp, plot = T)
testZeroInflation(familyXstation.temp)
testDispersion(familyXstation.temp)
# temp does not make model better

# humidity?

familyXstation.humidity <- glmmTMB(n_500m ~ Family/Station + humidity_scaled + (1|Segment), data =  n_colub_boid.na.removed, family = tweedie)

simulateResiduals(fittedModel = familyXstation.humidity, plot = T)
testZeroInflation(familyXstation.humidity)
testDispersion(familyXstation.humidity)
# humidity does not make model better

# humidity and temp

familyXstation.humidity.temp <- glmmTMB(n_500m ~ Family/Station + humidity_scaled + temp_scaled + (1|Segment), data =  n_colub_boid.na.removed, family = tweedie)

simulateResiduals(fittedModel = familyXstation.humidity.temp, plot = T)
testZeroInflation(familyXstation.humidity.temp)
testDispersion(familyXstation.humidity.temp)
# humidity does not make model better

Anova(familyXstation)
```


# Figures for detection rates

```{r poster figure}

frog_plot <- ggplot(frogs.mean.se, aes(x = reorder(Family, -mean), y = mean, fill = Station, shape = Station)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = spec_dodge, width = 0.2, linewidth = 0.2) +
  geom_point(position = spec_dodge, color = "black") + scale_shape_discrete(labels = c("Unhunted", "Hunted")) + theme(legend.title = element_blank()) + ylab("Individuals per 500m")+ labs(x=NULL) + labs(title = "Frog detection rates") +  theme(legend.title = element_blank())+ theme_bw() +
  theme(axis.text = element_text(size = 12, color = "black")) +theme(axis.text.x = element_text(color = "black", face = "plain", angle = 45, vjust = 1)) 

lz_plot <- ggplot(lz_fam_df, aes(x = reorder(x = Family, -mean_n_per_500m), y = mean_n_per_500m, fill = Station,shape = Station)) +
  geom_errorbar(aes(ymin = mean_n_per_500m - se, ymax = mean_n_per_500m + se), position = spec_dodge, width = 0.2, linewidth = 0.2) + labs(fill = NULL)+
  geom_point(position = spec_dodge, color = "black") + scale_shape_discrete(labels = c("Unhunted", "Hunted")) + theme(legend.title = element_blank()) + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") +   theme(axis.text.x = element_text(color = "black", face = "plain", angle = 45)) + labs(x = NULL) + theme(legend.text = element_text(size = 12)) + labs(title = "Lizard detection rates") + theme_bw()+
  theme(axis.text = element_text(size = 12, color = "black")) + theme(axis.text.x = element_text(color = "black", face = "plain", angle = 45, vjust = 1)) 
 

lz_frog_plot <- ggarrange(frog_plot, lz_plot, labels = c("a)", "b)"), common.legend = TRUE)

lz_frog_plot 


```

```{r smamm and snake}

smamm_plot <- ggplot(smams.distance.mean.se, aes(x = Station, y = mean, shape = Station)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = spec_dodge, width = 0.2, linewidth = 0.3) + labs(fill = NULL)+
  geom_point(position = spec_dodge, size = 1) + scale_shape_discrete(labels = c("Unhunted", "Hunted")) + 
  theme_minimal() + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") + labs(x = NULL)+ theme(legend.title = element_blank()) +
  theme(axis.text = element_text(size = 12, color = "black"))+ scale_x_discrete(labels = c("Unhunted", "Hunted")) +
  theme(axis.text = element_text(size = 12, color = "black")) + theme(legend.text = element_text(size = 12))

smamm_plot

# boidae vs collubridae

boid_colub_plot <- ggplot(snakes.boid.colub.mean.se, aes(x = Family, y = mean, shape = Station)) +  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = spec_dodge, width = 0.2, linewidth = 0.3) + labs(fill = NULL) + theme_minimal() + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") + labs(x = NULL) +
  geom_point(position = spec_dodge, size = 1, color = "black") + scale_shape_discrete(labels = c("Unhunted", "Hunted")) + theme(legend.title = element_blank()) +
  theme(axis.text = element_text(size = 12, color = "black")) +
  theme(axis.text = element_text(size = 12, color = "black")) + theme(legend.text = element_text(size = 12))

boid_colub_plot

# diet type

snake_3_guild_plot <- ggplot(snakes.mean.se, aes(x = Diet_Type_3categories, y = mean, shape = Station)) +  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = spec_dodge, width = 0.2, linewidth = 0.3) + labs(fill = NULL) + theme_minimal() + theme(axis.text.y = element_text(color = "black")) + ylab("Individuals per 500m") + labs(x = NULL) + 
  geom_point(position = spec_dodge, size = 1, color = "black") + scale_x_discrete(labels = c("Mixed diet", "Mammal-based diet", "Excludes mammals")) + scale_shape_discrete(labels = c("Unhunted", "Hunted")) + theme(legend.title = element_blank()) +
  theme(axis.text = element_text(size = 12, color = "black")) + theme(legend.text = element_text(size = 12))

snake_3_guild_plot

det.plot <- ggarrange(smamm_plot, snake_3_guild_plot, boid_colub_plot, labels = c("a)", "b)", "c)"), nrow = 1, common.legend = TRUE)


det.plot

#ggsave(det.plot, filename= "smamm_snake.png")
```

