---
title: "Snake_Analysis"
author: "Owen McCool, Therese Lamperty"
date: "2025-07-16"
output: word_document
---

# Code Setup & package read in 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read in packages}
library(vegan)
library(stringr)
library(ggplot2)
library(tidyverse)
library(iNEXT) # species diversity
library(hillR) # species diversity
library(DHARMa) # checking models
library(car) # reporting model stats
library(glmmTMB)
library(unmarked) #density estimates with hierarchical distance sampling models
library(tidyr)
library(AICcmodavg) # has convenient functions for comparing models based on AIC

```

# Project Overview

Study Site: Research was split between two different zones within the greater Yasuni Biosphere Reserve. Tiputini Biodiversity Station (TBS) served as the faunally intact site, whereas Yasuni Research Station (YRS) served as the defaunated site. Both sites are situated along the Tiputini River within the province of Orellana, Ecuador. TBS, approximately 45 km east of YRS, is only accessible by boat. YRS is along the Maxus Oil Road, and is less integrated into the Amazonian ecosystem. Additionally, unsustainable hunting activities by the indigenous communities have been prominent in the surrounding region at YRS for multiple decades, as noted from 2014. This is thought to be the root cause of defaunation in this zone. 

	I transferred between the two research stations every two weeks, starting my first rotation at TBS on June 10. Time at each station was spent working on 3 separate projects: my herpetofauna surveys, frugivory surveys, and an overarching seed enemy and dispersal project. 

	The rainy season in Northeastern Ecuador is at its peak between March and July. I conducted my study between the months of June and August, 2024. The average high temperature for June in this region is 85.66° F, with the average low of 70.05° F. July has an average high temperature of 86.29° F and an average low of 69.13° F. August has an average high temperature of 91.35° F and an average low of 69.60° F. June receives 143mm of precipitation on average each year, whereas July receives 99.69mm and August receives 77.55mm. (https://weatherandclimate.com/ecuador/orellana/boca-tiputini) 
Put project description here (field sites, time frame of collection,sampling scheme, type of data (eg. species ID,temp, etc.) )

Surveys: Herpetofauna were sampled using visual surveys on foot by two surveyors/night. Trail systems at each site served as the sampling paths. The surveys were repeated around 6 times on each rotation while in the field, and resampled on the second rotations. The surveys typically began around 20:00-20:30 and lasted until 23:00-00:00 CST. At each site, habitat matches were compared and assessed and trails through that habitat were consistently walked. Two different types of habitat, Terra Firma (uplands) and Varzea (flooded forest) were surveyed on the herp nocturnal walks. In addition to this, two significant bodies of water were also surveyed at each site, primarily consisting of oxbow lakes with one year-round large pond. 

	Surveyors used FENIX HM60R headlamps while conducting night surveys. They carefully examined all surrounding habitat, including understory trees, shrubs, vines, water bodies, forest floor, and middle-canopy. Amphibians of the order Anura gave off reflective eyeshine and were easier to sample for, whereas members of Squamata did not. More information about this will be within the discussion section. Effort hours will be calculated based on survey time length and number of surveyors. 

	Additionally, environmental information was taken throughout the course of each survey. Temperature and humidity data was collected with a Mengshen meter in 30 minute intervals on surveys. Understory vegetation density was also measured every 15 minutes along each trail. A 1.5m PVC pipe was randomly placed upright just off the right side of the trail at each measuring point, and the number of leaves making contact with the pipe was counted. 

	Weather and wind were noted before each survey in four nominal categories, with weather either being “clear, partly cloudy, cloudy, rainy” and wind being “no wind, felt on face, branches move, trees move”. I also accounted for precipitation that had already occurred in 6-hour blocks as “time window of last rain event”. 

	Once an animal was found, the first step of the protocol was to note what time of night the find occurred. After this, the temperature and humidity was once again collected. A rangefinder was used to estimate the perpendicular distance of each animal from the trail. A GPS was used to mark and log the coordinates of each animal found. Animals were given a two-letter code based on the Order and Suborder followed by a number, indicating its position of being found. For example, the code of the 9th snake found of the season would be SK09. Species found were also identified to the lowest taxonomic level in the field, and photographs of individuals were taken both to get positive IDs later in the lab when not possible in the field, but also to avoid counting re-captures.

	Snakes found during non-survey times were also recorded. These individuals were split into one of two categories of detection. The first is an “Opportunistic Encounter” (OE), which is any snake that was found by me while out working on other non-herpetofauna projects in the field. The other type of detection is the Third Party Encounter (TPE). This was any snake that was found by other research groups during my rotations at each station, identified to lowest possible taxonomy. The same data was taken for these individuals. GPS coordinates, time of detection, trail, observer, estimation of distance from trail, and habitat type. Because the third-party observers often didn’t have access to the climate data logger, this information was retrieved later on from each station’s respective weather tower data, based on the time of detection. 

Snakes were given a classification as to whether they were small mammal consumers or not. Information on this was retrieved from literature such as online publications, the book Reptiles and Amphibians of the Amazon by R.D. and Patricia Bartlett, and the expert field book Reptiles of Ecuador. 

# Read in and prep data

## Read in

This data was cleaned some in initial steps in 2024-2025 (see archived code). Note that temp and humidity data was removed for 4 surveys at YRS that took place after July 11, 2025, during which the team lost confidence in the data collection instrument. 

```{r data read in}

#1. all snake records
snakes.master <- read.csv(file = "Data/snake.master.csv")

#2. environmental data, which also has rodent encounters
enviro <- read.csv(file = "Data/enviro.csv")

#3. segment data, has lengths of segments
segs <- read.csv("Data/Segment_Distances.csv")

#4. amphibian and lizard data collected in 2025
amph_lz=read.csv("Data/Field_Season_2_datasheets(amphibians_and_lizards).csv")

#5. corresponding amph_lz env data
env=read.csv("Data/Field_Season_2_datasheets(Environment_Sheet).csv")

#6. this may be the same as 'segs' above
seg_dist=read.csv("Data/Field_Season_2_datasheets(Segment_Distances).csv")

#7. amphibian 2024 data
amph_2024 <- read.csv("Data/2024_amphibians.csv")

#8. lizard 2024 data
liz_2024 <- read.csv("Data/2024_lizards.csv")
```

## Data prep

We need to make a dataframe that contains the dates that each segment was surveyed, preferably also merged with the length of each of those segments.

```{r segment master list}
 
segs_expanded <- enviro %>%
  select(!c(6,9,11:45)) %>%
  separate_rows(Order, sep = ",")

# get ready to add the meters per segment. Need names to match
segs <- segs %>%
  rename(Segment = SEGMENT) %>%
  rename(Parent_Trail = PARENT.TRAIL) %>%
  rename(Seg_dist_m = DISTANCE_M)

# and fix the names in segs_expanded to match and to be easy to understand
segs_expanded <- segs_expanded %>%
  rename(Segment = Order) %>%
  rename(Parent_Trail_LongName = Trail_Walked)

# change this in enviro as well
enviro <- enviro %>%
  rename(Parent_Trail_LongName = Trail_Walked)

# join them together
segments_master <- left_join(segs_expanded, segs)

# there are some missing distances
# check for typos
unique(segs_expanded$Segment)

# there's just one where there is a hyphen separating two segments that need to be in separate rows
segs_expanded <- segs_expanded %>%
  mutate(Segment = ifelse(Segment == "NUMA-02-CHIO-03", 
                          "NUMA-02, CHIO-03", Segment)) %>%
  separate_rows(Segment, sep = ",")

# remove all the spaces
segs_expanded <- segs_expanded %>%
  mutate(Segment = trimws(Segment))

# fix one case where there is just a space between two segments that need to be in separate rows
segs_expanded <- segs_expanded %>%
  separate_rows(Segment, sep = " ")

# okay this looks more promising, try to join again
segments_master <- left_join(segs_expanded, segs)

# T note 4/21/2026 - check for duplicated entries - I think that CHIO-03 from 2024 is in here 2x

```

Okay now other little things with the datasets

```{r df prepping}

# make dates dates
snakes.master$Start_Date <- mdy(snakes.master$Start_Date)

snakes.master$Date <- mdy(snakes.master$Date)

enviro$Start_Date <- mdy(enviro$Start_Date)

segments_master$Start_Date<-mdy(segments_master$Start_Date)
segments_master$End_Date<-mdy(segments_master$End_Date)

# modify diet type to be binary = eats mammals, does not eat mammals
unique(snakes.master$Diet_Type)
# keep original too
snakes.master$Diet_Type_3categories <- snakes.master$Diet_Type

# mixed = "includes mammals"
snakes.master <- snakes.master %>%
  mutate(Diet_Type = ifelse(Diet_Type == "mixed", 
                          "includes_mammals", Diet_Type)) 
# mostly mammals = includes mammals
snakes.master <- snakes.master %>%
  mutate(Diet_Type = ifelse(Diet_Type == "mostly_mammals", 
                          "includes_mammals", Diet_Type))

# sampling effort
snakes.sur <- snakes.master %>%
  filter(Type_of_Encounter == "Survey") #piping with dplyr, make new dataframe with only surveyed snakes since those are only obs. with effort associated (not sure if this is needed since effort hours is now its own column)

#Now we want to put in an environment sheet because it has our effort hours, and we want to convert those to usable numeric values

# NOTE this is the time spent actively searching, not the time difference between start and stop of survey for the night

enviro$Effort_hrs <- sapply(strsplit(enviro$Searching_Time, ":"), function(x) {
  as.numeric(x[1]) + as.numeric(x[2]) / 60
}) #not really sure how this function works (from internet) but it makes the hours go from 2:30 to 2.50 hours, etc. 

# join date and time columns first
# note I added end date columns manually in the enviro datasheet to account for the 2 surveys that went past midnight
enviro$Survey_Start_Date_Time <- paste(enviro$Start_Date, enviro$Survey_Start_Time)

# use lubridate to make these date and time in R
enviro$Survey_Start_Date_Time <-ymd_hm(enviro$Survey_Start_Date_Time)

# repeat with end date and time
enviro$Survey_End_Date_Time <- paste(enviro$End_Date, enviro$Survey_End_Time)

enviro$Survey_End_Date_Time <-mdy_hm(enviro$Survey_End_Date_Time)

# small mammals 
 # add the n small mammals found per survey to the snake.sur dataframe because that has the effort hours already in it

# subset data - just want the n small mammmals and keep the date column to join by

small.mamms.temp <- enviro %>%
  select(Station, Start_Date, Number_of_Small_Mammals_Spotted, Parent_Trail_LongName) #creating temporary dataframe with just these parameters

small.mamms.temp$Number_of_Small_Mammals_Spotted <-  as.numeric(small.mamms.temp$Number_of_Small_Mammals_Spotted) #changing the n small mammals to a number from a character (idk why it was a character)
class(small.mamms.temp$Number_of_Small_Mammals_Spotted) #looks correct now

# isolate effort hours per survey
effort <- enviro %>%
  select(Start_Date, Effort_hrs)

# join effort to small mamms
small.mamms.temp <- left_join(small.mamms.temp, effort)

# clean 
rm(effort)

# get rate of n mammals per hour
small.mamms.temp$smammals.hr <- small.mamms.temp$Number_of_Small_Mammals_Spotted/small.mamms.temp$Effort_hrs

# small mammal dataframe done

# add segment lengths to snake sur df
snakes.sur <- snakes.sur %>%
  rename(Segment = Location)
```

```{r check and fix tricky join}

length(unique(segments_master$Segment)) # 89 unique segments surveyed
length(unique(segments_master$Start_Date)) #61 surveys

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Segment)) # snakes were found on 46 of them

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Snake_ID)) # 101 snakes found

length(unique(snakes.sur$Start_Date)) #60, what date is in segments master but not in snakes.sur? (it must be the duplicated one from start of season one)

missing_startdate <- anti_join(segments_master, snakes.sur, by = "Start_Date")
 
# FLAG 1: MAPO-01 MAPO-02 HARP-01 on 2024-06-21 are in segments master but not in the record of surveys that were pulled from snake.master dataframe. Can we confirm that these surveys happened? Was there just no enviro data that went with them?

# okay so do we have distances for all segments snakes were found on?

missing_segs <- anti_join(segments_master, snakes.sur, by = "Segment")

# 157 instances of segments being surveyed and no snakes were found?

# also, in theory, segments master has all the cases that a survey occurred, and so I can remove the cases of surveys with no snakes found from the snakes.sur dataframe before merging (double check with Owen that this sounds right)

snakes.sur.zeros.dropped <- snakes.sur %>%
 filter(!is.na(Snake_ID))

snakes.sur.zeros.dropped <- snakes.sur.zeros.dropped %>%
  select(Start_Date, Station, Segment, Snake_ID, Distance_From_Trail_m, Diet_Type, Diet_Type_3categories, Family, Genus, Species, Common_Name, Temperature_at_Time_of_Detection, Humidity_at_Time_of_Detection)

snakes.sur <- full_join(segments_master, snakes.sur.zeros.dropped, relationship = "many-to-many")

length(unique(snakes.sur$Segment)) # 89 unique segments surveyed

length(unique(snakes.sur$Start_Date)) #61 surveys

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Segment)) # snakes were found on 46 of them

snakes.sur %>% filter(!is.na(Snake_ID)) %>% summarize(n_distinct(Snake_ID)) # 101 snakes found

# FLAG 2: NOTE TO THERESE: SOMEHOW IN SNAKES.SUR, SK04 AND SK05 GET DUPLICATED BUT IT DOESN'T SEEM TO CAUSE ERRORS LATER ON DOWN THE LINE. EG, THE DENSITY SECTION HAS EXACTLY 101 SNAKES WHICH IS THE CORRECT AMOUNT SO I'M NOT SURE WHAT'S GOING WRONG HERE 

# *reply 4/21/2026 T: I think that I caught that and removed those duplicates at some point in the code. We could (and should) remove them earlier on though - just do that either in the raw datafile if it's truly an error or maybe here.

rm(snakes.sur.zeros.dropped, missing_startdate)

```

### 2024 amphibian and lizard data

```{r clean 2024 amph lz data}
# Start with amphibians
amph_2024 <- amph_2024[-2,]
colnames(amph_2024) <- amph_2024[1,]
amph_2024 <- amph_2024[-1,]

amph_2024 <- amph_2024 %>%
  rename(Time_Found = Time_Found_)
amph_2024 <- amph_2024 %>%
  mutate(Survey_Type = "Night", 
         X = NA, 
         X.1 = NA, 
         Size = NA,
         Diet_Type = NA,
         Primary_Diet_Component = NA)


# Now for lizards
liz_2024 <- liz_2024[-2,]
colnames(liz_2024) <- liz_2024[1,]
liz_2024 <- liz_2024[-1,]

liz_2024 <- liz_2024 %>%
  rename(Time_Found = Time_Found_)
liz_2024 <- liz_2024 %>%
  mutate(Survey_Type = "Night", 
         X = NA, 
         X.1 = NA, 
         Size = NA,
         Diet_Type = NA,
         Primary_Diet_Component = NA)

```

```{r organize and clean species data, warning=FALSE}

amph_lz=amph_lz %>%
  separate(Species,into=c("Genus","Species")) #turn Species column into Genus and Species columns

# replace unknown family with NA
amph_lz <- amph_lz %>%
  mutate(Family = ifelse(Family == "weird" | Family == "mystery" | Family == "weird ground frog in valley, looked like big head x treefrog" | Family == "unknown", NA, Family))

# replace blank Genus and Unknown Genus entries with NA
amph_lz <- amph_lz %>%
  mutate(Genus = ifelse(Genus == "" | Genus == "Unknown", NA, Genus))

amph_lz$Species= na_if(amph_lz$Species, "spp") #replace "spp" with NA

# Check to make sure  column names are the same
colnames(amph_2024)
colnames(amph_lz)

colnames(liz_2024)
colnames(amph_lz)

# MERGE ALL DATA INTO AMPH_LZ

amph_lz <- rbind(amph_lz, amph_2024, liz_2024)

rm(amph_2024)
rm(liz_2024)

rm_trail_spaces <- function(column) {
  clean_column <- trimws(column)
  return(clean_column)
}

# Get rid of duplicate stuff

amph_lz$Gross_ID <- rm_trail_spaces(amph_lz$Gross_ID)
amph_lz$Genus <- rm_trail_spaces(amph_lz$Genus)
amph_lz$Survey_Type <- rm_trail_spaces(amph_lz$Survey_Type)
amph_lz$Station <- rm_trail_spaces(amph_lz$Station)
amph_lz$Weather <- rm_trail_spaces(amph_lz$Weather)
amph_lz$Animal_ID <- rm_trail_spaces(amph_lz$Animal_ID)


```

```{r more prep and cleaning}
colnames(amph_lz)

amph_lz$Start_Date_Start_Time=paste(amph_lz$Date,amph_lz$Survey_Start_Time,sep=" ")
amph_lz$End_Date_End_Time=paste(amph_lz$Date,amph_lz$Survey_End_Time,sep=" ")

amph_lz$Start_Date_Start_Time=mdy_hm(amph_lz$Start_Date_Start_Time)
amph_lz$End_Date_End_Time=mdy_hm(amph_lz$End_Date_End_Time)

#### Date ####

amph_lz$Date=mdy(amph_lz$Date)

#### Survey_Type ####
amph_lz$Survey_Type=trimws(amph_lz$Survey_Type)

#### Survey_hr ####
amph_lz$Survey_hr=difftime(amph_lz$End_Date_End_Time,amph_lz$Start_Date_Start_Time,units="hours")

#### Gross_ID () ####
amph_lz$Gross_ID=trimws(amph_lz$Gross_ID)

#### Genus ####
amph_lz %>%
  distinct(Genus) %>% arrange(Genus)

#### Time_Found ####
amph_lz$Time_Found=hm(amph_lz$Time_Found)

```

We need to make a dataframe that contains the dates that each segment was surveyed, preferably also merged with the length of each of those segments.

```{r segment master list}

# env should have every segment ever surveyed ... and each time it was surveyed
segs_survey_record <- env %>%
  select(c(1,2,3,5)) %>%
  separate_rows(Order, sep = ",")

# and fix the names in segs_expanded to match and to be easy to understand
segs_survey_record <- segs_survey_record %>%
  rename(Segment = Order)

# there's just one where there is a hyphen separating two segments that need to be in separate rows
#segs_survey_record <- segs_survey_record %>%
 # mutate(Segment = ifelse(Segment == "NUMA-02-CHIO-03", 
  #                        "NUMA-02, CHIO-03", Segment)) %>%
  #separate_rows(Segment, sep = ",")

# remove all the spaces
segs_survey_record <- segs_survey_record %>%
  mutate(Segment = trimws(Segment))

# fix one case where there is just a space between two segments that need to be in separate rows
segs_survey_record <- segs_survey_record %>%
  separate_rows(Segment, sep = " ")

# get ready to add the meters per segment. Need names to match
seg_dist <- seg_dist %>%
  rename(Segment = SEGMENT) %>%
  rename(Parent_Trail = PARENT.TRAIL) %>%
  rename(Seg_dist_m = DISTANCE_.M.) %>%
  select(!Parent_Trail)

# first, do we have distances for all segments named in the segs_survey_record df?

# one typo - 	
#CEIBA-01 should be CEIB-01
segs_survey_record <- segs_survey_record %>%
  mutate(Segment = ifelse(Segment == "CEIBA-01", "CEIB-01", Segment))

segs_survey_record_with_distances_added <- left_join(segs_survey_record, seg_dist)

```

```{r join segments and survey info to df}

# reduce down to the columns we're interested in right now
amph_lz_segs <- amph_lz %>%
  select(Date, Station, Segment, Animal_ID, Distance_From_Trail_m, Gross_ID, Family, Genus, Species, Common_Name, Survey_Type, Habitat_Type, Height_of_Individual)

# fix date name to match
amph_lz_segs <- amph_lz_segs %>%
  rename(Start_Date = Date)

# Find segs in segments master that are not in amph_lz
segs_in_master_not_surveys <- anti_join(segments_master, amph_lz_segs, by = "Segment")

# Owen/Jacob we should do a check here, does it seem believable that 8 segments in 2025 had 0 specimens found, and that 6 segments in 2024 had 0 specimens found?

# vice versa
# should be 0
segs_in_surveys_not_master <- anti_join(amph_lz_segs, segments_master, by = "Segment")

#FLAG 3: Few things missing, all of which are afternoon surveys: 2024-06-05, TCHO-02 and MATA-01 afternoon. Also - 2025-06-23 HARP-04, -03, -02 and GUAC-04. We just need the corresponding segment length for these and the habitat type.

# T - 4/21/2026, stopping here - once we get the above seg data in, proceed with joining that info to the lizard/frog/salamander data 

# add that information to the survey data
amph_lz_segs <- left_join(segs_survey_record_with_distances_added, amph_lz_segs)

# make sure nothing disappeared
amph_lz_segs %>% group_by(Station, Gross_ID) %>% 
  summarize(total_found = n_distinct(Animal_ID),
            species_found = n_distinct(Species))

```

*Stopped in the above chunk (see 4/21/2026 note and all FLAG comments)*

# Data summaries

## Sampling Time and Distance

```{r survey samp effort}

# pull out mean and standard deviation overall and for each station
mean(enviro$Effort_hrs)
sd(enviro$Effort_hrs)

# per station - this will give output at bottom of code chunk and should be reported in the Results or Methods section
enviro %>%
  group_by(Station) %>%
  summarise(mean_duration = mean(Effort_hrs),
            standard_deviation = sd(Effort_hrs))

#Also want to see what the distance effort is

class(enviro$Meters) #says it is a character and can't be summed, we need to convert it
enviro$Meters <- as.numeric(enviro$Meters) #now they are numbers

sum(enviro$Meters[enviro$Station == "TBS"]) # 73,864 meters at TBS
sum(enviro$Meters[enviro$Station == "YRS"]) # 76,915 meters at YRS

# per night meters
enviro %>%
  group_by(Station) %>%
  summarise(mean_m_per_night = mean(Meters),
            standard_deviation = sd(Meters))
```

## Small mammals

```{r summarize small mammal numbers}
#Raw mammal counts

sum(small.mamms.temp$Number_of_Small_Mammals_Spotted[small.mamms.temp$Station == "YRS"])

sum(small.mamms.temp$Number_of_Small_Mammals_Spotted[small.mamms.temp$Station == "TBS"])

# I think this is good to report (total numbers) but maybe also with mean and sd of n encountered per survey at each station:

enviro %>%
  group_by(Station) %>%
  summarise(mean_detections_per_survey = mean(Number_of_Small_Mammals_Spotted),
            standard_deviation = sd(Number_of_Small_Mammals_Spotted))

```

# Small Mammals

## Smamm Per hr boxplot

```{r small mammal boxplot}

# make a boxplot!

smammals.plot <- ggplot(data = small.mamms.temp, aes(x = Station, y = smammals.hr, fill = Station)) + geom_boxplot() + theme_bw() + scale_fill_manual(values = c("cornflowerblue", "goldenrod1"), labels = c("Unhunted", "Hunted")) + xlab("Station") +
  theme(legend.title=element_blank()) + ylab("N Small Mammals Detected Per Hour") + ggtitle("Small Mammal Detection Rates by Station") 

smammals.plot #view the plot

# clean
rm(smammals.plot)

# save plot to computer files
#ggsave(smammals.plot, file="small_mammal_plot.png", width = 6, 
      # height = 4)
# can add dimension specifications with width = x and height = x
```

## Smamm Per hr mean and se

```{r smamm per hr se and mean}

# another way to plot that makes judging significance easier NEEDS EXPLAINING
smams.mean.se <- small.mamms.temp %>%
  group_by(Station) %>%
  summarise(mean = mean(smammals.hr),
            sd = sd(smammals.hr),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

smamm.fig <- ggplot(smams.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) +
  geom_jitter(data = small.mamms.temp, 
              aes(x = Station, y = smammals.hr), 
              width = 0.15, size = 1.5, alpha = 0.5) + coord_fixed(ratio = 0.75) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21, size = 3) + theme_bw() + scale_fill_manual(values = c("cornflowerblue", "goldenrod1")) + xlab("") + scale_color_discrete(labels = "") + ylab("Individuals/Hour") + ggtitle("Small Mammal Detections in Hunted\nand Unhunted Forest") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(
    legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(family = "Tahoma", margin = margin(t = 8)), axis.text.y = element_text(family = "Tahoma", margin = margin(l = 8)), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5, face = "bold", size = 12, margin = margin(b = 12))) 

smamm.fig #view plot, also seems significant NEED TO UNDERSTAND GGPLOT CODE

#ggsave(smamm.fig, file="small-mammal-mean-se.png", width = 6,  height = 6)

rm(smams.mean.se, smamm.fig)
```

## Smamm Model

```{r small mammal model}

# note: nbinom2 with offset term and zero inflation term will run and seems okay, but it doesn't pass validation cleanly. Other attempts (with and without random effects) cause model convergence problems. 

# Model as continuous response variable rather than raw count data

smam.cont <- glmmTMB(smammals.hr ~ Station + (1|Parent_Trail_LongName), data = small.mamms.temp, family = tweedie(link = "log"))

sim <- simulateResiduals(fittedModel = smam.cont, plot = T)

summary(smam.cont) # great, can report this

testDispersion(smam.cont, plot = F)
testZeroInflation(smam.cont, plot = F)

# could report in this format:
Anova(smam.cont)

#https://stats.stackexchange.com/questions/603155/seeking-feedback-on-the-interpretation-and-reporting-of-a-glmm-output
```


```{r clean env after smamm}
rm(sim, small.mamms.temp, smam.cont)
```

# Understory

## Understory Set-Up

Get understory data reformatted. We took these readings every 15 min. First step will be looking at variation between stations. 

```{r reformat understory}
understory_long <- enviro %>%
  select(Survey_Start_Date_Time, Station, Parent_Trail_LongName, Understory_Density_1:Understory_Density_13, Understory_Density_Final) %>%
  pivot_longer(names_to = "Measurement",
               cols = Understory_Density_1:Understory_Density_Final,
               values_to = "Understory_leaves", names_repair = "unique")

# remove NAs

understory_long <- understory_long %>%
  na.omit()

```

## Und Figure

Histogram per station

```{r understory histo}

under.hist <- ggplot(understory_long, aes(x= Understory_leaves)) +
  geom_histogram() +facet_wrap(~Station) + theme_bw()

under.hist

# summary
understory_long %>%
  na.omit() %>%
  group_by(Station) %>%
  summarise(mean = mean(Understory_leaves),
            sd = sd(Understory_leaves),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

```

And do a mean and se with points overlain

```{r understory mean se figure}

und.mean.se <- understory_long %>%
  na.omit() %>%
  group_by(Station) %>%
  summarise(mean = mean(Understory_leaves),
            sd = sd(Understory_leaves),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

und.fig <- ggplot(und.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) + geom_jitter(data = understory_long, aes( x = Station, y = Understory_leaves), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.45) + theme_bw() +
  scale_fill_manual(values = c("lightgreen","lightgreen")) + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5) + geom_point(pch = 21, size = 3) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Understory (vertical leaf overlap)") + ggtitle("Understory Density") + theme_bw(base_family = "Tahoma") + theme(plot.title = element_text(face = "bold", hjust = .5))  + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(legend.position = "none") + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

und.fig

#ggsave(und.fig, file="understory-mean-se.png", width = 6,  height = 4)

# rm(und.fig)
```

# Temperature

## Temp Set-Up

Get temperature data reformatted. We took these readings every 30 min and also every time an individual was found. First step will be looking at variation between stations. 

```{r reformat temperature}
# First, let's add column to differentiate by season (2024 or 2025)
enviro$Season <- factor(format(enviro$Start_Date, "%Y"))

# pull out per data record enviro data from snake.sur

env_at_detection <- snakes.sur %>%
  select(Station, Start_Date, Temperature_at_Time_of_Detection, Humidity_at_Time_of_Detection) %>% na.omit()

enviro2 <- left_join(enviro, env_at_detection)

temperature_long <- enviro2 %>%
  select(Survey_Start_Date_Time, Station, Season, Parent_Trail_LongName, 
         Temperature_1:Temperature_7, Temperature_Final, Temperature_at_Time_of_Detection) %>%
  pivot_longer(names_to = "Measurement",
               cols = Temperature_1:Temperature_at_Time_of_Detection,
               values_to = "Temperature", names_repair = "unique")
# remove NAs

temperature_long <- temperature_long %>%
  na.omit()
```

## Temp Figure

Histogram per station & season

```{r temperature histo}

temperature.hist <- ggplot(temperature_long, aes(x= Temperature)) +
  geom_histogram() +facet_grid(Season~Station) + theme_bw()

temperature.hist
```

And do a mean and se with points overlain

```{r temperature mean se figure}

temp.mean.se <- temperature_long %>%
  na.omit() %>%
  group_by(Station, Season) %>%
  summarise(mean = mean(Temperature),
            sd = sd(Temperature),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

temp.fig <- ggplot(temp.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) + geom_jitter(data = temperature_long, aes( x = Station, y = Temperature), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.75) + theme_bw() +
  scale_fill_manual(values = c("lightgreen","lightgreen")) + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5) + geom_point(pch = 21, size = 3) + facet_wrap(~Season) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("Measured Temperature") + ggtitle("Temps by Field Season & Station") + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(legend.position = "none") + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

temp.fig

#Interesting, seems 2025 was def colder at YRS, and a little bit overall as well

#ggsave(temp.fig, file="temp-mean-se.png", width = 6,  height = 4)

rm(temp.fig)
```

```{r clean after temp}
rm(temp.mean.se, temperature.hist)
```

# Humidity

## Hum Set-Up

Get temperature data reformatted. We took these readings every 30 min and also every time an individual was found. First step will be looking at variation between stations. 

```{r reformat humidity}

humidity_long <- enviro2 %>%
  select(Survey_Start_Date_Time, Station, Season, Parent_Trail_LongName, 
         Humidity_1:Humidity_7, Humidity_Final, Humidity_at_Time_of_Detection) %>%
  pivot_longer(names_to = "Measurement",
               cols = Humidity_1:Humidity_at_Time_of_Detection,
               values_to = "Humidity", names_repair = "unique")
# remove NAs

humidity_long <- humidity_long %>%
  na.omit()

# Sanity check to make sure pivoted temp and humidity DFs are same length
length(humidity_long$Survey_Start_Date_Time)
length(temperature_long$Survey_Start_Date_Time) 
```

## Hum Figure

Histogram per station & season

```{r humidity histo}

humidity.hist <- ggplot(humidity_long, aes(x= Humidity)) +
  geom_histogram() +facet_grid(Season~Station) + theme_bw()
humidity.hist
```

And do a mean and se with points overlain

```{r humidity mean se figure}

hum.mean.se <- humidity_long %>%
  na.omit() %>%
  group_by(Station, Season) %>%
  summarise(mean = mean(Humidity),
            sd = sd(Humidity),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

hum.fig <- ggplot(hum.mean.se, aes(x=Station, y=mean, fill = Station, group = Station)) + geom_jitter(data = humidity_long, aes( x = Station, y = Humidity), width = 0.1, size = 1.5, alpha = 0.2) + coord_fixed(ratio = 0.75) + theme_bw() +
  scale_fill_manual(values = c("lightgreen","lightgreen")) + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5) + geom_point(pch = 21, size = 3) + facet_wrap(~Season) + xlab("") + scale_color_discrete(labels = "") +
  theme(legend.title=element_blank()) + ylab("% Relative Humidity") + ggtitle("Humidity by Field Season & Station") + scale_x_discrete(labels=c("Unhunted", "Hunted")) + theme(legend.position = "none") + theme(axis.text = element_text(color = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

hum.fig

#Huge amount of variation in humidity, overall more humid at YRS but 
# definitely more humid in 2025, likely from all the huge rain events 

#ggsave(hum.fig, file="humidty-mean-se.png", width = 6,  height = 4)

#rm(hum.fig)
```

```{r clean env after hum}
rm(hum.fig, hum.mean.se, humidity.hist)
```

# Handling Humidity and Temp

We want to summarize humidity and temp in a way that allows us to incorporate it into our model. As of 2/23, I'm thinking the best way to do this is to take means per survey date.

Let's take the mean vals per date

```{r temp hum means} 

temp_date_means <- temperature_long %>%
  na.omit() %>%
  group_by(Survey_Start_Date_Time, Station) %>%
  reframe(mean_temp_F = mean(Temperature))

hum_date_means <- humidity_long %>%
  na.omit() %>%
  group_by(Survey_Start_Date_Time, Station) %>%
  reframe(mean_humidity_percent = mean(Humidity))

temp_hum_df <- left_join(temp_date_means, hum_date_means)

temp_hum_df$Start_Date <- as.Date(temp_hum_df$Survey_Start_Date_Time)

temp_hum_df <- temp_hum_df %>%
  select(-Survey_Start_Date_Time)

# clean
rm(temperature_long, humidity_long, temp_date_means, hum_date_means, understory_long)
```

# Density estimates

Note I think it's a good idea to follow the methods in this paper: Hierarchical distance sampling to estimate population sizes of common lizards across a desert ecoregion (in Zotero shared library)

## Binning

First, create distance bins. Also look at some things.

```{r review data more} 

# pull out year to make a way to distinguish between seasons

snakes.sur$Year <- factor(format(snakes.sur$Start_Date, "%Y"))

snakes.sur %>%
  group_by(Year, Station) %>%
    reframe(N_surveys = length(unique(Start_Date))) #Effort comparisons between stations

enviro %>%
  group_by(Season, Station) %>%
    reframe(Accum_hrs = sum(Effort_hrs))

# Deciding on distance bins in distance sampling involves selecting cutpoints that adequately capture the shape of the detection function, typically ensuring more bins near the line (where detection probability changes rapidly) and fewer further away. The goal is to minimize bias from data rounding while ensuring enough observations per bin for accurate model fitting. 

# Smaller bins closer to zero

# Truncate your data to remove the farthest detections - WE CANNOT DO THIS DUE TO THE BOA BIAS THAT OWEN POINTED OUT

# First, look at a histogram of distance from trail to get an idea of detection distribution

hist(snakes.sur$Distance_From_Trail_m, breaks = seq(0, 12, by = 0.5))

```

Proceed with specifying bins.

The chunk below divides observations into smaller bins closer to the trail and larger bins farther away. 0, 0.25, 0.5, 1, 1.5, 2, 2.5, 3.5, 5, 8.

```{r divide up bins, eval=FALSE} 

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 0.25, "0_0.25", NA))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 0.25 & Distance_From_Trail_m <= 0.5, "0.25_0.5", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 0.5 & Distance_From_Trail_m <= 1, "0.5_1", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 1.5, "1_1.5", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1.5 & Distance_From_Trail_m <= 2, "1.5_2", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 2.5, "2_2.5", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2.5 & Distance_From_Trail_m <=3.5, "2.5_3.5", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 3.5 & Distance_From_Trail_m <=5, "3.5_5", Distance_bin_m))

#snakes.sur <- snakes.sur %>%
 # mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 5, "5_8", Distance_bin_m))

# join temp and humidity means to the snakes.sur data
#snakes.sur <- left_join(snakes.sur, temp_hum_df)
```

This verions of binning (chunk below) divides chunks up into 1 m intervals until 5m from trail, which is grouped into 1 bin of distances 5-8m from trail (and I threw the 11.5m away obs in that group too).

```{r divide up bins broader}

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 1, "0_1", NA))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 2, "1_2", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 3, "2_3", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 3 & Distance_From_Trail_m <= 4, "3_4", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 4 & Distance_From_Trail_m <= 5, "4_5", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 5, "5_8", Distance_bin_m))

```

What I'm reading indicates with sparse data, larger bins are better. I'm trying a version here with much larger bins.

## 3 Diet types

Mixed, primarily mammal consumers, non-mammal consumers

```{r format for unmarked 3 diet types}

# isolate surveys for which no snakes were found
zeros.3.diets <- snakes.sur %>%
  filter(is.na(Snake_ID)) %>%
  select(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m)

# get them into the format we want (tally doesn't actually matter here)
z.tally.no.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

# for the nights when 0 snakes were found, make a dataframe of instances that specifies zero found for each diet type category

# no mammals
z.tally.no.mammals.3.diets$Diet_Type_3categories <- "no_mammals"

# mostly mammals
z.tally.mostly.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

z.tally.mostly.mammals.3.diets$Diet_Type_3categories <- "mostly_mammals"

# mixed

# mixed
z.tally.mixed.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

z.tally.mixed.3.diets$Diet_Type_3categories <- "mixed"

# stack them together
z.tallies.3 <- rbind(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

#clean
rm(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

# also, figure out why the zero detections on CHIO-03 6-11-2024 are entered twice, as are 6-23 MAPO03, 6-23 MURC-01, and 6-23 PUEN-01 (all 2024) - ignoring for now bc it's an error I'm pretty sure

# remove the n column (meaningless)
z.tallies.3 <- z.tallies.3 %>%
  select(!n)

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

length(surveys.no.zeros$Snake_ID) # SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally.3 <- surveys.no.zeros %>%
  group_by(Start_Date, Year, Station, Diet_Type_3categories, Segment, Seg_dist_m, Distance_bin_m) %>% tally()

# pivot wider that wider
density_est_df.3 <- survey.tally.3 %>% 
  pivot_wider(names_from = Distance_bin_m,
              values_from = n, 
              values_fill = 0)

# now, we can add the surveys and the segments with zero detections back

density_est_df.3 <- full_join(z.tallies.3, density_est_df.3)

# make NAs zeros - except the temp and humidty NAs

density_est_df.3 <- density_est_df.3 %>%mutate_all(~replace(., is.na(.), 0))

# clean
rm(survey.tally.3, surveys.no.zeros, z.tallies.3, zeros.3.diets)
```

```{r make umf three diet types}

density_est_df.3$Year <- as.factor(density_est_df.3$Year)
density_est_df.3$Diet_Type_3categories <- as.factor(density_est_df.3$Diet_Type_3categories)
density_est_df.3$Seg_dist_m <- as.numeric(density_est_df.3$Seg_dist_m)
density_est_df.3$Segment <- as.factor(density_est_df.3$Segment)
density_est_df.3$Station <- as.factor(density_est_df.3$Station)

# separate out the covariates
covs.3 <- density_est_df.3[,c("Station", "Year", "Diet_Type_3categories", "Segment", "Seg_dist_m")]

umf.3 <- unmarkedFrameDS(y=as.matrix(density_est_df.3[,7:ncol(density_est_df.3)]), siteCovs=data.frame(covs.3), dist.breaks=c(0, 1, 2, 3, 4, 5, 8), unitsIn="m", survey="line", tlength=covs.3$Seg_dist_m)

hist(umf.3, xlab="distance (m)", main="", cex.lab=0.8, cex.axis=0.8)

```

### GOF function

Goodness of fit tests

```{r gof function}

# using this resource: https://www.montana.edu/screel/teaching/bioe-440r-521/bioe-440-distance-sampling.html

# Goodness of fit test: good fit if results of these tests show p > 0.05. First, fit a model: use your best-supported model (Lowest AIC). Then check GOF for that model.

# Function returning three fit-statistics.

fitstats <- function(fm) {

  observed <- getY(fm@data)

  expected <- fitted(fm)

  resids <- residuals(fm)

  sse <- sum(resids^2)

  chisq <- sum((observed - expected)^2 / expected)

  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2)

  out <- c(SSE=sse, Chisq=chisq, freemanTukey=freeTuke)

  return(out)

}

```

## Model 3 Diet Types

```{r mods three diet types no env covs}

null.3 <- distsamp(~1 ~ 1, umf.3, unitsOut = "kmsq")
summary(null.3)

Diet_det <- distsamp(~ Diet_Type_3categories ~ 1, umf.3, unitsOut = "kmsq")
summary(Diet_det)# diet type as detection covariate improves model

Year.3 <- distsamp(~ Diet_Type_3categories ~ Year, umf.3, unitsOut = "kmsq")
summary(Year.3)
# doesn't significantly improve fit (< 2 AIC)

Station.3 <- distsamp(~Diet_Type_3categories ~ Station, umf.3, unitsOut = "kmsq")
summary(Station.3)
# station makes things worse

StationXGuild.3 <- distsamp(~Diet_Type_3categories ~ Station/Diet_Type_3categories, umf.3, unitsOut = "kmsq")
summary(StationXGuild.3)
# Make a table to compare models

# put mods in a list
mod.set.3diets <- list(null.3 = null.3, Year.3 = Year.3, StationXGuild.3 = StationXGuild.3, Station.3 = Station.3, Diet_det = Diet_det)

# make table
aic_table.3 <- aictab(cand.set = mod.set.3diets, sort = TRUE)

aic_table.3

```

### GOF 3 diet types

```{r goodness of fit three guilds}
#pb_3cats <- parboot(StationXGuild.3, fitstats, nsim=1000)
#pb_3cats
# fine
```

## Predicted vals 3 diet types

```{r predicted values three guilds}

new.data3 <- density_est_df.3 %>%
  ungroup() %>%
  dplyr::select(Station, Diet_Type_3categories) %>%
  distinct()

#Predict

predictions3 <- predict(StationXGuild.3, type="state", newdata=new.data3, appendData=TRUE)

```

```{r plot three guilds, message=FALSE, warning=FALSE}
predictions3$Diet_Type_3categories <- factor(
  predictions3$Diet_Type_3categories,
  levels = c("no_mammals", "mixed", "mostly_mammals", 
  labels = c("No Mammals", "Mixed", "Mostly Mammals"))
) #make sure that the order goes nomamm > mixed > mostly mamm

p_3cats<-ggplot(predictions3, aes(x=Station, y=Predicted, fill = Diet_Type_3categories, group = Station)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = .2) + geom_point(pch = 21, size = 4) + theme_bw() + scale_fill_manual(values = c("goldenrod1", "green4", "black")) + xlab("") + scale_color_discrete(labels = "") +  scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake Density Estimates by Dietary Guild") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(text = element_text(family = "Tahoma")) + theme(axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12, margin = margin(b = 10))) + scale_fill_manual(
  name = "Diet Type",
  values = c("goldenrod1", "green4", "black"),
  labels = c("No mammals", "Mixed", "Mostly mammals")
) + facet_wrap(~Diet_Type_3categories, scales = "free", labeller = labeller(Diet_Type_3categories = c(
    no_mammals = "No Mammals",
    mixed = "Mixed",
    mostly_mammals = "Mostly Mammals"
  ))
)

p_3cats

#ggsave(p_3cats, file="snk_dens_3diet.png", width = 8, height = 5)


```

Appears to be elevated numbers in hunted forest in the guild that is mostly mammal consumers.

## Family level

The two most detected families are Boidae and Colubridae. Limitting the analysis to these 2 families allows the models to run without error with Family as a covariate. Including all families detected creates model issues with uncertainty because the other families were only detected 1 or 2 times. 

```{r format for unmarked based on fams}

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m <= 1, "0_1", NA))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 1 & Distance_From_Trail_m <= 2, "2", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 2 & Distance_From_Trail_m <= 4, "2_4", Distance_bin_m))

snakes.sur <- snakes.sur %>%
  mutate(Distance_bin_m = ifelse(Distance_From_Trail_m > 4, "4_8", Distance_bin_m))

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

# SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally <- surveys.no.zeros %>%
  filter(Family != "Viperidae") %>%
filter(Family != "Elapidae") %>%
  group_by(Start_Date, Year, Station, Family, Segment, Seg_dist_m, Distance_bin_m) %>% tally()

# generate all possible combinations based on family to get zeros

df_long <- survey.tally %>%
  as_tibble() %>%
  complete(Family, Distance_bin_m, nesting(Start_Date, Year, Station, Segment, Seg_dist_m), fill = list(n = 0))

# also, which families eat what?
snakes.sur %>%
  group_by(Family) %>%
  reframe(unique(Diet_Type)) # colubridae is only possibility for mixed in our data?

# pivot wider that wider
density_fam_df <- df_long %>% 
  pivot_wider(names_from = Distance_bin_m,
              values_from = n, 
              values_fill = 0)
```

## Model family level density

```{r model at fam level}
# separate out the covariates

# model not running bc of sparse data at fam level - removed elapidae, which only had 1 detection, and Viperidae which only had 2

covs.fam <- density_fam_df[,c("Station", "Year", "Family", "Segment", "Seg_dist_m")]

umf.fam <- unmarkedFrameDS(y=as.matrix(density_fam_df[,7:ncol(density_fam_df)]), siteCovs=data.frame(covs.fam), dist.breaks=c( 0,1,2,4,8), unitsIn="m", survey="line", tlength=covs.fam$Seg_dist_m)

# run models
# null
null.fam <- distsamp(~1 ~ 1, umf.fam, unitsOut = "kmsq")
summary(null.fam)

# family as detection covariate, no density covariate
fam.null  <- distsamp(~Family ~ 1, umf.fam, unitsOut = "kmsq")
summary(fam.null)

# add year as density covariate
year <- distsamp(~ Family ~ Year, umf.fam, unitsOut = "kmsq")
summary(year)

# add station as density covariate
station <- distsamp(~Family ~ Station, umf.fam, unitsOut = "kmsq")
summary(station)

# add station and station x family interaction term 

station_stationXfam <- distsamp(~Family~ Station/Family, umf.fam, unitsOut = "kmsq")

summary(station_stationXfam)
# best mod by far

# put mods in a list
mod.set.family <- list(null.fam = null.fam, fam.null = fam.null, year = year, station = station, station_stationXfam = station_stationXfam)

# make table
aic_table.fam <- aictab(cand.set = mod.set.family, sort = TRUE)

aic_table.fam
```

## Predicted vals fam

```{r predicted values three guilds}

new.data.fam <- density_fam_df %>%
  ungroup() %>%
  dplyr::select(Station, Family) %>%
  distinct()
  

#Predict

predictions.fam <- predict(station_stationXfam, type="state", newdata=new.data.fam, appendData=TRUE)


```


```{r plot fam, message=FALSE, warning=FALSE}

p_3cats_fam<-ggplot(predictions.fam, aes(x=Station, y=Predicted, fill = Family, group = Station)) +
  geom_errorbar(aes(ymin = Predicted - SE, ymax = Predicted + SE), width = .2) + geom_point(pch = 21, size = 3) + theme_bw() + scale_fill_manual(values = c("purple2","goldenrod1", "black")) + xlab("") + scale_color_discrete(labels = "") +  scale_y_continuous(name = expression(bold(paste("Snakes /  ", km^bold("2"))))) + ggtitle("Snake Density Estimates by Family") +
  scale_x_discrete(labels = c("Unhunted", "Hunted")) + theme(text = element_text(family = "Tahoma"), axis.text = element_text(color = "black", size = 12), axis.title = element_text(family = "Tahoma"), plot.title = element_text(family = "Tahoma", 
      hjust = 0.5,face = "bold", size = 12)) 

p_3cats_fam

#ggsave(p_3cats_fam, file="snk_dens_3diet_fam.png", width = 8, height = 5)
```

Appears to be elevated numbers in hunted forest in the family that is mostly mammal consumers.


# Community composition

## Pie Charts

Below I mess around with some basic community composition visualizations.

```{r species occurrences}

# just get an idea of n detections from all survey types per station, including the instances with no ID
raw_count <- snakes.master %>%
  group_by(Station) %>%
  reframe(n = n())

# total at YRS is 111, at TBS is 108 (great, super comparable numbers when pooled)

# subfamily level may help
# create this column and populate it:
snakes.master$Subfamily <- NA
snakes.master <- snakes.master %>%
  dplyr::mutate(Subfamily = ifelse(Genus == "Siphlophis" | Genus == "Imantodes" | Genus == "Dipsas" | Genus == "Drepanoides" | Genus == "Oxyrhopus" | Genus == "Leptodeira" | Genus == "Hydrops" | Genus == "Helicops" | Genus == "Atractus" | Genus == "Erythrolamprus" | Genus == "Taeniophallus" | Genus == "Pseudoboa", "Dipsadinae", 
                            ifelse(Genus == "Epicrates" | Genus == "Corallus" | Genus == "Eunectes", "Boinae",
                                   ifelse(Genus == "Chironius" | Genus == "Drymobius" | Genus == "Tantilla" | Genus == "Drymarchon" | Genus == "Oxybelis" | Genus == "Spilotes" | Genus == "Dendrophidion" | Genus =="Leptophis", "Colubrinae",
                                          ifelse(Genus == "Bothrops" | Genus == "Bothrocophias", "Crotalinae",
                                                 ifelse(Genus == "Xenoxybelis", "Xenodontinae",
                                                        ifelse(Genus == "Micrurus", "Elapinae", Subfamily)))))))

#Siphlophis = Dipsadinae
#Imantodes = Dipsadinae 
#Epicrates = Boinae
#Dipsas = Dipsadinae
#Drepanoides = Dipsadinae  
#Chironius = Colubrinae
#Corallus = Boinae
#Oxyrhopus = Dipsadinae
#Leptodeira = Dipsadinae
#Hydrops = Dipsadinae
#Bothrops = Crotalinae
#Xenoxybelis = Xenodontinae
#Drymobius = Colubrinae
#Micrurus = Elapinae
#Helicops = Dipsadinae
#Eunectes = Boinae   
#Atractus = Dipsadinae
#Tantilla = Colubrinae
#Drymarchon = Colubrinae
#Erythrolamprus = Dipsadinae
#Taeniophallus = Dipsadinae
#Bothrocophias = Crotalinae
#Spilotes = Colubrinae
#Pseudoboa = Dipsadinae
#Oxybelis = Colubrinae
#Dendrophidion = Colubrinae
#Leptophis = Colubrinae
#Bothrocophias = Crotalinae

# make an edge list of n times an individual was seen at each station for each subfamily group
sub.fam.edgelist <- snakes.master %>%
  group_by(Station, Subfamily) %>%
  tally() %>%
  na.omit() 

# piechart
ggplot(sub.fam.edgelist, aes(x="", y= n, fill=Subfamily)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) + # remove background, grid, numeric labels
  scale_fill_brewer(palette="Set1") # remove background, grid, numeric labels

# only the survey snake data, at subfamily level
surveys <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Subfamily) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys, aes(x="", y= n, fill=Subfamily)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) + 
  scale_fill_brewer(palette="Set1") 

# Genus level, all detection methods:
all.genus <- snakes.master %>%
  group_by(Station, Genus) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(all.genus, aes(x="", y= n, fill=Genus)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station)

# Genus level, surveys only:
surveys.genus <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Genus) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys.genus, aes(x="", y= n, fill=Genus)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) 

# family

surveys.family <- snakes.master %>%
  filter(Method_of_Detection == "Found on survey") %>%
  group_by(Station, Family) %>%
  tally() %>%
  na.omit() 

# pie chart
ggplot(surveys.family, aes(x="", y= n, fill=Family)) +
  geom_bar(stat="identity", width=0.5, color="black") +
  coord_polar("y", start=0) +
  theme_void() + facet_wrap(~Station) 
```


## All records community composition

### Jaccard PERMANOVA 

```{r jaccard permanova}

# going to re-run this since some snakes were missing previously
snakes.master$ID <- paste(snakes.master$Station, snakes.master$Date, snakes.master$Snake_ID, sep = "//")

edgelist.full <- snakes.master %>%
  filter(Genus != "unknown") %>%
  group_by(Station, Species, Date, ID) %>%
  tally() %>%
  na.omit()

# now, let's use dplyr again to transpose (flip) your sp.abun dataframe so that now, you have the species as column names and the number of times they were found at each trail and date will be the data reported (expect lots of 0s and 1s)

snake.occ.mat <- edgelist.full %>% 
  pivot_wider(names_from = Species, #makes the column into the rownames
              values_from = n, 
              values_fill = 0) %>% #will fill spaces with no snakes as 0
  column_to_rownames(var = "ID") #will make this new combo the unique identifier as rownames

community <- snake.occ.mat[, c(3:42)] #made matrix from all data columns with n snake data

predictors <- snake.occ.mat[, c(1,2)] #separating non-snake stuff (station and zone)

j.dis.mat <- vegdist(community, method = "jaccard") #jaccard only accounts for presence absence 

pnova.j <- adonis2(j.dis.mat ~ predictors$Station) #adonis looks at specific differences between groups, our group factor is station

# look at results
print(pnova.j) # marginally sig/nonsig
```

### Jaccard NMDS

```{r jaccard nmds visual}

```

## Survey records community composition

### Jaccard PERMANOVA

### Jaccard NMDS

### Horn-Morisita PERMANOVA

### Horn-Morisita NMDS

# Diversity

```{r diversity & hill numbers}

# Diversity

# need data in a new species x site matrix
unique(snakes.sur$Species)

temp <- snakes.sur %>%
  group_by(Station, Species) %>% tally()

temp <- temp %>%
  na.omit()

# Need to modify the matrix so we have an n column, etc.


# get the new format, species as row names, station as columns
div.mat <- temp %>% 
  pivot_wider(names_from = Station, 
              values_from = n, 
              values_fill = 0) %>%
  column_to_rownames(var = "Species")

# now we use div.mat to get the diversity and 95% CIs
# new function uses iNEXT package (confusingly also the name of the function)
species.diversity = iNEXT(div.mat, q = c(0,1,2), datatype = "abundance", nboot = 200)

# extract values we need
species.diversity.output <- species.diversity[["iNextEst"]][["coverage_based"]]

div.vals <- species.diversity.output %>% 
  filter(Assemblage == "TBS" & m == 100 | Assemblage == "YRS" & m == 100)

# plot diversity
diversity.plot <- ggplot(data = div.vals, aes(x = Assemblage, y = qD, fill = as.factor(Order.q), shape = as.factor(Order.q))) + theme_light() + geom_pointrange(aes(ymin = qD.LCL, ymax = qD.UCL), position = position_dodge2(width = 0.35), color = "black", size = 0.6) + xlab(NULL) + ylab("Effective Number of Species") + scale_x_discrete(labels = c("Unhunted", "Hunted")) + scale_fill_manual(values = c("red", "goldenrod", "cornflowerblue"), labels = c("q = 0\n(richness)", "q = 1\n(exp. Shannon)", "q = 2\n(inv. Simpson)")) + coord_fixed(ratio = 0.1) + scale_shape_manual(values = c(21, 22, 23)) +
  theme(axis.title.y = element_text(size = 11, colour = "black", family = "Tahoma", 
                                    margin = margin(r = 10))) +
  theme(axis.title.x = element_text(size = 11, colour = "black", family = "Tahoma")) +
  theme(axis.text = element_text(size = 11, colour = "black", family = "Tahoma")) +
  theme(strip.text.x = element_text(size = 11, colour = "black", family = "Tahoma")) + 
  theme(legend.title = element_blank()) + theme(legend.text = element_text(size = 10)) +
  theme(plot.title = element_text(size = 11, colour = "black", family = "Tahoma", 
                                  hjust = .5, face = "bold")) + theme(legend.position = "bottom") +
  guides(fill = guide_legend(override.aes = list(shape = c(21, 22, 23)))) + guides(shape = "none") + ggtitle(str_wrap("Station-Level Diversity Measured with Hill Numbers (q = 0, 1, 2)", width=40))

diversity.plot

#ggsave("snk_diversity_plot.png", plot = diversity.plot, width = 6, height = 6.5)

```

Species accumulation curves

```{r accum curves}

ggiNEXT(species.diversity, type=1, facet.var="Order.q", grey=TRUE) 

```

Right so no difference there.

# Raw detection rates over distance

```{r raw detection rates}

# isolate surveys for which no snakes were found
zeros.3.diets <- snakes.sur %>%
  filter(is.na(Snake_ID)) %>%
  select(Start_Date, Station, Diet_Type_3categories, Segment, Seg_dist_m)

# get them into the format we want (tally doesn't actually matter here)
z.tally.no.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

# for the nights when 0 snakes were found, make a dataframe of instances that specifies zero found for each diet type category

# no mammals
z.tally.no.mammals.3.diets$Diet_Type_3categories <- "no_mammals"

# mostly mammals
z.tally.mostly.mammals.3.diets <- zeros.3.diets %>%
  group_by(Start_Date, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

z.tally.mostly.mammals.3.diets$Diet_Type_3categories <- "mostly_mammals"

# mixed

# mixed
z.tally.mixed.3.diets <- zeros.3.diets %>%
  group_by(Start_Date,Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

z.tally.mixed.3.diets$Diet_Type_3categories <- "mixed"

# stack them together
z.tallies.3 <- rbind(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

#clean
rm(z.tally.no.mammals.3.diets, z.tally.mostly.mammals.3.diets, z.tally.mixed.3.diets)

# also, figure out why the zero detections on CHIO-03 6-11-2024 are entered twice, as are 6-23 MAPO03, 6-23 MURC-01, and 6-23 PUEN-01 (all 2024) - ignoring for now bc it's an error I'm pretty sure

# remove the n column (meaningless)
z.tallies.3 <- z.tallies.3 %>%
  select(!n)

# also isolate surveys that DID have snakes detected
surveys.no.zeros <- snakes.sur %>%
  filter(!is.na(Snake_ID))

length(surveys.no.zeros$Snake_ID) # SK04 and SK05 are duplicated
# remove the duplicate entries

surveys.no.zeros <- surveys.no.zeros %>%
  filter(!duplicated(Snake_ID))

# tally counts - df without zeros
survey.tally.3 <- surveys.no.zeros %>%
  group_by(Start_Date, Station, Diet_Type_3categories, Segment, Seg_dist_m) %>% tally()

# now, we can add the surveys and the segments with zero detections back

raw_detection_df <- full_join(z.tallies.3, survey.tally.3)

# make NAs zeros
raw_detection_df <- raw_detection_df %>%
  mutate(n = ifelse(is.na(n), 0, n))

# I did this fast and it should be checked

detection_rates <- raw_detection_df %>%
  group_by(Station, Diet_Type_3categories, Start_Date) %>%
  summarize(meters_surveyed = sum(Seg_dist_m),
            individuals_found = sum(n))

sum(detection_rates$individuals_found)
detection_rates$n_per_500m <- (detection_rates$individuals_found/detection_rates$meters_surveyed)*500

sum(detection_rates$individuals_found)
detection_rates$n_per_500m <- (detection_rates$individuals_found/detection_rates$meters_surveyed)*500

detection_rates$n_per_1000m <- (detection_rates$individuals_found/detection_rates$meters_surveyed)*10000
# summarize and visualize
detections.mean.se <- detection_rates %>%
  group_by(Station, Diet_Type_3categories) %>%
  summarise(mean = mean(n_per_1000m),
            sd = sd(n_per_1000m),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

ggplot(detections.mean.se, aes(x=Station, y=mean, fill = Diet_Type_3categories)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21, size = 3) + theme_bw() + scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "black")) + xlab("") + scale_color_discrete(labels = "") + ylab("Detections/1km") + ggtitle("Snake Detections per 1km by Diet Type") + facet_wrap(~Diet_Type_3categories, scales = "free")

# need to also report raw overall detections regardless of diet type per station 
```
