---
title: "ECHerp"
author: "Vogin and Lamperty"
date: "2025-10-07"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Overview

This study looks at how hunting affects herpetofaunal communities, specifically those of frog and lizards. Overhunting can lead to defaunation, the decrease in large mammal populations. This has many ecological consequences, including cascading trophic effects. As large mammal populations decrease, small mammal population tend to increase due to decreased predation and competition. Mammal-eating snake populations may increase as a result of this from the increased prey base. As juveniles, these snakes feed on other prey, including frogs and lizards, which means potentially higher predation pressure where snake populations are higher.

To test this, we conducted visual detection surveys at a hunted and an unhunted site in order to determine if the community compositions of frogs and lizards differ. The hunted site is located at the Yasuni Research Station (YRS). YRS has experienced decades of hunting from the Waorani, the indigienous people of the Ecuadorian Amazon, leading to significant defaunation. The unhunted site is located at the Tiputini Biodiversity Station (TBS). TBS is located in a remote and, therefore, has very little human disturbance and is faunally intact.

Data was collected during an eight week period, from June- July of 2025. The time was split evenly between the two sites, rotating between the sites for a total of two visits at each site. Multiple visual detection surveys, including days and nights, were conducted each week. The surveys involved slowly walking the trail systems, scanning for frogs and lizards. When an individual was found, data was collected, including: species, location, time, and weather.

# Package Read In
```{r read in packages}
library(tidyverse) # data manip
library(lubridate) # date and time manipulation
library(vegan) # community composition/dissimilarity analyses
library(iNEXT) # diversity analyses
library(unmarked) # density estimates
```

# Data Read In and Clean
```{r read in data}
amph_lz=read.csv("Data/Field_Season_2_datasheets(amphibians_and_lizards).csv")
env=read.csv("Data/Field_Season_2_datasheets(Environment_Sheet).csv")
seg_dist=read.csv("Data/Field_Season_2_datasheets(Segment_Distances).csv")


# Read in 2024 data
amph_2024 <- read.csv("2024_amphibians.csv")
liz_2024 <- read.csv("2024_lizards.csv")
```


```{r clean 2024 data}
# Start with amphibians
amph_2024 <- amph_2024[-2,]
colnames(amph_2024) <- amph_2024[1,]
amph_2024 <- amph_2024[-1,]

amph_2024 <- amph_2024 %>%
  rename(Time_Found = Time_Found_)
amph_2024 <- amph_2024 %>%
  mutate(Survey_Type = "Night", 
         X = NA, 
         X.1 = NA, 
         Size = NA,
         Diet_Type = NA,
         Primary_Diet_Component = NA)


# Now for lizards
liz_2024 <- liz_2024[-2,]
colnames(liz_2024) <- liz_2024[1,]
liz_2024 <- liz_2024[-1,]

liz_2024 <- liz_2024 %>%
  rename(Time_Found = Time_Found_)
liz_2024 <- liz_2024 %>%
  mutate(Survey_Type = "Night", 
         X = NA, 
         X.1 = NA, 
         Size = NA,
         Diet_Type = NA,
         Primary_Diet_Component = NA)

```

```{r organize and clean species data, warning=FALSE}

amph_lz=amph_lz %>%
  separate(Species,into=c("Genus","Species")) #turn Species column into Genus and Species columns

# replace unknown family with NA
amph_lz <- amph_lz %>%
  mutate(Family = ifelse(Family == "weird" | Family == "mystery" | Family == "weird ground frog in valley, looked like big head x treefrog" | Family == "unknown", NA, Family))

# replace blank Genus and Unknown Genus entries with NA
amph_lz <- amph_lz %>%
  mutate(Genus = ifelse(Genus == "" | Genus == "Unknown", NA, Genus))

amph_lz$Species= na_if(amph_lz$Species, "spp") #replace "spp" with NA

# Check to make sure  column names are the same
colnames(amph_2024)
colnames(amph_lz)

colnames(liz_2024)
colnames(amph_lz)

# MERGE ALL DATA INTO AMPH_LZ

amph_lz <- rbind(amph_lz, amph_2024, liz_2024)

rm(amph_2024)
rm(liz_2024)

rm_trail_spaces <- function(column) {
  clean_column <- trimws(column)
  return(clean_column)
}

# Get rid of duplicate stuff

amph_lz$Gross_ID <- rm_trail_spaces(amph_lz$Gross_ID)
amph_lz$Genus <- rm_trail_spaces(amph_lz$Genus)
amph_lz$Survey_Type <- rm_trail_spaces(amph_lz$Survey_Type)
amph_lz$Station <- rm_trail_spaces(amph_lz$Station)
amph_lz$Weather <- rm_trail_spaces(amph_lz$Weather)
amph_lz$Animal_ID <- rm_trail_spaces(amph_lz$Animal_ID)

## WHAT'S GOING ON WITH FG594?

```

```{r check for typos}
colnames(amph_lz)

amph_lz$Start_Date_Start_Time=paste(amph_lz$Date,amph_lz$Survey_Start_Time,sep=" ")
amph_lz$End_Date_End_Time=paste(amph_lz$Date,amph_lz$Survey_End_Time,sep=" ")

amph_lz$Start_Date_Start_Time=mdy_hm(amph_lz$Start_Date_Start_Time)
amph_lz$End_Date_End_Time=mdy_hm(amph_lz$End_Date_End_Time)

#### Date ####

amph_lz$Date=mdy(amph_lz$Date)

#### Survey_Type ####
amph_lz$Survey_Type=trimws(amph_lz$Survey_Type)

#### Survey_hr ####
amph_lz$Survey_hr=difftime(amph_lz$End_Date_End_Time,amph_lz$Start_Date_Start_Time,units="hours")

#### Gross_ID () ####
amph_lz$Gross_ID=trimws(amph_lz$Gross_ID)

#### Genus ####
amph_lz %>%
  distinct(Genus) %>% arrange(Genus)

#### Time_Found ####
amph_lz$Time_Found=hm(amph_lz$Time_Found)

```

We need to make a dataframe that contains the dates that each segment was surveyed, preferably also merged with the length of each of those segments.

```{r segment master list}

# env should have every segment ever surveyed ... and each time it was surveyed
segs_survey_record <- env %>%
  select(c(1,2,3,5)) %>%
  separate_rows(Order, sep = ",")

# and fix the names in segs_expanded to match and to be easy to understand
segs_survey_record <- segs_survey_record %>%
  rename(Segment = Order)

# there's just one where there is a hyphen separating two segments that need to be in separate rows
#segs_survey_record <- segs_survey_record %>%
 # mutate(Segment = ifelse(Segment == "NUMA-02-CHIO-03", 
  #                        "NUMA-02, CHIO-03", Segment)) %>%
  #separate_rows(Segment, sep = ",")

# remove all the spaces
segs_survey_record <- segs_survey_record %>%
  mutate(Segment = trimws(Segment))

# fix one case where there is just a space between two segments that need to be in separate rows
segs_survey_record <- segs_survey_record %>%
  separate_rows(Segment, sep = " ")

# get ready to add the meters per segment. Need names to match
seg_dist <- seg_dist %>%
  rename(Segment = SEGMENT) %>%
  rename(Parent_Trail = PARENT.TRAIL) %>%
  rename(Seg_dist_m = DISTANCE_.M.) %>%
  select(!Parent_Trail)

# first, do we have distances for all segments named in the segs_survey_record df?

# one typo - 	
#CEIBA-01 should be CEIB-01
segs_survey_record <- segs_survey_record %>%
  mutate(Segment = ifelse(Segment == "CEIBA-01", "CEIB-01", Segment))

segs_survey_record_with_distances_added <- left_join(segs_survey_record, seg_dist)

```

```{r summarize findings some}

# Jacob: I think this could be a good thing to make into a table (use R or excel) to report in your results. Re-label "YRS" and "TBS" in your table as "Hunted" and "Unhunted" though
amph_lz %>% group_by(Station, Gross_ID) %>% 
  summarize(total_found = n_distinct(Animal_ID),
            species_found = n_distinct(Species, na.rm = TRUE)) 


```

```{r join segments and survey info to df}

# reduce down to the columns we're interested in right now
amph_lz_segs <- amph_lz %>%
  select(Date, Station, Segment, Animal_ID, Distance_From_Trail_m, Gross_ID, Family, Genus, Species, Common_Name, Survey_Type, Habitat_Type, Height_of_Individual)

# fix date name to match
amph_lz_segs <- amph_lz_segs %>%
  rename(Start_Date = Date)

# make seg master date a date object
segs_survey_record_with_distances_added$Start_Date <- mdy(segs_survey_record_with_distances_added$Date)

segs_survey_record_with_distances_added <- segs_survey_record_with_distances_added %>%
  select(-Date)

# Find segs in segments master that are not in amph_lz
segs_in_master_not_surveys <- anti_join(segs_survey_record_with_distances_added, amph_lz_segs, by = "Segment")
# ok so 16 times a seg was surveyed without anything being found. Seems beleivable.
unique(segs_in_master_not_surveys$Segment)
# Jacob please double check this makes sense

# vice versa
# should be 0
segs_in_surveys_not_master <- anti_join(amph_lz_segs, segs_survey_record_with_distances_added, by = "Segment")

# great, add that information to the survey data
amph_lz_segs <- left_join(segs_survey_record_with_distances_added, amph_lz_segs)

# make sure nothing disappeared
amph_lz_segs %>% group_by(Station, Gross_ID) %>% 
  summarize(total_found = n_distinct(Animal_ID),
            species_found = n_distinct(Species))

# seems ok. NA is when no individuals were found on a segment I believe
```

```{r summarize what was surveyed}
# per night meters - this will be reported in Methods
env %>%
  group_by(Station) %>%
  summarise(mean_m_per_survey = mean(Meters),
            standard_deviation = sd(Meters))

```

# Detection rate per 500m

```{r detection rate}

# isolate surveys that DID have detections
surveys.no.zeros <- amph_lz_segs %>%
  filter(!is.na(Animal_ID))

# tally counts - df without zeros
survey.tally <- surveys.no.zeros %>%
  group_by(Start_Date, Station, Gross_ID, Segment, Seg_dist_m) %>% tally()

# generate all possible combinations based on family to get zeros

df_long <- survey.tally %>%
  as_tibble() %>%
  complete(Gross_ID, nesting(Start_Date, Station, Segment, Seg_dist_m), fill = list(n = 0))

# how many individuals found per 500m, on average?
df_long$n_per_500m <- (df_long$n/df_long$Seg_dist_m)*500

# summarize and visualize
detections.mean.se <- df_long %>%
  group_by(Station, Gross_ID) %>%
  summarise(mean = mean(n_per_500m),
            sd = sd(n_per_500m),
            n = n()) %>%
  mutate(se = sd / sqrt(n))

# different version without the individual data points
mean_detections_taxa <- ggplot(detections.mean.se, aes(x=Station, y=mean, fill = Gross_ID)) +
geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .2) + geom_point(pch = 21, size = 3, fill = "black") + theme_bw() + xlab("") + scale_color_discrete(labels = "") + ylab("Detections/500m") +
  theme(legend.position="none") + scale_x_discrete(labels=c("Unhunted", "Hunted")) + ggtitle("Mean number of detections per 500m by taxa") + facet_wrap(~Gross_ID, scales = "free") + theme(plot.title = element_text(size = 16))+
  theme(strip.text = element_text(size = 12)) +
  theme(axis.text.y.left = element_text(size = 14)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12))

mean_detections_taxa

#ggsave(mean_detections_taxa, file = "mean_detections_by_taxa.png")

```

```{r basic analysis}
# read about Kruskall Wallis tests
# structure: Dependent Variable ~ Independent Variable

# Frogs
frogs <- df_long %>%
  filter(Gross_ID == "Frog")

frog_test <- kruskal.test(n_per_500m ~ Station, data = frogs)
print(frog_test)

# Salamanders
salamanders <- df_long %>%
  filter(Gross_ID == "Salamander")

salamander_test <- kruskal.test(n_per_500m ~ Station, data = salamanders)

print(salamander_test)

# Lizards
lizards <- df_long %>%
  filter(Gross_ID == "Lizard")

lizard_test <- kruskal.test(n_per_500m ~ Station, data = lizards)

print(lizard_test)
```

```{r mixed effects model}

# If time for Jacob to learn, we could just do a glmm
library(glmmTMB)

taxa_glmm <- glmmTMB(n ~ Station/Gross_ID + (1|Segment), family = nbinom2(), ziformula = ~1, data = df_long) #ran through a few families using the rate (continuous) and this is what works (using count data, random effect of segment, and accounting for zero inflation and a negative binom family)

summary(taxa_glmm)

#chi squared table output
library(car)
Anova(taxa_glmm)

library(DHARMa)
# check model
sim <- simulateResiduals(fittedModel = taxa_glmm, plot = T)

testDispersion(taxa_glmm, plot = F)
testZeroInflation(taxa_glmm, plot = F)
```